---
title: "Weekly Maps **Cx. pipiens** and **Cx. tarsalis** abundance: SLC 2025 field season"
author: "Norah Saarman"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  # Saves PNGs to the root /figures folder
  fig.path = "../figures/" 
)
```

Weekly maps of **Culex pipiens** and **Culex tarsalis** abundance across SLC for the 2025 field season, overlay onto Google satellite image.

Load libraries:

```{r}
library(ggmap)
library(ggplot2)
library(sf)
library(dplyr)
library(gifski)
#library(magick)
```

# SQRT Weekly Maps (Square-Root Scaling)

Loop over all weeks with square-root scaling

Notes: 

Weekly file contained multiple observations per site and species (from different trap types or sampling dates), so added code to combine multiple observations from different trap types and sampling dates per week.

Also applied a square-root transformation to the variable before it is mapped to point size.


### Zoom 11

```{r all-weeks-zoom11, warning=FALSE}

# Weeks to plot
weeks <- 15:40

# Read the two species files once and combine them
pipiens_2025  <- read.csv("../data/pipiens_2025.csv")
tarsalis_2025 <- read.csv("../data/tarsalis_2025.csv")

all_data <- bind_rows(pipiens_2025, tarsalis_2025)

# 1. Register your Google Maps API key
register_google(key = "AIzaSyCdMfZPEp_xxkhdQrN2gfcGdBNpl1ga7ng")

# 2. Build a bounding box
lon_min <- -112.8
lon_max <- -111.2
lat_min <- 40.6
lat_max <- 41

# 3. Download the satellite map using the bounding box
bg <- get_map(
  location = c(
    left   = lon_min,
    bottom = lat_min,
    right  = lon_max,
    top    = lat_max
  ),
  maptype = "satellite",
  zoom = 11   #zoom in with zoom = 12
)

# 4. Find the global max across all weeks so the size scale is identical
global_max <- all_data %>%
  filter(disease_week %in% weeks) %>%
  group_by(disease_week, site_code, site_name, longitude, latitude, species) %>%
  summarise(count = sum(count, na.rm = TRUE), .groups = "drop") %>%
  summarise(global_max = max(count, na.rm = TRUE)) %>%
  pull(global_max)

print(paste("global max:", global_max))

# 5. Loop over all weeks
for (w in weeks) {

  spatial_data_raw <- all_data %>%
    filter(disease_week == w)
  
  # Combine multiple observations from different trap types and sampling dates per week
  spatial_data <- spatial_data_raw %>%
    group_by(site_code, site_name, longitude, latitude, species) %>%
    summarise(count = sum(count, na.rm = TRUE), .groups = "drop")

  # Reorder species so pipiens is plotted last (on top)
  spatial_data$species <- factor(
    spatial_data$species,
    levels = c("Culex tarsalis", "Culex pipiens")
  )

  map <- ggmap(bg) +

    # 1. Draw tarsalis FIRST (underneath)
    geom_point(
      data = subset(spatial_data, species == "Culex tarsalis"),
      aes(
        x = longitude,
        y = latitude,
        size = count,
        color = species,
        shape = species,
        alpha = species
      )
    ) +

    # 2. Draw pipiens SECOND (on top)
    geom_point(
      data = subset(spatial_data, species == "Culex pipiens"),
      aes(
        x = longitude,
        y = latitude,
        size = count,
        color = species,
        shape = species,
        alpha = species
      )
    ) +

    scale_color_manual(
      values = c(
        "Culex pipiens"  = "#00E5FF",
        "Culex tarsalis" = "#FF00AA"
      ),
      name = "Species",
      drop = FALSE
    ) +

    scale_shape_manual(
      values = c(
        "Culex pipiens"  = 16,  # circle
        "Culex tarsalis" = 16   # circle
      ),
      name = "Species",
      drop = FALSE
    ) +

    scale_alpha_manual(
      values = c(
        "Culex pipiens"  = 0.6,
        "Culex tarsalis" = 0.6
      ),
      guide = "none",
      drop = FALSE
    ) +

    scale_size_continuous(
      trans = "sqrt",
      range = c(3, 20),
      limits = c(1, global_max),
      breaks = c(1, 10, 50, 500, 5000),
      name = "Count"
    ) +
    
    #scale_size( # calls continuous automatically
    #  range = c(3, 20),
    #  limits = c(1, global_max),
    #  breaks = c(1, 10, 50, 100, 500, 5000),
    #  name = "Count"
    #) +

    labs(
      title = paste("Week", w),
      x = "Longitude",
      y = "Latitude"
    ) +

    theme_minimal()

  print(map)

  file_name <- paste0(
    "/uufs/chpc.utah.edu/common/home/saarman-group1/pip_tars_SLC2025/figures/d_weeks_2025_maps/zoom11/week_",
    w,
    "_zoom11.png"
  )
  
  # save one high res image
  ggsave(filename = file_name, plot = map, width = 8, height = 8, dpi = 300)


  gif_name <- paste0(
    "/uufs/chpc.utah.edu/common/home/saarman-group1/pip_tars_SLC2025/figures/d_weeks_2025_maps/zoom11/week_",
    w,
    "_zoom11_gif.png"
  )
  
  # save one low res image for gif
  ggsave(filename = gif_name, plot = map, width = 5, height = 5, dpi = 140)
}

```
  
  
### Zoom 12

```{r all-weeks-zoom12, warning=FALSE}

# Weeks to plot
weeks <- 15:40

# Read the two species files once and combine them
pipiens_2025  <- read.csv("../data/pipiens_2025.csv")
tarsalis_2025 <- read.csv("../data/tarsalis_2025.csv")

all_data <- bind_rows(pipiens_2025, tarsalis_2025)

# 1. Register your Google Maps API key
register_google(key = "AIzaSyCdMfZPEp_xxkhdQrN2gfcGdBNpl1ga7ng")

# 2. Build a bounding box
lon_min <- -112.8
lon_max <- -111.2
lat_min <- 40.6
lat_max <- 41

# 3. Download the satellite map using the bounding box
bg <- get_map(
  location = c(
    left   = lon_min,
    bottom = lat_min,
    right  = lon_max,
    top    = lat_max
  ),
  maptype = "satellite",
  zoom = 12   #zoom out with zoom = 11
)

# 4. Find the global max across all weeks so the size scale is identical
global_max <- all_data %>%
  filter(disease_week %in% weeks) %>%
  group_by(disease_week, site_code, site_name, longitude, latitude, species) %>%
  summarise(count = sum(count, na.rm = TRUE), .groups = "drop") %>%
  summarise(global_max = max(count, na.rm = TRUE)) %>%
  pull(global_max)

print(paste("global max:", global_max))

# 5. Loop over all weeks
for (w in weeks) {

  spatial_data_raw <- all_data %>%
    filter(disease_week == w)
  
  # Combine multiple observations from different trap types and sampling dates per week
  spatial_data <- spatial_data_raw %>%
    group_by(site_code, site_name, longitude, latitude, species) %>%
    summarise(count = sum(count, na.rm = TRUE), .groups = "drop")

  # Reorder species so pipiens is plotted last (on top)
  spatial_data$species <- factor(
    spatial_data$species,
    levels = c("Culex tarsalis", "Culex pipiens")
  )

  map <- ggmap(bg) +

    # 1. Draw tarsalis FIRST (underneath)
    geom_point(
      data = subset(spatial_data, species == "Culex tarsalis"),
      aes(
        x = longitude,
        y = latitude,
        size = count,
        color = species,
        shape = species,
        alpha = species
      )
    ) +

    # 2. Draw pipiens SECOND (on top)
    geom_point(
      data = subset(spatial_data, species == "Culex pipiens"),
      aes(
        x = longitude,
        y = latitude,
        size = count,
        color = species,
        shape = species,
        alpha = species
      )
    ) +

    scale_color_manual(
      values = c(
        "Culex pipiens"  = "#00E5FF",
        "Culex tarsalis" = "#FF00AA"
      ),
      name = "Species",
      drop = FALSE
    ) +

    scale_shape_manual(
      values = c(
        "Culex pipiens"  = 16,  # circle
        "Culex tarsalis" = 16   # circle
      ),
      name = "Species",
      drop = FALSE
    ) +

    scale_alpha_manual(
      values = c(
        "Culex pipiens"  = 0.6,
        "Culex tarsalis" = 0.6
      ),
      guide = "none",
      drop = FALSE
    ) +

    scale_size_continuous(
      trans = "sqrt",
      range = c(3, 20),
      limits = c(1, global_max),
      breaks = c(1, 10, 50, 500, 5000),
      name = "Count"
    ) +
    
    #scale_size( # calls continuous automatically
    #  range = c(3, 20),
    #  limits = c(1, global_max),
    #  breaks = c(1, 10, 50, 100, 500, 5000),
    #  name = "Count"
    #) +

    labs(
      title = paste("Week", w),
      x = "Longitude",
      y = "Latitude"
    ) +

    theme_minimal()

  print(map)

  file_name <- paste0(
    "/uufs/chpc.utah.edu/common/home/saarman-group1/pip_tars_SLC2025/figures/d_weeks_2025_maps/zoom12/week_",
    w,
    "_zoom12.png"
  )

  ggsave(filename = file_name, plot = map, width = 8, height = 8, dpi = 300)
  
  
  gif_name <- paste0(
    "/uufs/chpc.utah.edu/common/home/saarman-group1/pip_tars_SLC2025/figures/d_weeks_2025_maps/zoom12/week_",
    w,
    "_zoom12_gif.png"
  )
  
  # save one low res image for gif
  ggsave(filename = gif_name, plot = map, width = 5, height = 5, dpi = 140)
}
```
  
### SQRT weekly GIFS for Zoom 11 and 12

```{r create-gifs}
library(gifski)

# Create GIF for zoom 11
gif_dir <- "/uufs/chpc.utah.edu/common/home/saarman-group1/pip_tars_SLC2025/figures/d_weeks_2025_maps/zoom11"

gif_imgs <- list.files(
  gif_dir,
  pattern = "week_.*_zoom11_gif\\.png$",
  full.names = TRUE
)

gif_weeks <- as.numeric(sub(".*week_([0-9]+).*", "\\1", basename(gif_imgs)))
gif_imgs <- gif_imgs[order(gif_weeks)]

gifski(
  png_files = gif_imgs,
  gif_file  = file.path(gif_dir, "mosquito_2025_zoom11.gif"),
  width     = 700,
  height    = 700,
  delay     = 0.6
)


# Create GIF for zoom 12
gif_dir <- "/uufs/chpc.utah.edu/common/home/saarman-group1/pip_tars_SLC2025/figures/d_weeks_2025_maps/zoom12"

gif_imgs <- list.files(
  gif_dir,
  pattern = "week_.*_zoom12_gif\\.png$",
  full.names = TRUE
)

gif_weeks <- as.numeric(sub(".*week_([0-9]+).*", "\\1", basename(gif_imgs)))
gif_imgs <- gif_imgs[order(gif_weeks)]

gifski(
  png_files = gif_imgs,
  gif_file  = file.path(gif_dir, "mosquito_2025_zoom12.gif"),
  width     = 700,
  height    = 700,
  delay     = 0.6
)
```

# SQRT Seasonal  Seasonal averages (Square-Root Scaling)

```{r}
library(dplyr)

# read data
pipiens  <- read.csv("../data/pipiens_2025.csv")
tarsalis <- read.csv("../data/tarsalis_2025.csv")

all_data <- bind_rows(pipiens, tarsalis)

# define seasons from week numbers
all_data <- all_data %>%
  mutate(
    season_calc = case_when(
      disease_week %in% 15:22 ~ "early",
      disease_week %in% 23:31 ~ "mid",
      disease_week %in% 32:40 ~ "late",
      TRUE ~ NA_character_
    )
  )

# compute mean counts across weeks
seasonal_avg <- all_data %>%
  filter(!is.na(season_calc)) %>%
  group_by(season_calc, site_code, site_name, longitude, latitude, species, disease_week) %>%
  summarise(weekly_count = sum(count, na.rm = TRUE), .groups = "drop") %>%
  group_by(season_calc, site_code, site_name, longitude, latitude, species) %>%
  summarise(mean_count = mean(weekly_count, na.rm = TRUE), .groups = "drop")

seasonal_avg
```

### Zoom 11
```{r all-seasons-zoom11, warning=FALSE}

# Seasons to plot
seasons <- c("early", "mid", "late")

# Read the two species files once and combine them
pipiens_2025  <- read.csv("../data/pipiens_2025.csv")
tarsalis_2025 <- read.csv("../data/tarsalis_2025.csv")

all_data <- bind_rows(pipiens_2025, tarsalis_2025) %>%
  mutate(
    season_plot = case_when(
      disease_week %in% 15:22 ~ "early",
      disease_week %in% 23:31 ~ "mid",
      disease_week %in% 32:40 ~ "late",
      TRUE ~ NA_character_
    )
  )

# 1. Register your Google Maps API key
register_google(key = "AIzaSyCdMfZPEp_xxkhdQrN2gfcGdBNpl1ga7ng")

# 2. Build a bounding box
lon_min <- -112.8
lon_max <- -111.2
lat_min <- 40.6
lat_max <- 41

# 3. Download the satellite map using the bounding box
bg <- get_map(
  location = c(
    left   = lon_min,
    bottom = lat_min,
    right  = lon_max,
    top    = lat_max
  ),
  maptype = "satellite",
  zoom = 11   #zoom in with zoom = 12
)

# 4. Find the global max across all seasons so the size scale is identical
global_max <- all_data %>%
  filter(!is.na(season_plot)) %>%
  group_by(season_plot, site_code, site_name, longitude, latitude, species, disease_week) %>%
  summarise(count = sum(count, na.rm = TRUE), .groups = "drop") %>%
  group_by(season_plot, site_code, site_name, longitude, latitude, species) %>%
  summarise(mean_count = mean(count, na.rm = TRUE), .groups = "drop") %>%
  summarise(global_max = max(mean_count, na.rm = TRUE)) %>%
  pull(global_max)

print(paste("global max:", global_max))

# 5. Loop over all seasons
for (s in seasons) {

  spatial_data_raw <- all_data %>%
    filter(season_plot == s)
  
  # Combine multiple observations from different trap types and sampling dates per week,
  # then average across weeks within season
  spatial_data <- spatial_data_raw %>%
    group_by(site_code, site_name, longitude, latitude, species, disease_week) %>%
    summarise(count = sum(count, na.rm = TRUE), .groups = "drop") %>%
    group_by(site_code, site_name, longitude, latitude, species) %>%
    summarise(mean_count = mean(count, na.rm = TRUE), .groups = "drop")

  # Reorder species so pipiens is plotted last (on top)
  spatial_data$species <- factor(
    spatial_data$species,
    levels = c("Culex tarsalis", "Culex pipiens")
  )

  map <- ggmap(bg) +

    # 1. Draw tarsalis FIRST (underneath)
    geom_point(
      data = subset(spatial_data, species == "Culex tarsalis"),
      aes(
        x = longitude,
        y = latitude,
        size = mean_count,
        color = species,
        shape = species,
        alpha = species
      )
    ) +

    # 2. Draw pipiens SECOND (on top)
    geom_point(
      data = subset(spatial_data, species == "Culex pipiens"),
      aes(
        x = longitude,
        y = latitude,
        size = mean_count,
        color = species,
        shape = species,
        alpha = species
      )
    ) +

    scale_color_manual(
      values = c(
        "Culex pipiens"  = "#00E5FF",
        "Culex tarsalis" = "#FF00AA"
      ),
      name = "Species",
      drop = FALSE
    ) +

    scale_shape_manual(
      values = c(
        "Culex pipiens"  = 16,  # circle
        "Culex tarsalis" = 16   # circle
      ),
      name = "Species",
      drop = FALSE
    ) +

    scale_alpha_manual(
      values = c(
        "Culex pipiens"  = 0.6,
        "Culex tarsalis" = 0.6
      ),
      guide = "none",
      drop = FALSE
    ) +

    scale_size_continuous(
      trans = "sqrt",
      range = c(3, 20),
      limits = c(1, global_max),
      breaks = c(1, 10, 50, 500, 5000),
      name = "Mean count"
    ) +

    labs(
      title = paste(tools::toTitleCase(s), "season"),
      x = "Longitude",
      y = "Latitude"
    ) +

    theme_minimal()

  print(map)

  file_name <- paste0(
    "/uufs/chpc.utah.edu/common/home/saarman-group1/pip_tars_SLC2025/figures/d_weeks_2025_maps/zoom11/",
    s,
    "_zoom11.png"
  )
  
  # save one high res image
  ggsave(filename = file_name, plot = map, width = 8, height = 8, dpi = 300)
}
```
  
### Zoom 12

```{r all-seasons-zoom12, warning=FALSE}

# Seasons to plot
seasons <- c("early", "mid", "late")

# Read the two species files once and combine them
pipiens_2025  <- read.csv("../data/pipiens_2025.csv")
tarsalis_2025 <- read.csv("../data/tarsalis_2025.csv")

all_data <- bind_rows(pipiens_2025, tarsalis_2025) %>%
  mutate(
    season_plot = case_when(
      disease_week %in% 15:22 ~ "early",
      disease_week %in% 23:31 ~ "mid",
      disease_week %in% 32:40 ~ "late",
      TRUE ~ NA_character_
    )
  )

# 1. Register your Google Maps API key
register_google(key = "AIzaSyCdMfZPEp_xxkhdQrN2gfcGdBNpl1ga7ng")

# 2. Build a bounding box
lon_min <- -112.8
lon_max <- -111.2
lat_min <- 40.6
lat_max <- 41

# 3. Download the satellite map using the bounding box
bg <- get_map(
  location = c(
    left   = lon_min,
    bottom = lat_min,
    right  = lon_max,
    top    = lat_max
  ),
  maptype = "satellite",
  zoom = 12   #zoom out with zoom = 11
)

# 4. Find the global max across all seasons so the size scale is identical
global_max <- all_data %>%
  filter(!is.na(season_plot)) %>%
  group_by(season_plot, site_code, site_name, longitude, latitude, species, disease_week) %>%
  summarise(count = sum(count, na.rm = TRUE), .groups = "drop") %>%
  group_by(season_plot, site_code, site_name, longitude, latitude, species) %>%
  summarise(mean_count = mean(count, na.rm = TRUE), .groups = "drop") %>%
  summarise(global_max = max(mean_count, na.rm = TRUE)) %>%
  pull(global_max)

print(paste("global max:", global_max))

# 5. Loop over all seasons
for (s in seasons) {

  spatial_data_raw <- all_data %>%
    filter(season_plot == s)
  
  # Combine multiple observations from different trap types and sampling dates per week,
  # then average across weeks within season
  spatial_data <- spatial_data_raw %>%
    group_by(site_code, site_name, longitude, latitude, species, disease_week) %>%
    summarise(count = sum(count, na.rm = TRUE), .groups = "drop") %>%
    group_by(site_code, site_name, longitude, latitude, species) %>%
    summarise(mean_count = mean(count, na.rm = TRUE), .groups = "drop")

  # Reorder species so pipiens is plotted last (on top)
  spatial_data$species <- factor(
    spatial_data$species,
    levels = c("Culex tarsalis", "Culex pipiens")
  )

  map <- ggmap(bg) +

    # 1. Draw tarsalis FIRST (underneath)
    geom_point(
      data = subset(spatial_data, species == "Culex tarsalis"),
      aes(
        x = longitude,
        y = latitude,
        size = mean_count,
        color = species,
        shape = species,
        alpha = species
      )
    ) +

    # 2. Draw pipiens SECOND (on top)
    geom_point(
      data = subset(spatial_data, species == "Culex pipiens"),
      aes(
        x = longitude,
        y = latitude,
        size = mean_count,
        color = species,
        shape = species,
        alpha = species
      )
    ) +

    scale_color_manual(
      values = c(
        "Culex pipiens"  = "#00E5FF",
        "Culex tarsalis" = "#FF00AA"
      ),
      name = "Species",
      drop = FALSE
    ) +

    scale_shape_manual(
      values = c(
        "Culex pipiens"  = 16,  # circle
        "Culex tarsalis" = 16   # circle
      ),
      name = "Species",
      drop = FALSE
    ) +

    scale_alpha_manual(
      values = c(
        "Culex pipiens"  = 0.6,
        "Culex tarsalis" = 0.6
      ),
      guide = "none",
      drop = FALSE
    ) +

    scale_size_continuous(
      trans = "sqrt",
      range = c(3, 20),
      limits = c(1, global_max),
      breaks = c(1, 10, 50, 500, 5000),
      name = "Mean count"
    ) +

    labs(
      title = paste(tools::toTitleCase(s), "season"),
      x = "Longitude",
      y = "Latitude"
    ) +

    theme_minimal()

  print(map)

  file_name <- paste0(
    "/uufs/chpc.utah.edu/common/home/saarman-group1/pip_tars_SLC2025/figures/d_weeks_2025_maps/zoom12/",
    s,
    "_zoom12.png"
  )

  ggsave(filename = file_name, plot = map, width = 8, height = 8, dpi = 300)
}

```


# Raw Weekly Maps (Raw Scaling)

Loop over all weeks with raw scaling.

Notes:

Weekly file contained multiple observations per site and species (from different trap types or sampling dates), so added code to combine multiple observations from different trap types and sampling dates per week.

Point size is mapped directly to the raw count value for visualization only.

### Zoom 11
```{r z11-raw}

# Weeks to plot
weeks <- 15:40

# Read the two species files once and combine them
pipiens_2025  <- read.csv("../data/pipiens_2025.csv")
tarsalis_2025 <- read.csv("../data/tarsalis_2025.csv")

all_data <- bind_rows(pipiens_2025, tarsalis_2025)

# 1. Register your Google Maps API key
register_google(key = "AIzaSyCdMfZPEp_xxkhdQrN2gfcGdBNpl1ga7ng")

# 2. Build a bounding box
lon_min <- -112.8
lon_max <- -111.2
lat_min <- 40.6
lat_max <- 41

# 3. Download the satellite map using the bounding box
bg <- get_map(
  location = c(
    left   = lon_min,
    bottom = lat_min,
    right  = lon_max,
    top    = lat_max
  ),
  maptype = "satellite",
  zoom = 11
)

# 4. Find the global max across all weeks so the size scale is identical
global_max <- all_data %>%
  filter(disease_week %in% weeks) %>%
  group_by(disease_week, site_code, site_name, longitude, latitude, species) %>%
  summarise(count = sum(count, na.rm = TRUE), .groups = "drop") %>%
  summarise(global_max = max(count, na.rm = TRUE)) %>%
  pull(global_max)

print(paste("global max:", global_max))

# 5. Loop over all weeks
for (w in weeks) {

  spatial_data_raw <- all_data %>%
    filter(disease_week == w)

  # Combine multiple observations from different trap types and sampling dates per week
  spatial_data <- spatial_data_raw %>%
    group_by(site_code, site_name, longitude, latitude, species) %>%
    summarise(count = sum(count, na.rm = TRUE), .groups = "drop")

  # Reorder species so pipiens is plotted last (on top)
  spatial_data$species <- factor(
    spatial_data$species,
    levels = c("Culex tarsalis", "Culex pipiens")
  )

  map <- ggmap(bg) +

    # 1. Draw tarsalis FIRST (underneath)
    geom_point(
      data = subset(spatial_data, species == "Culex tarsalis"),
      aes(
        x = longitude,
        y = latitude,
        size = count,
        color = species,
        shape = species,
        alpha = species
      )
    ) +

    # 2. Draw pipiens SECOND (on top)
    geom_point(
      data = subset(spatial_data, species == "Culex pipiens"),
      aes(
        x = longitude,
        y = latitude,
        size = count,
        color = species,
        shape = species,
        alpha = species
      )
    ) +

    scale_color_manual(
      values = c(
        "Culex pipiens"  = "#00E5FF",
        "Culex tarsalis" = "#FF00AA"
      ),
      name = "Species",
      drop = FALSE
    ) +

    scale_shape_manual(
      values = c(
        "Culex pipiens"  = 16,
        "Culex tarsalis" = 16
      ),
      name = "Species",
      drop = FALSE
    ) +

    scale_alpha_manual(
      values = c(
        "Culex pipiens"  = 0.6,
        "Culex tarsalis" = 0.6
      ),
      guide = "none",
      drop = FALSE
    ) +

    scale_size_continuous(
      range = c(3, 20),
      limits = c(1, global_max),
      breaks = c(1, 10, 50, 500, 5000),
      name = "Count"
    ) +

    labs(
      title = paste("Week", w),
      x = "Longitude",
      y = "Latitude"
    ) +

    theme_minimal()

  print(map)

  file_name <- paste0(
    "/uufs/chpc.utah.edu/common/home/saarman-group1/pip_tars_SLC2025/figures/d_weeks_2025_maps/zoom11/week_",
    w,
    "_zoom11_raw.png"
  )

  # save one high res image
  ggsave(filename = file_name, plot = map, width = 8, height = 8, dpi = 300)

  gif_name <- paste0(
    "/uufs/chpc.utah.edu/common/home/saarman-group1/pip_tars_SLC2025/figures/d_weeks_2025_maps/zoom11/week_",
    w,
    "_zoom11_gif_raw.png"
  )

  # save one low res image for gif
  ggsave(filename = gif_name, plot = map, width = 5, height = 5, dpi = 140)
}
```

### Zoom 12

```{r z12-raw}
# Weeks to plot
weeks <- 15:40

# Read the two species files once and combine them
pipiens_2025  <- read.csv("../data/pipiens_2025.csv")
tarsalis_2025 <- read.csv("../data/tarsalis_2025.csv")

all_data <- bind_rows(pipiens_2025, tarsalis_2025)

# 1. Register your Google Maps API key
register_google(key = "AIzaSyCdMfZPEp_xxkhdQrN2gfcGdBNpl1ga7ng")

# 2. Build a bounding box
lon_min <- -112.8
lon_max <- -111.2
lat_min <- 40.6
lat_max <- 41

# 3. Download the satellite map using the bounding box
bg <- get_map(
  location = c(
    left   = lon_min,
    bottom = lat_min,
    right  = lon_max,
    top    = lat_max
  ),
  maptype = "satellite",
  zoom = 12
)

# 4. Find the global max across all weeks so the size scale is identical
global_max <- all_data %>%
  filter(disease_week %in% weeks) %>%
  group_by(disease_week, site_code, site_name, longitude, latitude, species) %>%
  summarise(count = sum(count, na.rm = TRUE), .groups = "drop") %>%
  summarise(global_max = max(count, na.rm = TRUE)) %>%
  pull(global_max)

print(paste("global max:", global_max))

# 5. Loop over all weeks
for (w in weeks) {

  spatial_data_raw <- all_data %>%
    filter(disease_week == w)

  # Combine multiple observations from different trap types and sampling dates per week
  spatial_data <- spatial_data_raw %>%
    group_by(site_code, site_name, longitude, latitude, species) %>%
    summarise(count = sum(count, na.rm = TRUE), .groups = "drop")

  # Reorder species so pipiens is plotted last (on top)
  spatial_data$species <- factor(
    spatial_data$species,
    levels = c("Culex tarsalis", "Culex pipiens")
  )

  map <- ggmap(bg) +

    # 1. Draw tarsalis FIRST (underneath)
    geom_point(
      data = subset(spatial_data, species == "Culex tarsalis"),
      aes(
        x = longitude,
        y = latitude,
        size = count,
        color = species,
        shape = species,
        alpha = species
      )
    ) +

    # 2. Draw pipiens SECOND (on top)
    geom_point(
      data = subset(spatial_data, species == "Culex pipiens"),
      aes(
        x = longitude,
        y = latitude,
        size = count,
        color = species,
        shape = species,
        alpha = species
      )
    ) +

    scale_color_manual(
      values = c(
        "Culex pipiens"  = "#00E5FF",
        "Culex tarsalis" = "#FF00AA"
      ),
      name = "Species",
      drop = FALSE
    ) +

    scale_shape_manual(
      values = c(
        "Culex pipiens"  = 16,
        "Culex tarsalis" = 16
      ),
      name = "Species",
      drop = FALSE
    ) +

    scale_alpha_manual(
      values = c(
        "Culex pipiens"  = 0.6,
        "Culex tarsalis" = 0.6
      ),
      guide = "none",
      drop = FALSE
    ) +

    scale_size_continuous(
      range = c(3, 20),
      limits = c(1, global_max),
      breaks = c(1, 10, 50, 500, 5000),
      name = "Count"
    ) +

    labs(
      title = paste("Week", w),
      x = "Longitude",
      y = "Latitude"
    ) +

    theme_minimal()

  print(map)

  file_name <- paste0(
    "/uufs/chpc.utah.edu/common/home/saarman-group1/pip_tars_SLC2025/figures/d_weeks_2025_maps/zoom12/week_",
    w,
    "_zoom12_raw.png"
  )

  ggsave(filename = file_name, plot = map, width = 8, height = 8, dpi = 300)

  gif_name <- paste0(
    "/uufs/chpc.utah.edu/common/home/saarman-group1/pip_tars_SLC2025/figures/d_weeks_2025_maps/zoom12/week_",
    w,
    "_zoom12_gif_raw.png"
  )

  # save one low res image for gif
  ggsave(filename = gif_name, plot = map, width = 5, height = 5, dpi = 140)
}
```

### Raw weekly GIFs for raw, zoom 11 and 12
```{r gifs-raw}
library(gifski)

# Create GIF for zoom 11
gif_dir <- "/uufs/chpc.utah.edu/common/home/saarman-group1/pip_tars_SLC2025/figures/d_weeks_2025_maps/zoom11"

gif_imgs <- list.files(
  gif_dir,
  pattern = "week_.*_zoom11_gif\\_raw.png$",
  full.names = TRUE
)

gif_weeks <- as.numeric(sub(".*week_([0-9]+).*", "\\1", basename(gif_imgs)))
gif_imgs <- gif_imgs[order(gif_weeks)]

gifski(
  png_files = gif_imgs,
  gif_file  = file.path(gif_dir, "mosquito_2025_zoom11_raw.gif"),
  width     = 700,
  height    = 700,
  delay     = 0.6
)

# Create GIF for zoom 12
gif_dir <- "/uufs/chpc.utah.edu/common/home/saarman-group1/pip_tars_SLC2025/figures/d_weeks_2025_maps/zoom12"

gif_imgs <- list.files(
  gif_dir,
  pattern = "week_.*_zoom12_gif\\_raw.png$",
  full.names = TRUE
)

gif_weeks <- as.numeric(sub(".*week_([0-9]+).*", "\\1", basename(gif_imgs)))
gif_imgs <- gif_imgs[order(gif_weeks)]

gifski(
  png_files = gif_imgs,
  gif_file  = file.path(gif_dir, "mosquito_2025_zoom12_raw.gif"),
  width     = 700,
  height    = 700,
  delay     = 0.6
)
```
  
# Uinta Basin map for Chapter 1

Google map
```{r uinta-google}
library(tidyverse)
library(ggmap)
library(ggrepel)

# Register Google key 
register_google(key = "AIzaSyCdMfZPEp_xxkhdQrN2gfcGdBNpl1ga7ng")

uinta_sites <- tribble(
  ~Locality,        ~lat,        ~lon,        ~group,
  "Ouray",          40.0837,    -109.6628,   "Not detected",
  "Randlett",       40.2311,    -109.8216,   "Cx. pipiens morphotype",
  "Sunshine",       40.2656,    -109.8509,   "Cx. pipiens confirmed",
  "Fort Duchesne",  40.2757,    -109.8534,   "Not detected",
  "Thunder Ridge",  40.3652,    -109.8219,   "Not detected",
  "Whiterocks",     40.4583,    -109.9339,   "Not detected",
  "Neola",          40.4473,    -110.0285,   "Not detected",
  "Yellowstone",    40.3289,    -109.8911,   "Cx. pipiens confirmed",
  "Hilltop",        40.2948,    -109.8977,   "Not detected",
  "Arcadia",        40.2069,    -110.1855,   "Cx. pipiens morphotype",
  "Myton",          40.2002,    -110.0612,   "Not detected",
  "Wetlands",       40.1931,    -109.9128,   "Not detected",
  "Library",        40 + 27/60 + 21.6/3600, -(109 + 31/60 + 30.9/3600), "Not detected",
  "Boothill",       40 + 26/60 + 18.7/3600, -(109 + 35/60 + 11.8/3600), "Cx. pipiens confirmed",
  "Fausett Ranch",  40 + 19/60 + 51.8/3600, -(109 + 50/60 + 54.4/3600), "Not detected"
)

lon_min <- -110.25
lon_max <- -109.45
lat_min <- 40.02
lat_max <- 40.52

bg <- get_map(
  location = c(
    left = lon_min,
    bottom = lat_min,
    right = lon_max,
    top = lat_max
  ),
  maptype = "terrain",
  zoom = 10
)

group_levels <- c(
  "Cx. pipiens confirmed",
  "Cx. pipiens morphotype",
  "Not detected"
)

uinta_sites <- uinta_sites %>%
  mutate(group = factor(group, levels = group_levels))

ggmap(bg) +
  geom_point(
    data = uinta_sites,
    aes(x = lon, y = lat, fill = group, shape = group),
    size = 4,
    color = "grey20",
    stroke = 0.8
  ) +
  geom_text_repel(
    data = uinta_sites,
    aes(x = lon, y = lat, label = Locality),
    size = 5,
    fontface = "bold",
    box.padding = 0.4,
    point.padding = 0.3,
    segment.color = "grey40",
    max.overlaps = Inf
  ) +
  scale_fill_manual(
  name = NULL,
  breaks = group_levels,
  values = c(
    "Cx. pipiens confirmed" = "#FF2DA0",
    "Cx. pipiens morphotype" = "purple",
    "Not detected" = "#1bc8ea"
  )
) +
scale_shape_manual(
  name = NULL,
  breaks = group_levels,
  values = c(
    "Cx. pipiens confirmed" = 24,
    "Cx. pipiens morphotype" = 22,
    "Not detected" = 21
  )
) +
guides(
  fill = guide_legend(
    override.aes = list(
      shape = c(24, 22, 21),
      fill = c("#FF2DA0", "purple", "#1bc8ea"),
      color = "grey20",
      size = 5
    )
  ),
  shape = "none"
) +
  coord_quickmap(
    xlim = c(lon_min, lon_max),
    ylim = c(lat_min, lat_max),
    expand = FALSE
  ) +
  labs(fill = NULL) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 12),
    plot.margin = margin(10, 10, 10, 10)
  )
```



```{r sdm}
# ============================================================
# Uinta Basin map with SDM suitability > 50%
# ============================================================

library(tidyverse)
library(ggmap)
library(terra)

# Register Google key 
register_google(key = "AIzaSyCdMfZPEp_xxkhdQrN2gfcGdBNpl1ga7ng")

# -------------------------
# Map extent
# -------------------------
lon_min <- -110.25
lon_max <- -109.45
lat_min <- 40.02
lat_max <- 40.52

bg <- get_map(
  location = c(
    left   = lon_min,
    bottom = lat_min,
    right  = lon_max,
    top    = lat_max
  ),
  maptype = "terrain",
  zoom = 10
)


# -------------------------
# Load raw SDM raster from Gorris et al.
# -------------------------

sdm_pipiens <- terra::rast("../../ace2/gis/culex_pipiens_meansuitability.nc")

# Crop to Uinta Basin
uinta_ext <- ext(lon_min, lon_max, lat_min, lat_max)
sdm_uinta <- crop(sdm_pipiens, uinta_ext)

# Convert to dataframe
sdm_df <- as.data.frame(sdm_uinta, xy = TRUE, na.rm = TRUE)

# Rename suitability column
names(sdm_df)[3] <- "suitability"

# -------------------------
# Plot continuous suitability
# -------------------------
ggmap(bg) +
  geom_raster(
    data = sdm_df,
    aes(
      x = x,
      y = y,
      fill = suitability
    ),
    alpha = 0.55,
    interpolate = TRUE
  ) +
  scale_fill_gradient(
    low = "transparent",
    high = "#FF2DA0",
    name = "Predicted\nsuitability",
    limits = c(0.35, .61)
  ) +
  coord_quickmap(
    xlim = c(lon_min, lon_max),
    ylim = c(lat_min, lat_max),
    expand = FALSE
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11),
    plot.margin = margin(10, 10, 10, 10)
  )
```
```{r}
# Wider map extent
lon_min <- -110.35
lon_max <- -109.35
lat_min <- 39.95
lat_max <- 40.58

bg <- get_map(
  location = c(
    left   = lon_min,
    bottom = lat_min,
    right  = lon_max,
    top    = lat_max
  ),
  maptype = "terrain",
  zoom = 10
)

sdm_pipiens <- terra::rast(
  "../../ace2/gis/culex_pipiens_meansuitability.nc"
)

uinta_ext <- terra::ext(lon_min, lon_max, lat_min, lat_max)

sdm_uinta <- terra::crop(sdm_pipiens, uinta_ext)

sdm_df <- as.data.frame(sdm_uinta, xy = TRUE, na.rm = TRUE)
names(sdm_df)[3] <- "suitability"

sdm_df_50 <- sdm_df %>%
  filter(suitability > 0.3)

ggmap(bg) +
  geom_raster(
    data = sdm_df_50,
    aes(x = x, y = y, fill = suitability),
    alpha = 0.75,
    interpolate = TRUE
  ) +
  scale_fill_gradient(
    low = "#FDBBD7",
    high = "#FF2DA0",
    limits = c(0.3, 0.61),
    name = "Predicted\nsuitability"
  ) +
  coord_quickmap(
    xlim = c(lon_min, lon_max),
    ylim = c(lat_min, lat_max),
    expand = FALSE
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11),
    plot.margin = margin(10, 10, 10, 10)
  )
```