---
title: "GLMM **Cx. pipiens** and **Cx. tarsalis** abundance: SLC 2025 field season"
author: "Norah Saarman"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Research Topic:** testing whether habitat and seasonal partitioning between Culex pipiens s.l. and Culex tarsalis shapes West Nile Virus (WNV) dynamics across urban–rural gradients. 

**Core hypothesis:** early/mid-season amplification dominated by pipiens in urban areas, later spillover involving tarsalis moving into urban/peri-urban areas.

**Approach:** Preliminary results visualized via mapping, with species identity and abundance as primary response variables. Model mosquito abundance and proportions using GLMMs count ~ season*urbanization + trap_type + (1|site/date), family = poisson(link = "log"):   
 - Response variable = mosquito abundance  
 - Predictors = season*urbanization  
 - The trap type could be important, so we will add that as a fixed effect (covariate)... is this correct? We do think that the response variable of count of mosquitoes depends on trap type, since tarsalis seems to be more attracted to CO2 than pipiens, and we want to quantify that effect. Note that poisson model does not give a fixed offset (due to the log link)... The structure of this model means that it will estimate an effect that scales with the total number of mosquitos caught, which is exactly what we want.  
 - The data are grouped into sites and are also linked through time, so we'll add those as random effects. I think the sites should be coded as factors, **but I'm not sure what format to use for the date. I think it should be disease week so that week 18 is treated closer to 19 than 20, etc., but I'm not totally confident in this.**   
 - The family = poisson (link = "log")... why again?  

Load libraries
```{r}
library(tidyverse) # for data wrangling
library(glmmTMB)   # for model fitting
library(DHARMa)    # for residual plots
library(emmeans)   # for estimating marginal effects
library(multcomp)  # for statistical comparisons on fitted models
library(dplyr)     # for mutating dataframe to change labels in dataset

```

# Salamander example

Just to understand a bit better what it looks like when residuals don't "look good":
```{r salamander}
# use the salamanders dataset from glmmTMB:
data("Salamanders")
head(Salamanders)

# we'll model salamander count as a function of species (spp) and mining presence (mined).
# Because the data are grouped into sites, we'll add that as a random effect.
# first, look at the response distribution

hist(Salamanders$count)
hist(log(Salamanders$count))

# count data, so start with poisson model:
fit_pois <- glmmTMB(count ~ spp*mined + (1 | site),
                    family = poisson(link = "log"),
                    data = Salamanders)

summary(fit_pois)
# look at the residuals:
simulateResiduals(fit_pois, plot = T)
# this doesn't look great. It seems like there is a trend in the residuals with
# greater error at higher values. Maybe the negative binomial distribution can help with this

#continue salamander dataset example
# this doesn't look great. It seems like there is a trend in the residuals with
# greater error at higher values. Maybe the negative binomial distribution can help with this

fit_nbinom <- glmmTMB(count ~ spp*mined + (1 | site),
                      family = nbinom1(link = "log"),
                      data = Salamanders)
simulateResiduals(fit_nbinom, plot = T)

# this looks much better. Now we can look at the marginal effects:

# First, just look at the effect of species, doing a 1-way pairwise comparison
em_spp = emmeans(fit_nbinom, pairwise ~ spp, type = "response")
em_spp

cl = cld(em_spp, Letters = letters)

ggplot(data = cl, aes(x = spp, y = response)) +
  geom_point(size=2.5, color="black") +
  geom_errorbar(aes(x=spp, ymin = asymp.LCL,
                    ymax = asymp.UCL),
                width = 0.2, size=1, color="black") +
  geom_text(aes(label = gsub(" ", "", .group)),
            position = position_nudge(x = 0.3)) +
  ggeasy::easy_remove_axes("both", "title")


# Next, let's see how the effects vary with different mined levels in a 2-way emmeans comparison
em_spp2 = emmeans(fit_nbinom, pairwise ~ spp | mined, type = "response")
em_spp2

cl2 = cld(em_spp2, Letters = letters)

ggplot(data = cl2, aes(x = spp, y = response)) +
  geom_point(size=2.5, color="black") +
  geom_errorbar(aes(x=spp, ymin = asymp.LCL,
                    ymax = asymp.UCL),
                width = 0.2, size=1, color="black") +
  facet_wrap(~ mined, #nrow = 2,
             labeller = label_both) +
  geom_text(aes(label = gsub(" ", "", .group)),
            position = position_nudge(x = 0.4)) +
  ggeasy::easy_remove_axes("y", "title")
```

  
# Culex tarsalis

 
  
```{r tarsalis}
## tarsalis datasets from SLCMAD:
tarsalis <- read.csv("../data/tarsalis_2025.csv")

tarsalis <- tarsalis %>%
  mutate(
    urbanization = factor(
      urban_cat,
      levels = c("rural", "peri", "urban"),
      labels = c("rural", "peri", "urban")
    ),
    season = factor(
      season,
      levels = c("early", "mid", "late"),
      labels = c("early", "mid", "late")
    )
  )

head(tarsalis, 5)
```


# GLMM by season and urbanization, each mosquito species separately

Start by binning into discrete seasons: early, mid, late

```{r seasons-bins}
table(tarsalis$season, tarsalis$disease_week)
```
  
First, look at the response distribution
```{r response}
# sub in tarsalis dataset here instead of salamander 
hist(tarsalis$count) 
hist(log(tarsalis$count))

```

## Random effects (1 | site_name/disease_week)

Fit the model with poisson(link="log"), using random effects (1 | site_name/disease_week)

```{r poisson-log-link-week}
# tarsalis count data, starting with poisson model:
fit_pois <- glmmTMB(count ~ season*urbanization + trap_type + (1 | site_name/disease_week),
                    family = poisson(link = "log"),
                    data = tarsalis)
summary(fit_pois)
```

Look at the residuals:
```{r resids-poisson-log-link-week}
simulateResiduals(fit_pois, plot = T)
```
## Random effects (1 | site_name/collection_date)

Fit the model with poisson(link="log"), using random effects (1 | site_name/collection_date)

```{r poisson-log-link-date}
# tarsalis count data, starting with poisson model:
fit_pois_date <- glmmTMB(count ~ season*urbanization + trap_type + (1 | site_name/collection_date),
                    family = poisson(link = "log"),
                    data = tarsalis)
summary(fit_pois_date)

# Look at the residuals
simulateResiduals(fit_pois_date, plot = T)
```

I think it still doesn't look great... KS test shows significant deviation, dispersion test shows significant deviation, and there is a funnel shape. Maybe the negative binomial distribution can help with this as it helped with the salamanders? But this is all beyond me... give it a shot!

```{r nbinom-log-link-date}

fit_nbinom <- glmmTMB(count ~ season*urbanization + trap_type + (1 | site_name/collection_date),
                    family = nbinom1(link = "log"),
                    data = tarsalis)

simulateResiduals(fit_nbinom, plot = T)
```
  
I think it looks the same!!! So, stick with the poisson log-link? But I do think that random effects (1 | site_name/collection_date) looked better than disease_week. Not sure if that is good justification to use collection_date though.

## Marginal effects across seasons

Sticking with model with poisson(link="log"), using random effects (1 | site_name/collection_date) `fit_pois_date`

glmmTMB(count ~ season*urbanization + trap_type + (1 | site_name/collection_date),
                    family = nbinom1(link = "log"),
                    data = tarsalis)

```{r tar-marginal}
# First, just look at the effect of season, doing a 1-way pairwise comparison
em_season_tar = emmeans(fit_pois_date, pairwise ~ season, type = "response")
em_season_tar

cl = cld(em_season_tar, Letters = letters)

ggplot(data = cl, aes(x = season, y = rate)) +
    geom_point(size=2.5, color="black") +
    geom_errorbar(aes(x=season, ymin = asymp.LCL,
                      ymax = asymp.UCL),
                  width = 0.2, size=1, color="black") +
    geom_text(aes(label = gsub(" ", "", .group)),
              position = position_nudge(x = 0.3)) +
    ggeasy::easy_remove_axes("both", "title")

# Next, let's see how the effects vary with different mined levels in a 2-way emmeans comparison
em_spp2 = emmeans(fit_pois_date, pairwise ~ season | urbanization, type = "response")
em_spp2

cl2 <- cld(em_spp2$emmeans, Letters = letters)

ggplot(data = cl2, aes(x = season, y = rate)) +
    geom_point(size = 2.5, color = "black") +
    geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL),
                  width = 0.2, size = 1, color = "black") +
    facet_wrap(~ urbanization, labeller = label_both) +
    geom_text(aes(label = gsub(" ", "", .group)),
              position = position_nudge(x = 0.4)) +
    labs(
        title = "Seasonal abundance of Culex tarsalis across urbanization classes",
        x = "Season",
        y = "Predicted abundance"
    ) +
    ggeasy::easy_remove_axes("y", "title") +
    theme(plot.title = element_text(hjust = 0.5))

```

## Marginal effects across urbanization
```{r tar-marginal-urbanization}
# First, just look at the effect of season, doing a 1-way pairwise comparison
em_urban_tar = emmeans(fit_pois_date, pairwise ~ season, type = "response")
em_urban_tar

cl = cld(em_urban_tar, Letters = letters)

ggplot(data = cl, aes(x = season, y = rate)) +
    geom_point(size=2.5, color="black") +
    geom_errorbar(aes(x=season, ymin = asymp.LCL,
                      ymax = asymp.UCL),
                  width = 0.2, size=1, color="black") +
    geom_text(aes(label = gsub(" ", "", .group)),
              position = position_nudge(x = 0.3)) +
    ggeasy::easy_remove_axes("both", "title")

# Next, let's see how the effects vary with different mined levels in a 2-way emmeans comparison
em_spp2 = emmeans(fit_pois_date, pairwise ~ urbanization | season, type = "response")
em_spp2

cl2 <- cld(em_spp2$emmeans, Letters = letters)

ggplot(data = cl2, aes(x = urbanization, y = rate)) +
    geom_point(size = 2.5, color = "black") +
    geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL),
                  width = 0.2, size = 1, color = "black") +
    facet_wrap(~ season, labeller = label_both) +
    geom_text(aes(label = gsub(" ", "", .group)),
              position = position_nudge(x = 0.4)) +
    labs(
        title = "Abundance of Culex tarsalis across season",
        x = "Urbanization",
        y = "Predicted abundance"
    ) +
    ggeasy::easy_remove_axes("y", "title") +
    theme(plot.title = element_text(hjust = 0.5))

```
# Culex pipiens
  
```{r pipiens}
## pipiens datasets from SLCMAD:
pipiens <- read.csv("../data/pipiens_2025.csv")

head(pipiens, 5)
table(pipiens$season, pipiens$urban_cat)

library(dplyr)

pipiens <- pipiens %>%
  mutate(
    urban_cat = trimws(tolower(urban_cat)),
    urbanization = factor(
      urban_cat,
      levels = c("rural", "peri", "urban"),
      labels = c("rural", "peri", "urban")
    ),
    season = factor(
      season,
      levels = c("early", "mid", "late"),
      labels = c("early", "mid", "late")
    )
  )

head(pipiens, 5)
```


# GLMM by season and urbanization, each mosquito species separately

Start by binning into discrete seasons: early, mid, late

```{r pip-seasons-bins}
table(pipiens$season, pipiens$disease_week)

table(pipiens$season, pipiens$urbanization)
```
  
First, look at the response distribution
```{r pip-response}
# sub in pipiens dataset here instead of salamander 
hist(pipiens$count) 
hist(log(pipiens$count))

```

## Random effects (1 | site_name/disease_week)

Fit the model with poisson(link="log"), using random effects (1 | site_name/disease_week)

```{r pip-poisson-log-link-week}
# pipiens count data, starting with poisson model:
fit_pois <- glmmTMB(count ~ season*urbanization + trap_type + (1 | site_name/disease_week),
                    family = poisson(link = "log"),
                    data = pipiens)
summary(fit_pois)
```

Look at the residuals:
```{r pip-resids-poisson-log-link-week}
simulateResiduals(fit_pois, plot = T)
```
## Random effects (1 | site_name/collection_date)

Fit the model with poisson(link="log"), using random effects (1 | site_name/collection_date)

```{r pip-poisson-log-link-date}
# pipiens count data, starting with poisson model:
fit_pois_date <- glmmTMB(count ~ season*urbanization + trap_type + (1 | site_name/collection_date),
                    family = poisson(link = "log"),
                    data = pipiens)
summary(fit_pois_date)

# Look at the residuals
simulateResiduals(fit_pois_date, plot = T)
```

I think it still doesn't look great... KS test shows significant deviation, dispersion test shows significant deviation, and there is a funnel shape. Maybe the negative binomial distribution can help with this as it helped with the salamanders? But this is all beyond me... give it a shot!

```{r pip-nbinom-log-link-date}
fit_nbinom <- glmmTMB(count ~ season*urbanization + trap_type + (1 | site_name/collection_date),
                    family = nbinom1(link = "log"),
                    data = pipiens)

simulateResiduals(fit_nbinom, plot = T)
```
  
I think it looks the same!!! So, stick with the poisson log-link? But I do think that random effects (1 | site_name/collection_date) looked better than disease_week. Not sure if that is good justification to use collection_date though.

## Marginal effects

Sticking with model with poisson(link="log"), using random effects (1 | site_name/collection_date) `fit_pois_date`

glmmTMB(count ~ season*urbanization + trap_type + (1 | site_name/collection_date),
                    family = nbinom1(link = "log"),
                    data = pipiens)

```{r pip-marginal}
# First, just look at the effect of season, doing a 1-way pairwise comparison
em_season_pip = emmeans(fit_pois_date, pairwise ~ season, type = "response")
em_season_pip

cl = cld(em_season_pip, Letters = letters)

ggplot(data = cl, aes(x = season, y = rate)) +
    geom_point(size=2.5, color="black") +
    geom_errorbar(aes(x=season, ymin = asymp.LCL,
                      ymax = asymp.UCL),
                  width = 0.2, size=1, color="black") +
    geom_text(aes(label = gsub(" ", "", .group)),
              position = position_nudge(x = 0.3)) +
    ggeasy::easy_remove_axes("both", "title")

# Next, let's see how the effects vary in a 2-way emmeans comparison
em_spp2 = emmeans(fit_pois_date, pairwise ~ season | urbanization, type = "response")
em_spp2

cl2 <- cld(em_spp2$emmeans, Letters = letters)

ggplot(data = cl2, aes(x = season, y = rate)) +
    geom_point(size = 2.5, color = "black") +
    geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL),
                  width = 0.2, size = 1, color = "black") +
    facet_wrap(~ urbanization, labeller = label_both) +
    geom_text(aes(label = gsub(" ", "", .group)),
              position = position_nudge(x = 0.4)) +
    labs(
        title = "Seasonal abundance of Culex pipiens across urbanization classes",
        x = "Season",
        y = "Predicted abundance"
    ) +
    ggeasy::easy_remove_axes("y", "title") +
    theme(plot.title = element_text(hjust = 0.5))

```
  

## Marginal effects across urbanization
```{r pip-marginal-urbanization}
# First, just look at the effect of season, doing a 1-way pairwise comparison
em_urban_pip = emmeans(fit_pois_date, pairwise ~ season, type = "response")
em_urban_pip

cl = cld(em_urban_pip, Letters = letters)

ggplot(data = cl, aes(x = season, y = rate)) +
    geom_point(size=2.5, color="black") +
    geom_errorbar(aes(x=season, ymin = asymp.LCL,
                      ymax = asymp.UCL),
                  width = 0.2, size=1, color="black") +
    geom_text(aes(label = gsub(" ", "", .group)),
              position = position_nudge(x = 0.3)) +
    ggeasy::easy_remove_axes("both", "title")

# Next, let's see how the effects vary with different mined levels in a 2-way emmeans comparison
em_spp2 = emmeans(fit_pois_date, pairwise ~ urbanization | season, type = "response")
em_spp2

cl2 <- cld(em_spp2$emmeans, Letters = letters)

ggplot(data = cl2, aes(x = urbanization, y = rate)) +
    geom_point(size = 2.5, color = "black") +
    geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL),
                  width = 0.2, size = 1, color = "black") +
    facet_wrap(~ season, labeller = label_both) +
    geom_text(aes(label = gsub(" ", "", .group)),
              position = position_nudge(x = 0.4)) +
    labs(
        title = "Abundance of Culex pipiens across season",
        x = "Urbanization",
        y = "Predicted abundance"
    ) +
    ggeasy::easy_remove_axes("y", "title") +
    theme(plot.title = element_text(hjust = 0.5))

```
