---
title: "GAM **Cx. pipiens** and **Cx. tarsalis**: SLC 2025 field season"
author: "Norah Saarman"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

# Setup
```{r 1-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  # Saves PNGs to the root /figures folder
  fig.path = "../figures/" 
)
```

**Research Topic:** testing whether habitat and seasonal partitioning between Culex pipiens s.l. and Culex tarsalis shapes West Nile Virus (WNV) dynamics across urban–rural gradients. 

**Core hypothesis:** early/mid-season amplification dominated by pipiens in urban areas, later spillover involving tarsalis moving into urban/peri-urban areas.

**Approach:** Preliminary results visualized via mapping, with species identity and abundance as primary response variables. Model mosquito abundance and proportions using GLMM with GAM smoothing:

count ~ season*urbanization + trap_type + (1|site/date), family = poisson(link = "log"):   

 - Response variable = mosquito abundance  
 - Predictors = season*urbanization  
 - The trap type could be important, so we will add that as a fixed effect (covariate)... is this correct? We do think that the response variable of count of mosquitoes depends on trap type, since tarsalis seems to be more attracted to CO2 than pipiens, and we want to quantify that effect. Note that poisson model does not give a fixed offset (due to the log link)... The structure of this model means that it will estimate an effect that scales with the total number of mosquitos caught, which is exactly what we want.  
 - The data are grouped into sites and are also linked through time, so we'll add those as random effects. I think the sites should be coded as factors, **but I'm not sure what format to use for the date. I think it should be disease week so that week 18 is treated closer to 19 than 20, etc., but I'm not totally confident in this.** 
 - The family = poisson (link = "log")... why again?  
 
**For simple model:** 
count ~ disease_week*urbanization + (1|site/date), family = poisson(link = "log")


Load libraries
```{r}
library(tidyverse) # for data wrangling
library(glmmTMB)   # for model fitting
library(DHARMa)    # for residual plots
library(mgcViz)    # for residual plots
library(emmeans)   # for estimating marginal effects
library(multcomp)  # for statistical comparisons on fitted models
library(dplyr)     # for mutating dataframe to change labels in dataset
library(mgcv)      # fits GAM
library(broom)
library(ggplot2)
```
  
# Prepare Data

## Combined data 
```{r}
## tarsalis datasets from SLCMAD:
tarsalis <- read.csv("../data/tarsalis_2025.csv")
## pipiens datasets from SLCMAD:
pipiens <- read.csv("../data/pipiens_2025.csv")
## combine

combined <- bind_rows(tarsalis, pipiens)

## Set factor levels
combined <- combined %>%
  mutate(
    species = factor(
      species,
      levels = c("Culex pipiens", "Culex tarsalis")
    ),
    urban_cat = trimws(tolower(urban_cat)),
    urbanization = factor(
      urban_cat,
      levels = c("rural", "peri", "urban")
    ),
    season = factor(
      season,
      levels = c("early", "mid", "late")
    )
  )

#check
table(combined$species)
table(combined$season, combined$species)

combined <- combined %>%
  mutate(
    species = factor(species),
    urbanization = factor(urbanization),
    trap_type = factor(trap_type),
    site_name = factor(site_name),
    disease_week = as.numeric(disease_week)
  )
```
In the GLMM, (1 | site_name/collection_date), which handles clustering of repeated observations taken at the same site on the same date with:
- a random intercept for site_name  
- and a random intercept for each site_name:collection_date combination

GAM equivalent, in mgcv, the closest analogue is to create an interaction ID and include it as another random-effect smooth.

First create the grouping variable:

```{r grouping}
combined <- combined %>%
  mutate(
    site_date = interaction(site_name, collection_date, drop = TRUE)
  )
```
Check if the grouping variable occurs often:
```{r group-check}
length(unique(combined$site_date))
nrow(combined)
table(table(combined$site_date))
```
Yes, more than half of the observations are impacted... but many of them are different trap-types. This is a decision to make, to include or not to include? Site_date adds shared noise from shared environment within a sampling event. Since we really care more about ecological patterns over time, and are already including trap-type, is it really needed? Does it change the result?

Let's start with a simple comparison of including ONLY site_name.

**NOTE:** I'm also worried that we might need to use the * trap_type to fully capture species-specific trap effects.

# Single model GAM with combined by species (random effect = site name)

## Fit single GAM
```{r fit-gam-combined-by-spp}

# Model 1: Single spline
# Models a universal pattern in time and allows the magnitude to vary across traps.
gam_offset <- gam(
  count ~ species * trap_type +
    s(disease_week, k = 15) + 
    s(site_name, bs = "re"),
  data = combined,
  family = nb(link = "log"),
  method = "REML"
)

# Model 2: factor spline (bs = "fs")
# Fits variable patterns for each group pooled toward a universal pattern, and you don't include a "by =" argument
# The second models a universal pattern plus trap-specific deviations from that pattern.
gam_variable_shape <- gam(
  count ~ species * trap_type +
    s(disease_week, k = 15) + 
    s(disease_week, species, bs = "fs", k = 15) + 
    s(site_name, bs = "re"),
  data = combined,
  family = nb(link = "log"),
  method = "REML"
)

# Model 3: Independent splines that are not pooled then you use the by = argument to specify the grouping and don't need to include the "bs =" part.
gam_by <- gam(
  count ~ species * trap_type +
    s(disease_week, by = species, k = 15) + # splines not pooled
    s(site_name, bs = "re"),
  data = combined,
  family = nb(link = "log"),
  method = "REML"
)

AIC(
  gam_offset,
  gam_variable_shape,
  gam_by
)

# Model 3 with independent splines wins
```

```{r what-k}
# Now what k value?

# Fit GAM model with site_name only
gam_by_k10 <- gam(
  count ~ species *  trap_type +
    s(disease_week, by = species, k = 10, m = 2) +
    s(site_name, bs = "re"),
  data = combined,
  family = nb(link="log"),
  method = "REML"
)

# Fit GAM model with site_name only
gam_by_k15 <- gam(
  count ~ species * trap_type +
    s(disease_week, by = species, k = 15, m = 2) +
    s(site_name, bs = "re"),
  data = combined,
  family = nb(link="log"),
  method = "REML"
)

AIC(gam_by_k10, gam_by_k15)

```
  
## Check Fit and Plot Smooths
```{r check-gam-site-name-only}
#Summary of GAM fit
summary(gam_by_k15)

#Check if smooths are hitting their basis limits
gam.check(gam_by_k15)
```
K 15 is slightly better

Considering that the best AIC is when we use "by" and have totally independent fits, I think its best to move forward with species-specific gams.


# Trap_type: Urban only, with paired trap types:

"Within Urban sites, does trap type affect abundance? fit the model on Urban observations only"

Analyze only sites with both trap types present to avoid confounding trap effects with site/urbanization effects.
  
```{r identify-pairs}

# Identify sites with BOTH trap types

paired_sites <- combined %>%
  group_by(site_name) %>%
  summarize(
    n_traps = n_distinct(trap_type),
    .groups = "drop"
  ) %>%
  filter(n_traps == 2) %>%
  pull(site_name)

paired_sites

```
  

## Cx. pipiens abundance by trap type Urban

```{r paired-urban-pip}
# pull out pipiens
pipiens <- combined[combined$species == "Culex pipiens",]
head(pipiens$site_name)

urban_pip <- pipiens %>%
  filter(
    urbanization == "urban",
    site_name %in% paired_sites
  )

table(urban_pip$trap_type)
length(unique(urban_pip$site_name))


# Model 1: Single spline
# Models a universal pattern in time and allows the magnitude to vary across traps.
urban_pip_gam_offset <- gam(
  count ~ trap_type + 
    s(disease_week, k = 15) + 
    s(site_name, bs = "re"),
  data = urban_pip,
  family = nb(link = "log"),
  method = "REML"
)

# Model 2: factor spline (bs = "fs")
# Fits variable patterns for each group pooled toward a universal pattern, and you don't include a "by =" argument
# The second models a universal pattern plus trap-specific deviations from that pattern.
urban_pip_gam_variable_shape <- gam(
  count ~ trap_type + 
    s(disease_week, k = 15) + 
    s(disease_week, trap_type, bs = "fs", k = 15) + 
    s(site_name, bs = "re"),
  data = urban_pip,
  family = nb(link = "log"),
  method = "REML"
)

# Model 3: Independent splines that are not pooled then you use the by = argument to specify the grouping and don't need to include the "bs =" part.
urban_pip_gam_by <- gam(
  count ~ trap_type +
    s(disease_week, by = trap_type, k = 15) + # splines not pooled
    s(site_name, bs = "re"),
  data = urban_pip,
  family = nb(link = "log"),
  method = "REML"
)

AIC(
  urban_pip_gam_offset,
  urban_pip_gam_variable_shape,
  urban_pip_gam_by
)

summary(urban_pip_gam_variable_shape)

```
For Cx. pipiens, I think Model 1 with offset spline is best, but not too different from Model 2, so since it overlaps with Cx. tar's best, should we go with that?

## Cx. tarsalis abundance by trap type Urban

```{r paired-urban-tar}
# pull out tarsalis
tarsalis <- combined[combined$species == "Culex tarsalis",]
head(tarsalis$site_name)

urban_tar <- tarsalis %>%
  filter(
    urbanization == "urban",
    site_name %in% paired_sites
  )

table(urban_tar$trap_type)
length(unique(urban_tar$site_name))


# Model 1: Single spline
# Models a universal pattern in time and allows the magnitude to vary across traps.
urban_tar_gam_offset <- gam(
  count ~ trap_type + 
    s(disease_week, k = 15) + 
    s(site_name, bs = "re"),
  data = urban_tar,
  family = nb(link="log"),
  method = "REML"
)

# Model 2: factor spline (bs = "fs")
# Fits variable patterns for each group pooled toward a universal pattern, and you don't include a "by =" argument
# The second models a universal pattern plus trap-specific deviations from that pattern.
urban_tar_gam_variable_shape <- gam(
  count ~ trap_type + 
    s(disease_week, k = 15) + 
    s(disease_week, trap_type, bs = "fs", k = 15) + 
    s(site_name, bs = "re"),
  data = urban_tar,
  family = nb(link="log"),
  method = "REML"
)

# Model 3: Independent splines that are not pooled then you use the by = argument to specify the grouping and don't need to include the "bs =" part.
urban_tar_gam_by <- gam(
  count ~ trap_type +
    s(disease_week, by = trap_type, k = 15) + # splines not pooled
    s(site_name, bs = "re"),
  data = urban_tar,
  family = nb(link="log"),
  method = "REML"
)

AIC(
  urban_tar_gam_offset,
  urban_tar_gam_variable_shape,
  urban_tar_gam_by
)

summary(urban_tar_gam_variable_shape)
```

For Cx. tarsalis, Model 3 with independent splines is best, but not too different from Model 2, so since it matches Cx. pip's best, should we go with that?

## Smooths plots by trap type

Plot with a secondary y-axis with a scaling factor between trap types. This is easiest if you rescale only the GRVD predictions for plotting, but label the right axis back in GRVD units.

```{r smooths}
# Colors by species, same as before
cols <- c(
  "Culex pipiens"  = "#1bc8ea",
  "Culex tarsalis" = "#FF2DA0"
)

# Helper function for urban trap-type GAMs
predict_urban_trap_gam <- function(model, data, species_label) {
  
  newdata <- expand.grid(
    disease_week = seq(
      min(data$disease_week, na.rm = TRUE),
      max(data$disease_week, na.rm = TRUE),
      by = 1
    ),
    trap_type = levels(data$trap_type),
    site_name = levels(data$site_name)[1]
  )
  
  newdata$trap_type <- factor(newdata$trap_type, levels = levels(data$trap_type))
  newdata$site_name <- factor(newdata$site_name, levels = levels(data$site_name))
  
  pred <- predict(
    model,
    newdata = newdata,
    type = "link",
    se.fit = TRUE,
    exclude = "s(site_name)"
  )
  
  newdata %>%
    mutate(
      species = species_label,
      fit_link = pred$fit,
      se_link = pred$se.fit,
      fit = exp(fit_link),
      lower = exp(fit_link - 1.96 * se_link),
      upper = exp(fit_link + 1.96 * se_link)
    )
}

# Predictions from both urban models
pred_urban_trap <- bind_rows(
  predict_urban_trap_gam(
    urban_pip_gam_variable_shape,
    urban_pip,
    "Culex pipiens"
  ),
  predict_urban_trap_gam(
    urban_tar_gam_variable_shape,
    urban_tar,
    "Culex tarsalis"
  )
)


# Split predictions by trap type to calculate scale factor
co2_range <- pred_urban_trap %>%
  filter(trap_type == "CO2") %>%
  summarize(max_fit = max(upper, na.rm = TRUE)) %>%
  pull(max_fit)

grvd_range <- pred_urban_trap %>%
  filter(trap_type == "GRVD") %>%
  summarize(max_fit = max(upper, na.rm = TRUE)) %>%
  pull(max_fit)

scale_factor <- co2_range / grvd_range

scale_by_species <- pred_urban_trap %>%
  group_by(species) %>%
  summarize(
    co2_max = max(upper[trap_type == "CO2"], na.rm = TRUE),
    grvd_max = max(upper[trap_type == "GRVD"], na.rm = TRUE),
    scale_factor = co2_max / grvd_max,
    .groups = "drop"
  )

pred_urban_trap_plot <- pred_urban_trap %>%
  left_join(scale_by_species, by = "species") %>%
  mutate(
    fit_plot   = ifelse(trap_type == "GRVD", fit * scale_factor, fit),
    lower_plot = ifelse(trap_type == "GRVD", lower * scale_factor, lower),
    upper_plot = ifelse(trap_type == "GRVD", upper * scale_factor, upper)
  )

library(dplyr)
library(ggplot2)
library(patchwork)

plot_species_panel <- function(pred_df, species_name, show_legend = TRUE) {
  
  dat <- pred_df %>%
    filter(species == species_name)
  
  scale_factor <- dat %>%
    summarize(
      co2_max = max(upper[trap_type == "CO2"], na.rm = TRUE),
      grvd_max = max(upper[trap_type == "GRVD"], na.rm = TRUE),
      scale_factor = co2_max / grvd_max
    ) %>%
    pull(scale_factor)
  
  dat <- dat %>%
    mutate(
      fit_plot   = ifelse(trap_type == "GRVD", fit * scale_factor, fit),
      lower_plot = ifelse(trap_type == "GRVD", lower * scale_factor, lower),
      upper_plot = ifelse(trap_type == "GRVD", upper * scale_factor, upper)
    )
  
  ggplot(
    dat,
    aes(
      x = disease_week,
      y = fit_plot,
      color = species,
      fill = species,
      linetype = trap_type
    )
  ) +
    geom_ribbon(
      aes(
        ymin = lower_plot,
        ymax = upper_plot,
        group = interaction(species, trap_type)
      ),
      alpha = 0.18,
      color = NA
    ) +
    geom_line(
  aes(
    group = interaction(species, trap_type),
    linewidth = trap_type
  )
) +
scale_linewidth_manual(
  values = c(
    "CO2" = 0.8,
    "GRVD" = 1.5
  ),
  guide = "none"
) +
    geom_vline(
      xintercept = 33,
      linetype = "dashed",
      color = "black",
      linewidth = 0.5
    ) +
    annotate(
      "text",
      x = 33,
      y = Inf,
      label = "1st WNV cases",
      vjust = 1,
      hjust = -0.07,
      size = 3
    ) +
    scale_color_manual(
      values = cols,
      limits = names(cols),
      drop = FALSE
    ) +
    scale_fill_manual(
      values = cols,
      limits = names(cols),
      drop = FALSE
    ) +
    scale_y_continuous(
      name = "Predicted: CO2",
      sec.axis = sec_axis(
        trans = ~ . / scale_factor,
        name = "Predicted: GRVD"
      )
    ) +
    guides(
      color = guide_legend(
        order = 1,
        override.aes = list(
          linetype = "solid",
          linewidth = 1.2,
          alpha = .75
        )
      ),
      fill = "none",
      linetype = guide_legend(order = 2)
    ) +
    labs(
      title = NULL,
      x = "Disease week",
      color = "Species",
      fill = "Species",
      linetype = "Trap type"
    ) +
    theme_classic() +
    theme(
      legend.position = ifelse(show_legend, "right", "none"),
      plot.title = element_blank(),
      strip.background = element_blank(),
      strip.text = element_text(size = 12)
    )
}

plot_pip <- plot_species_panel(
  pred_urban_trap,
  "Culex pipiens",
  show_legend = TRUE
)

plot_tar <- plot_species_panel(
  pred_urban_trap,
  "Culex tarsalis",
  show_legend = FALSE
)

combined_plot <- wrap_plots(
  plot_pip,
  plot_tar,
  ncol = 1
) +
  plot_annotation(
    title = "Urban predicted count by trap type"
  ) &
  theme(
    plot.title = element_text(hjust = 0.5)
  )

combined_plot
```

  
## Interpretation

**Cx. pipiens:**

-   Estimated GRVD effect was -0.72 in the paired urban-only model

-   GRVD traps caught approximately half of the abundance caught in CO2 traps

    -   exp(-0.72) = 0.49

-   The shape of the seasonal smooth was very very similar among trap types, so I think using the same smooth for both trap types (trap type = fixed effect), works well for Cx. pipiens.

**Cx. tarsalis:**

-   Estimated GRVD effect was -2.6898 in paired urban-only model.

-   GRVD traps caught approximately 7% of the abundance caught in CO2 traps.

    -   exp(-2.6898) = 0.0678 (6.78%)

-   It also looks like for tarsalis, the GRVD trap counts being low enough to potentially not reliable capture the full seasonal smooth, so we should try to visualize with CO2 traps wherever possible.

# Species-Specific GAM 

separate models for each species, count by urbanization + trap_type (random effect = site name)

## Cx. pipiens GAM
```{r fit-gam-separated-spp}
# pull out pipiens
pipiens <- combined %>%
  filter(species == "Culex pipiens") %>%
  droplevels()


# Model 2: factor spline (bs = "fs")
# Fits variable patterns for each group pooled toward a universal pattern, and you don't include a "by =" argument
# The second models a universal pattern plus trap-specific deviations from that pattern.
pip_gam <- gam(
  count ~ urbanization + trap_type + 
    # s(disease_week, k = 15) +   ## Remove in final?
    s(disease_week, urbanization, bs = "fs", k = 15, m = 3) + 
    s(site_name, bs = "re"),
  data = pipiens,
  family = nb(link = "log"),
  method = "REML"
)

```
### Check smooths: pipiens
```{r check-gam-pip}
#Summary of GAM fit
summary(pip_gam)

#AIC for 
cat("GAM model pip AIC: ", AIC(pip_gam), "\n")

#Check if smooths are hitting their basis limits
gam.check(pip_gam)

# plot the smooths for pip
plot(pip_gam, select = 1, shade = TRUE, main = "GAM Smooth for pip x rural")
plot(pip_gam, select = 2, shade = TRUE, main = "GAM Smooth for pip x peri")
plot(pip_gam, select = 3, shade = TRUE, main = "GAM Smooth for pip x urban")
```


### Map residuals: pipiens
```{r spatial-plot-residuals-pip}

# Make a list of coordinates
site_coords_pip <- pipiens %>%
  dplyr::select(site_name, latitude, longitude) %>%
  distinct()

# Extract data actually used in the model
# Add coordinates and residuals
pipiens_data_res <- model.frame(pip_gam) %>%
  mutate(resid = residuals(pip_gam, type = "pearson")) %>%
  left_join(site_coords_pip, by = "site_name")

# Transform to make it easier to see spatial patterns
pipiens_data_res$resid_log <- log10(pipiens_data_res$resid + 2)

# plot by lat/long
ggplot(pipiens_data_res, aes(x = longitude, y = latitude, color = resid_log)) +
  geom_point(size = 3) +
  coord_fixed() +
  scale_color_viridis_c() +
  theme_minimal()

```
  
### Weekly residuals: pipiens  
```{r extract-residuals-pip}

# Plot residuals by disease week

ggplot(pipiens_data_res,
       aes(x = disease_week, y = resid)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se = FALSE, color = "blue") +
  theme_bw()

#residual distributions for each week separately
ggplot(pipiens_data_res,
       aes(x = factor(disease_week), y = resid)) +
  geom_boxplot() +
  theme_bw() +
  labs(
    x = "Disease Week",
    y = "Pearson Residuals",
    title = "Residual Distribution by Week - pipiens"
  )
```

  
### DHARMa: pip

```{r dharma-pip}

# pipiens
sim_pip <- simulateResiduals(pip_gam, n = 1000)
plot(sim_pip)

testDispersion(sim_pip)

# Aggregate DHARMa residuals by site for pipiens
sim_pip_site <- recalculateResiduals(
  sim_pip,
  group = pipiens$site_name
)

site_coords_pip <- pipiens %>%
  dplyr::select(site_name, latitude, longitude) %>%
  distinct() %>%
  arrange(site_name)

testSpatialAutocorrelation(
  sim_pip_site,
  x = site_coords_pip$longitude,
  y = site_coords_pip$latitude
)

```

  
  
## Cx. tarsalis GAM
```{r tarsalis}
# pull out tarsalis
tarsalis <- combined %>%
  filter(species == "Culex tarsalis") %>%
  droplevels() 

# Model 2: factor spline (bs = "fs")
# Fits variable patterns for each group pooled toward a universal pattern, and you don't include a "by =" argument
# The second models a universal pattern plus trap-specific deviations from that pattern.
tar_gam <- gam(
  count ~ urbanization + trap_type + 
    # s(disease_week, k = 15) +   ## Remove in final?
    s(disease_week, urbanization, bs = "fs", k = 15, m = 3) + 
    s(site_name, bs = "re"),
  data = tarsalis,
  family = nb(link = "log"),
  method = "REML"
)

```
    
### Check smooths: tarsalis
```{r check-gam-tar}
#Summary of GAM fit
summary(tar_gam)

#AIC for 
cat("GAM model tar AIC: ", AIC(tar_gam), "\n")

#Check if smooths are hitting their basis limits
gam.check(tar_gam)

# plot the smooths for tar
plot(tar_gam, select = 1, shade = TRUE, main = "GAM Smooth for tar x rural")
plot(tar_gam, select = 2, shade = TRUE, main = "GAM Smooth for tar x peri")
plot(tar_gam, select = 3, shade = TRUE, main = "GAM Smooth for tar x urban")
```
  
### Map residuals: tarsalis
```{r spatial-plot-residuals-tar}

# Make a list of coordinates
site_coords_tar <- tarsalis %>%
  dplyr::select(site_name, latitude, longitude) %>%
  distinct()

# Extract data actually used in the model
# Add coordinates and residuals
tarsalis_data_res <- model.frame(tar_gam) %>%
  mutate(resid = residuals(tar_gam, type = "pearson")) %>%
  left_join(site_coords_tar, by = "site_name")

# Transform to make it easier to see spatial patterns
tarsalis_data_res$resid_log <- log10(tarsalis_data_res$resid + 2)

# plot by lat/long
ggplot(tarsalis_data_res, aes(x = longitude, y = latitude, color = resid_log)) +
  geom_point(size = 3) +
  coord_fixed() +
  scale_color_viridis_c() +
  theme_minimal()

```

### Weekly residuals: tarsalis
```{r extract-residuals-tar}
 

# Plot residuals by disease week

ggplot(tarsalis_data_res,
       aes(x = disease_week, y = resid)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se = FALSE, color = "blue") +
  theme_bw()

#residual distributions for each week separately
ggplot(tarsalis_data_res,
       aes(x = factor(disease_week), y = resid)) +
  geom_boxplot() +
  theme_bw() +
  labs(
    x = "Disease Week",
    y = "Pearson Residuals",
    title = "Residual Distribution by Week - tarsalis"
  )
```
  
### DHARMa: tarsalis

```{r dharma-tar}
# tarsalis
sim_tar <- simulateResiduals(tar_gam, n = 1000)
plot(sim_tar)

testDispersion(sim_tar)

# Aggregate DHARMa residuals by site for tarsalis
sim_tar_site <- recalculateResiduals(
  sim_tar,
  group = tarsalis$site_name
)

site_coords_tar <- tarsalis %>%
  dplyr::select(site_name, latitude, longitude) %>%
  distinct() %>%
  arrange(site_name)

testSpatialAutocorrelation(
  sim_tar_site,
  x = site_coords_tar$longitude,
  y = site_coords_tar$latitude
)


# check whether very large counts are dominating the fit:
ggplot(tarsalis, aes(x = count)) +
  geom_histogram(bins = 50) +
  scale_x_log10() +
  theme_bw()

# Is underdispersion trap-specific?
tarsalis_data_res <- model.frame(tar_gam) %>%
  mutate(
    resid = residuals(tar_gam, type = "pearson"),
    fitted = fitted(tar_gam)
  )

ggplot(tarsalis_data_res, aes(x = trap_type, y = resid)) +
  geom_boxplot() +
  theme_bw()

ggplot(tarsalis_data_res, aes(x = urbanization, y = resid)) +
  geom_boxplot() +
  theme_bw()

tarsalis_data_res <- model.frame(tar_gam) %>%
  mutate(
    fitted = fitted(tar_gam)
  )

```
  
Something strange is happening with tarsalis model, with failed dispersion test when we use.DHARMa nonparametric dispersion test via sd of residuals fitted vs. simulated:   
  -   smooth: m=3, k=25  
  -   data:  simulationOutput  
  -   dispersion = 0.48346, p-value < 2.2e-16  
  -   alternative hypothesis: two.sided  

Is this DHARMa is reacting to strong structured signal
repeated measures, high explanatory power?

# Plot pip vs. tar from separate models 
  
### Effect size (forest) plot from separate models

```{r effect-2-models}
coef_pip <- tidy(pip_gam, parametric = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(species = "Culex pipiens")

coef_tar <- tidy(tar_gam, parametric = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(species = "Culex tarsalis")

coef_df_sep <- bind_rows(coef_pip, coef_tar) %>%
  mutate(
    effect = exp(estimate),
    lower = exp(estimate - 1.96 * std.error),
    upper = exp(estimate + 1.96 * std.error),
    term_clean = recode(
      term,
      "urbanizationperi" = "Peri relative to rural",
      "urbanizationurban" = "Urban relative to rural",
      "trap_typeGRVD" = "GRVD relative to CO2"
    )
  )

ggplot(coef_df_sep, aes(x = effect, y = term_clean, color = species)) +
  geom_point(position = position_dodge(width = 0.6), size = 3) +
  geom_errorbarh(
    aes(xmin = lower, xmax = upper),
    position = position_dodge(width = 0.6),
    height = 0.2
  ) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_x_log10() +
  scale_color_manual(values = cols) +
  labs(
    x = "Multiplicative effect on abundance (log scale)",
    y = "",
    color = "Species",
    title = "Effect sizes from species-specific GAMs"
  ) +
  theme_minimal()
```
   
### Seasonal Smooth from all models
```{r viz-2-gams}

cols <- c(
  "Culex pipiens"  = "#1bc8ea",
  "Culex tarsalis" = "#FF2DA0"
)

predict_species_gam <- function(model, newdata, species_label) {
  
  newdata$urbanization <- factor(newdata$urbanization, levels = levels(combined$urbanization))
  newdata$trap_type <- factor(newdata$trap_type, levels = levels(combined$trap_type))
  newdata$site_name <- factor(newdata$site_name, levels = levels(combined$site_name))
  
  pred <- predict(
    model,
    newdata = newdata,
    type = "link",
    se.fit = TRUE,
    exclude = "s(site_name)"
  )
  
  newdata %>%
    mutate(
      species = species_label,
      fit_link = pred$fit,
      se_link = pred$se.fit,
      fit = exp(fit_link),
      lower = exp(fit_link - 1.96 * se_link),
      upper = exp(fit_link + 1.96 * se_link)
    )
}
# ----------------------
# 1. Seasonal effects across habitats
# ----------------------

newdata_season <- expand.grid(
  disease_week = seq(min(combined$disease_week), max(combined$disease_week), by = 1),
  urbanization = "rural",
  trap_type = "GRVD",
  site_name = levels(combined$site_name)[1]
)

pred_season <- bind_rows(
  predict_species_gam(pip_gam, newdata_season, "Culex pipiens"),
  predict_species_gam(tar_gam, newdata_season, "Culex tarsalis")
)

ggplot(pred_season, aes(x = disease_week, y = fit, color = species, fill = species)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = cols) +
  scale_fill_manual(values = cols) +
  labs(
    x = "Disease week",
    y = "Predicted abundance",
    color = "Species",
    fill = "Species",
    title = "Predicted seasonal abundance from species-specific GAMs"
  ) +
  theme_minimal()

# ----------------------
# 2. Predicted abundance by urbanization: CO2 traps
# ----------------------

newdat_site_CO2 <- expand.grid(
  disease_week = seq(min(combined$disease_week), max(combined$disease_week), by = 1),
  urbanization = levels(combined$urbanization),
  trap_type = "CO2",
  site_name = levels(combined$site_name)[1]
)

pred_site_CO2 <- bind_rows(
  predict_species_gam(pip_gam, newdat_site_CO2, "Culex pipiens"),
  predict_species_gam(tar_gam, newdat_site_CO2, "Culex tarsalis")
)

ggplot(pred_site_CO2, aes(x = disease_week, y = fit, color = species, group = species)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = species), alpha = 0.25, color = NA) +
  geom_vline(xintercept = 33, linetype = "dashed", color = "black", linewidth = 0.5) +
  annotate("text", x = 33, y = Inf, label = "1st WNV cases",
           vjust = 1, hjust = -0.07, size = 3) +
  facet_wrap(~ urbanization, scales = "free_y", ncol = 1) +
  scale_color_manual(values = cols) +
  scale_fill_manual(values = cols) +
  labs(
    title = "Predicted abundance from species-specific GAMs (CO2 traps)",
    x = "Disease week",
    y = "Predicted abundance",
    color = "Species",
    fill = "Species"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5),
    strip.background = element_blank(),
    strip.text = element_text(size = 12)
  )

# ----------------------
# 3. Plot predicted for pipiens/tarsalis on their own to allow pipiens to be visible
# ----------------------
# pipiens only:
ggplot(pred_site_CO2[pred_site_CO2$species=="Culex pipiens",], aes(x = disease_week, y = fit, color = species, group = species)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = species), alpha = 0.25, color = NA) +
  geom_vline(xintercept = 33, linetype = "dashed", color = "black", linewidth = 0.5) +
  annotate("text", x = 33, y = Inf, label = "1st WNV",
           vjust = 1, hjust = -0.07, size = 3) +
  facet_wrap(~ urbanization, scales = "free_y", ncol = 1) +
  scale_color_manual(values = cols) +
  scale_fill_manual(values = cols) +
  labs(
    title = "Predicted abundance from species-specific GAMs (CO2 traps)",
    x = "Disease week",
    y = "Predicted abundance",
    color = "Species",
    fill = "Species"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5),
    strip.background = element_blank(),
    strip.text = element_text(size = 12)
  )

# tarsalis only
ggplot(pred_site_CO2[pred_site_CO2$species=="Culex tarsalis",], aes(x = disease_week, y = fit, color = species, group = species)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = species), alpha = 0.25, color = NA) +
  geom_vline(xintercept = 33, linetype = "dashed", color = "black", linewidth = 0.5) +
  annotate("text", x = 33, y = Inf, label = "1st WNV",
           vjust = 1, hjust = -0.07, size = 3) +
  facet_wrap(~ urbanization, scales = "free_y", ncol = 1) +
  scale_color_manual(values = cols) +
  scale_fill_manual(values = cols) +
  labs(
    title = "Predicted abundance from species-specific GAMs (CO2 traps)",
    x = "Disease week",
    y = "Predicted abundance",
    color = "Species",
    fill = "Species"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5),
    strip.background = element_blank(),
    strip.text = element_text(size = 12)
  )

# ----------------------
# 4. Urbanization effect figure
# ----------------------
newdata_urban <- expand.grid(
  urbanization = levels(combined$urbanization),
  trap_type = "GRVD",
  disease_week = median(combined$disease_week, na.rm = TRUE),
  site_name = levels(combined$site_name)[1]
)

pred_urban <- bind_rows(
  predict_species_gam(pip_gam, newdata_urban, "Culex pipiens"),
  predict_species_gam(tar_gam, newdata_urban, "Culex tarsalis")
) %>%
  mutate(urbanization = factor(urbanization, levels = c("rural", "peri", "urban")))

ggplot(pred_urban, aes(x = urbanization, y = fit, color = species, group = species)) +
  geom_point(position = position_dodge(width = 0.3), size = 3) +
  geom_line(position = position_dodge(width = 0.3), linewidth = 1) +
  geom_errorbar(
    aes(ymin = lower, ymax = upper),
    position = position_dodge(width = 0.3),
    width = 0.2
  ) +
  scale_color_manual(values = cols) +
  labs(
    x = "Urbanization",
    y = "Predicted abundance",
    color = "Species",
    title = "Urbanization effects from species-specific GAMs"
  ) +
  theme_minimal()
# ----------------------
# 5. Trap type effect figure
# ----------------------

newdata_trap <- expand.grid(
  urbanization = levels(combined$urbanization),
  trap_type = levels(combined$trap_type),
  disease_week = median(combined$disease_week, na.rm = TRUE),
  site_name = levels(combined$site_name)[1]
)

pred_trap <- bind_rows(
  predict_species_gam(pip_gam, newdata_trap, "Culex pipiens"),
  predict_species_gam(tar_gam, newdata_trap, "Culex tarsalis")
) %>%
  mutate(urbanization = factor(urbanization, levels = c("rural", "peri", "urban")))

ggplot(pred_trap, aes(x = trap_type, y = fit, color = species, group = species)) +
  geom_point(position = position_dodge(width = 0.3), size = 3) +
  geom_line(position = position_dodge(width = 0.3), linewidth = 1) +
  geom_errorbar(
    aes(ymin = lower, ymax = upper),
    position = position_dodge(width = 0.3),
    width = 0.2
  ) +
  facet_wrap(~ urbanization) +
  scale_color_manual(values = cols) +
  labs(
    x = "Trap type",
    y = "Predicted abundance",
    color = "Species",
    title = "Trap type effects from species-specific GAMs"
  ) +
  theme_minimal()

```

### Plot relative abundance figure

```{r rel-abund-2-models}
newdata_rel <- expand.grid(
  urbanization = levels(combined$urbanization),
  disease_week = median(combined$disease_week, na.rm = TRUE),
  trap_type = "CO2",
  site_name = levels(combined$site_name)[1]
)

pred_rel <- bind_rows(
  predict_species_gam(pip_gam, newdata_rel, "Culex pipiens"),
  predict_species_gam(tar_gam, newdata_rel, "Culex tarsalis")
) %>%
  mutate(
    species = factor(species, levels = c("Culex pipiens", "Culex tarsalis")),
    urbanization = factor(urbanization, levels = c("rural", "peri", "urban"))
  ) %>%
  group_by(urbanization) %>%
  mutate(
    prop = fit / sum(fit),
    prop_lower = lower / sum(upper),
    prop_upper = upper / sum(lower),
    prop_lower = pmax(0, prop_lower),
    prop_upper = pmin(1, prop_upper)
  ) %>%
  ungroup()

ggplot(pred_rel, aes(x = urbanization, y = prop, color = species, group = species)) +
  geom_point(size = 4) +
  geom_line(linewidth = 1) +
  geom_errorbar(
    aes(ymin = prop_lower, ymax = prop_upper),
    width = 0.1
  ) +
  scale_color_manual(values = cols) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Urbanization",
    y = "Relative abundance",
    color = NULL,
    title = "Predicted relative abundance (CO2 traps)") +
  theme_classic() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5)
  )
```
   
