---
title: "GAM **Cx. pipiens** and **Cx. tarsalis** abundance: SLC 2025 field season"
author: "Norah Saarman"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Research Topic:** testing whether habitat and seasonal partitioning between Culex pipiens s.l. and Culex tarsalis shapes West Nile Virus (WNV) dynamics across urban–rural gradients. 

**Core hypothesis:** early/mid-season amplification dominated by pipiens in urban areas, later spillover involving tarsalis moving into urban/peri-urban areas.

**Approach:** Preliminary results visualized via mapping, with species identity and abundance as primary response variables. Model mosquito abundance and proportions using GLMMs count ~ season*urbanization + trap_type + (1|site/date), family = poisson(link = "log"):   
 - Response variable = mosquito abundance  
 - Predictors = season*urbanization  
 - The trap type could be important, so we will add that as a fixed effect (covariate)... is this correct? We do think that the response variable of count of mosquitoes depends on trap type, since tarsalis seems to be more attracted to CO2 than pipiens, and we want to quantify that effect. Note that poisson model does not give a fixed offset (due to the log link)... The structure of this model means that it will estimate an effect that scales with the total number of mosquitos caught, which is exactly what we want.  
 - The data are grouped into sites and are also linked through time, so we'll add those as random effects. I think the sites should be coded as factors, **but I'm not sure what format to use for the date. I think it should be disease week so that week 18 is treated closer to 19 than 20, etc., but I'm not totally confident in this.**   
 - The family = poisson (link = "log")... why again?  
 
**For simple model:** 
count ~ season*urbanization + (1|site/date), family = poisson(link = "log")


Load libraries
```{r}
library(tidyverse) # for data wrangling
library(glmmTMB)   # for model fitting
library(DHARMa)    # for residual plots
library(emmeans)   # for estimating marginal effects
library(multcomp)  # for statistical comparisons on fitted models
library(dplyr)     # for mutating dataframe to change labels in dataset

```
  
# Prepare Data

## Culex tarsalis data
  
```{r tarsalis}
## tarsalis datasets from SLCMAD:
tarsalis <- read.csv("../data/tarsalis_2025.csv")

tarsalis <- tarsalis %>%
  mutate(
    urbanization = factor(
      urban_cat,
      levels = c("rural", "peri", "urban"),
      labels = c("rural", "peri", "urban")
    ),
    season = factor(
      season,
      levels = c("early", "mid", "late"),
      labels = c("early", "mid", "late")
    )
  )

head(tarsalis, 5)
```

## Culex pipiens data
  
```{r pipiens}
## pipiens datasets from SLCMAD:
pipiens <- read.csv("../data/pipiens_2025.csv")

head(pipiens, 5)
table(pipiens$season, pipiens$urban_cat)

library(dplyr)

pipiens <- pipiens %>%
  mutate(
    urban_cat = trimws(tolower(urban_cat)),
    urbanization = factor(
      urban_cat,
      levels = c("rural", "peri", "urban"),
      labels = c("rural", "peri", "urban")
    ),
    season = factor(
      season,
      levels = c("early", "mid", "late"),
      labels = c("early", "mid", "late")
    )
  )

head(pipiens, 5)
```

## Combined data 
```{r}
## tarsalis datasets from SLCMAD:
tarsalis <- read.csv("../data/tarsalis_2025.csv")
## pipiens datasets from SLCMAD:
pipiens <- read.csv("../data/pipiens_2025.csv")
## combine
library(dplyr)
combined <- bind_rows(tarsalis, pipiens)

## Set factor levels
combined <- combined %>%
  mutate(
    species = factor(
      species,
      levels = c("Culex pipiens", "Culex tarsalis")
    ),
    urban_cat = trimws(tolower(urban_cat)),
    urbanization = factor(
      urban_cat,
      levels = c("rural", "peri", "urban")
    ),
    season = factor(
      season,
      levels = c("early", "mid", "late")
    )
  )

#check
table(combined$species)
table(combined$season, combined$species)
```
# GLMM by species * season (and species * season * urbanization) for visualization

## GLMM with species * season, including trap type:
```{r simple}
fit_combined <- glmmTMB(
  count ~ species * season + trap_type + (1 | site_name/collection_date),
  family = nbinom1(link = "log"),
  data = combined
)
```

Marginal Means:
```{r marginal-means}
library(emmeans)

em <- emmeans(fit_combined, ~ species * season, type = "response")
em_df <- as.data.frame(em)

library(ggplot2)

ggplot(em_df, aes(x = season, y = response, color = species, group = species)) +
  geom_point(size = 2.5) +
  geom_line(linewidth = 1) +
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL), width = 0.15) +
  scale_color_manual(values = c(
    "Culex pipiens" = "#1bc8ea",
    "Culex tarsalis" = "#FF2DA0"
  )) +
  scale_fill_manual(values = c(
    "Culex pipiens" = "#1bc8ea",
    "Culex tarsalis" = "#FF2DA0"
  )) +
  labs(
    title = "Seasonal abundance",
    x = "Season",
    y = "Predicted abundance",
    color = "Species"
  ) +
    theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5),
    strip.background = element_blank(),
    strip.text = element_text(size = 12)
  )
```
## GLMM with species * season * urbanization, including trap type:
```{r season-urbanization}
library(glmmTMB)
library(dplyr)

fit_combined <- glmmTMB(
  count ~ species * season * urbanization + trap_type +
    (1 | site_name/collection_date),
  family = nbinom1(link = "log"),
  data = combined
)
```
### Plot marginal means by species and season within each urbanization class
```{r marg-with-urbanization, warning=FALSE}
library(emmeans)

em <- emmeans(
  fit_combined,
  ~ species * season | urbanization,
  type = "response"
)

em_df <- as.data.frame(em)

library(ggplot2)

ggplot(em_df, aes(x = season, y = response, color = species, group = species)) +
  geom_point(size = 2.5) +
  geom_line(linewidth = 1) +
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL), width = 0.15) +
    scale_color_manual(values = c(
    "Culex pipiens" = "#1bc8ea",
    "Culex tarsalis" = "#FF2DA0"
  )) +
  scale_fill_manual(values = c(
    "Culex pipiens" = "#1bc8ea",
    "Culex tarsalis" = "#FF2DA0"
  )) +
  
  facet_wrap(~ urbanization) +
  labs(
    title = "Seasonal abundance of Culex pipiens and Culex tarsalis across urbanization classes",
    x = "Season",
    y = "Predicted abundance",
    color = "Species"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5),
    strip.background = element_blank(),
    strip.text = element_text(size = 12)
  )

```

### Plot marginal means by species and season within each trap type
```{r mm-by-trap, warning=FALSE}
em <- emmeans(
  fit_combined,
  ~ species * season | trap_type,
  type = "response"
)

em_df <- as.data.frame(em)

# Handle column name
if ("response" %in% names(em_df)) {
  em_df <- em_df %>% rename(rate = response)
}

# Plot by trap type:
ggplot(em_df, aes(x = season, y = rate, color = species, group = species)) +
  geom_point(size = 2.5) +
  geom_line(linewidth = 1) +
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL), width = 0.15) +
  facet_wrap(~ trap_type) +
  labs(
    title = "Seasonal abundance by species and trap type",
    x = "Season",
    y = "Predicted abundance",
    color = "Species"
  ) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r mm-trap-urbanization}
#facet by both urbanization and trap type

em <- emmeans(
  fit_combined,
  ~ species * season | urbanization * trap_type,
  type = "response"
)

em_df <- as.data.frame(em)

if ("response" %in% names(em_df)) {
  em_df <- em_df %>% rename(rate = response)
}

#plot
ggplot(em_df, aes(x = season, y = rate, color = species, group = species)) +
  geom_point(size = 2.5) +
  geom_line(linewidth = 1) +
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL), width = 0.15) +
  scale_color_manual(values = c(
    "Culex pipiens" = "#1bc8ea",
    "Culex tarsalis" = "#FF2DA0"
  )) +
  scale_fill_manual(values = c(
    "Culex pipiens" = "#1bc8ea",
    "Culex tarsalis" = "#FF2DA0"
  )) +
  facet_grid(urbanization ~ trap_type, scales = "free_y") +
  labs(
    title = "Seasonal abundance by species, urbanization, and trap type",
    x = "Season",
    y = "Predicted abundance",
    color = "Species"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5),
    strip.background = element_blank(),
    strip.text = element_text(size = 12)
  )
```

# GAM-style smooth over weeks
  
```{r bam}
library(dplyr)
library(mgcv)

combined <- combined %>%
  mutate(
    species = factor(species),
    urbanization = factor(urbanization),
    trap_type = factor(trap_type),
    site_name = factor(site_name),
    disease_week = as.numeric(disease_week)
  )
```
In the GLMM, (1 | site_name/collection_date), which handles clustering of repeated observations taken at the same site on the same date with:
- a random intercept for site_name  
- and a random intercept for each site_name:collection_date combination

GAM equivalent, in mgcv, the closest analogue is to create an interaction ID and include it as another random-effect smooth.

First create the grouping variable:

```{r grouping}
combined <- combined %>%
  mutate(
    site_date = interaction(site_name, collection_date, drop = TRUE)
  )
```
Check if the grouping variable occurs often:
```{r group-check}
length(unique(combined$site_date))
nrow(combined)
table(table(combined$site_date))
```
Yes, more than half of the observations are impacted... but many of them are different trap-types. This is a decision to make, to include or not to include? Site_date adds shared noise from shared environment within a sampling event. Since we really care more about ecological patterns over time, and are already including trap-type, is it really needed? Does it change the result?

Let's start with a simple comparison of including ONLY site_name, or including both site_name and site_date:

### GAM with site_name ONLY:
```{r fit-gam-site-name-only}
# Fit GAM model with site_name only
fit_gam_site <- bam(
  count ~ species * urbanization +
    s(disease_week, by = species) +
    trap_type +
    s(site_name, bs = "re"),
  family = nb(),
  data = combined,
  method = "fREML"
)

```
Plot results with site_name ONLY:
```{r plot-gam-site-name-only}
# Create prediction data. 
# Pick one trap type so the comparison is clean
# GRVD because available across urbanization
newdat_site_GRVD <- expand.grid(
  disease_week = seq(min(combined$disease_week), max(combined$disease_week), by = 1),
  species = levels(combined$species),
  urbanization = levels(combined$urbanization),
  trap_type = "GRVD", 
  site_name = levels(combined$site_name)[1]
)

# CO2 for comparison
newdat_site_CO2 <- expand.grid(
  disease_week = seq(min(combined$disease_week), max(combined$disease_week), by = 1),
  species = levels(combined$species),
  urbanization = levels(combined$urbanization),
  trap_type = "CO2", 
  site_name = levels(combined$site_name)[1]
)

# Predict for GRVD on the link scale, then back-transform:
pred_site_GRVD <- predict(fit_gam_site, newdata = newdat_site_GRVD, type = "link", se.fit = TRUE)
newdat_site_GRVD <- newdat_site_GRVD %>%
  mutate(
    fit_link = pred_site_GRVD$fit,
    se_link  = pred_site_GRVD$se.fit,
    fit      = exp(fit_link),
    lower    = exp(fit_link - 1.96 * se_link),
    upper    = exp(fit_link + 1.96 * se_link)
  )

# Predict for CO2 on the link scale, then back-transform:
pred_site_CO2 <- predict(fit_gam_site, newdata = newdat_site_CO2, type = "link", se.fit = TRUE)
newdat_site_CO2 <- newdat_site_CO2 %>%
  mutate(
    fit_link = pred_site_CO2$fit,
    se_link  = pred_site_CO2$se.fit,
    fit      = exp(fit_link),
    lower    = exp(fit_link - 1.96 * se_link),
    upper    = exp(fit_link + 1.96 * se_link)
  )


# Plot for GRVD
ggplot(newdat_site_GRVD, aes(x = disease_week, y = fit, color = species, group = species)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = species), alpha = 0.25, color = NA) +
  
  geom_vline(xintercept = 33, linetype = "dashed", color = "black", linewidth = 0.5) +
 
  annotate("text",
           x = 33,
           y = Inf,
           label = "1st WNV cases",
           angle = 0,
           vjust = 1,
           hjust = -.07,
           size = 3) +
  
  facet_wrap(~ urbanization, scales = "free_y", ncol = 1) +
  
  scale_color_manual(values = c(
    "Culex pipiens" = "#1bc8ea",
    "Culex tarsalis" = "#FF2DA0"
  )) +
  scale_fill_manual(values = c(
    "Culex pipiens" = "#1bc8ea",
    "Culex tarsalis" = "#FF2DA0"
  )) +
  
  labs(
    title = "Predicted abundance (GRVD traps)",
    x = "Disease week",
    y = "Predicted abundance",
    color = "Species",
    fill = "Species"
  ) +
  
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5),
    strip.background = element_blank(),
    strip.text = element_text(size = 12)
  )


# Plot for CO2
ggplot(newdat_site_CO2, aes(x = disease_week, y = fit, color = species, group = species)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = species), alpha = 0.25, color = NA) +
  
  geom_vline(xintercept = 33, linetype = "dashed", color = "black", linewidth = 0.5) +
 
  annotate("text",
           x = 33,
           y = Inf,
           label = "1st WNV cases",
           angle = 0,
           vjust = 1,
           hjust = -.07,
           size = 3) +
  
  facet_wrap(~ urbanization, scales = "free_y", ncol = 1) +
  
  scale_color_manual(values = c(
    "Culex pipiens" = "#1bc8ea",
    "Culex tarsalis" = "#FF2DA0"
  )) +
  scale_fill_manual(values = c(
    "Culex pipiens" = "#1bc8ea",
    "Culex tarsalis" = "#FF2DA0"
  )) +
  
  labs(
    title = "Predicted abundance (CO2 traps)",
    x = "Disease week",
    y = "Predicted abundance",
    color = "Species",
    fill = "Species"
  ) +
  
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5),
    strip.background = element_blank(),
    strip.text = element_text(size = 12)
  )
```


### GAM with site_date as a combined group:
```{r fit-gam-site-date, eval = FALSE}
# Fit GAM model with site_date
fit_gam_site_date <- bam(
  count ~ species * urbanization +
    s(disease_week, by = species) +
    trap_type +
    s(site_date, bs = "re"),
  family = nb(),
  data = combined,
  method = "fREML"
)

```
Plot results with site_date as a combined group:
```{r plot-gam-site-date, eval = FALSE}
# Create prediction data. 
# Pick one trap type so the comparison is clean
# GRVD because available across urbanization
newdat_site_date_GRVD <- expand.grid(
  disease_week = seq(min(combined$disease_week), max(combined$disease_week), by = 1),
  species = levels(combined$species),
  urbanization = levels(combined$urbanization),
  trap_type = "GRVD", 
  site_date = levels(combined$site_date)[1]
)

# CO2 for comparison
newdat_site_date_CO2 <- expand.grid(
  disease_week = seq(min(combined$disease_week), max(combined$disease_week), by = 1),
  species = levels(combined$species),
  urbanization = levels(combined$urbanization),
  trap_type = "CO2", 
  site_date = levels(combined$site_date)[1]
)

# Predict for GRVD on the link scale, then back-transform:
pred_site_date_GRVD <- predict(fit_gam_site_date, newdata = newdat_site_date_GRVD, type = "link", se.fit = TRUE)
newdat_site_date_GRVD <- newdat_site_date_GRVD %>%
  mutate(
    fit_link = pred_site_date_GRVD$fit,
    se_link  = pred_site_date_GRVD$se.fit,
    fit      = exp(fit_link),
    lower    = exp(fit_link - 1.96 * se_link),
    upper    = exp(fit_link + 1.96 * se_link)
  )

# Predict for CO2 on the link scale, then back-transform:
pred_site_date_CO2 <- predict(fit_gam_site_date, newdata = newdat_site_date_CO2, type = "link", se.fit = TRUE)
newdat_site_date_CO2 <- newdat_site_date_CO2 %>%
  mutate(
    fit_link = pred_site_date_CO2$fit,
    se_link  = pred_site_date_CO2$se.fit,
    fit      = exp(fit_link),
    lower    = exp(fit_link - 1.96 * se_link),
    upper    = exp(fit_link + 1.96 * se_link)
  )


# Plot for GRVD
ggplot(newdat_site_date_GRVD, aes(x = disease_week, y = fit, color = species, group = species)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = species), alpha = 0.25, color = NA) +
  
  geom_vline(xintercept = 33, linetype = "dashed", color = "black", linewidth = 0.5) +
 
  annotate("text",
           x = 33,
           y = Inf,
           label = "1st WNV cases",
           angle = 0,
           vjust = 1,
           hjust = -.07,
           size = 3) +
  
  facet_wrap(~ urbanization, scales = "free_y", ncol = 1) +
  
  scale_color_manual(values = c(
    "Culex pipiens" = "#1bc8ea",
    "Culex tarsalis" = "#FF2DA0"
  )) +
  scale_fill_manual(values = c(
    "Culex pipiens" = "#1bc8ea",
    "Culex tarsalis" = "#FF2DA0"
  )) +
  
  labs(
    title = "Predicted abundance (GRVD traps)",
    x = "Disease week",
    y = "Predicted abundance",
    color = "Species",
    fill = "Species"
  ) +
  
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5),
    strip.background = element_blank(),
    strip.text = element_text(size = 12)
  )


# Plot for CO2
ggplot(newdat_site_date_CO2, aes(x = disease_week, y = fit, color = species, group = species)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = species), alpha = 0.25, color = NA) +
  
  geom_vline(xintercept = 33, linetype = "dashed", color = "black", linewidth = 0.5) +
 
  annotate("text",
           x = 33,
           y = Inf,
           label = "1st WNV cases",
           angle = 0,
           vjust = 1,
           hjust = -.07,
           size = 3) +
  
  facet_wrap(~ urbanization, scales = "free_y", ncol = 1) +
  
  scale_color_manual(values = c(
    "Culex pipiens" = "#1bc8ea",
    "Culex tarsalis" = "#FF2DA0"
  )) +
  scale_fill_manual(values = c(
    "Culex pipiens" = "#1bc8ea",
    "Culex tarsalis" = "#FF2DA0"
  )) +
  
  labs(
    title = "Predicted abundance (CO2 traps)",
    x = "Disease week",
    y = "Predicted abundance",
    color = "Species",
    fill = "Species"
  ) +
  
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5),
    strip.background = element_blank(),
    strip.text = element_text(size = 12)
  )
```


