---
title: "Bloodmeal + WNV positive GAM: SLC"
author: "Norah Saarman"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

# Setup
```{r 1-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  # Saves PNGs to the root /figures folder
  fig.path = "../figures/" 
)
```

```{r packages, warning=FALSE}
library(dplyr)
library(brms)
library(mgcv)
library(bayesplot)
library(ggplot2)
library(sf)
library(spdep)
library(tidyr)
library(maptiles)
library(ggspatial)
library(prettymapr)
library(tibble)

```

# Data 

Started with a folder in Emily's home directory called "saarman-group". I cleaned them up and added habitat data:  
 - "bloodmeal_master_FINAL.csv" --> "bloodmeal_Jun2026_NPS.csv"
 - "wnv_master_FINAL.csv" --> "wnv_Jun2026_NPS.csv"
 
 
```{r data}
bloodmeal <- read.csv("/uufs/chpc.utah.edu/common/home/saarman-group1/urbanindex/bloodmeal_Jun2026_NPS.csv")

wnv <- read.csv("/uufs/chpc.utah.edu/common/home/saarman-group1/urbanindex/wnv_Jun2026_NPS.csv")

```
# PCA of habitat/urbanization gradient  


### PCA - bloodmeal GPS points
```{r PCA-bloodmeal}
# Check mosq_species breakdown
table(bloodmeal$mosq_species,bloodmeal$urbanization)

# Pool trap types and check breakdown by trap
bloodmeal$trap_type2 <- bloodmeal$trap_type

# Mutate trap_type to pool walk-in and trash can
bloodmeal <- bloodmeal %>%
  mutate(trap_type2 = case_when(
    trap_type2 == "Walk-in" ~ "GRVD+",
    TRUE                     ~ trap_type2
  ))

bloodmeal <- bloodmeal %>%
  mutate(trap_type2 = case_when(
    trap_type2 == "Trash Can" ~ "GRVD+",
    TRUE                     ~ trap_type2
  ))

bloodmeal <- bloodmeal %>%
  mutate(trap_type2 = case_when(
    trap_type2 == "GRVD" ~ "GRVD+",
    TRUE                     ~ trap_type2
  ))

bloodmeal <- bloodmeal %>%
  mutate(trap_type2 = case_when(
    trap_type2 == "BOX" ~ "GRVD+",
    TRUE                     ~ trap_type2
  ))

table(bloodmeal$mosq_species,bloodmeal$urbanization,bloodmeal$trap_type2)

# Use only sites represented in the bloodmeal dataset to define the PCA
bloodmeal_habitat_sites <- bloodmeal %>%
  filter(site_code != "0") %>%
  distinct(
    site_code,
    impervious_500,
    canopy_500,
    dist_wetland_m,
    summer_ndvi_500
  ) %>%
  filter(complete.cases(
    impervious_500,
    canopy_500,
    dist_wetland_m,
    summer_ndvi_500
  ))

# Run PCA
bloodmeal_pca_vars <- bloodmeal_habitat_sites %>%
  dplyr::select(
    impervious_500,
    canopy_500,
    dist_wetland_m,
    summer_ndvi_500
  )

bloodmeal_pca <- prcomp(
  bloodmeal_pca_vars,
  center = TRUE,
  scale. = TRUE
)

summary(bloodmeal_pca)
round(bloodmeal_pca$rotation, 3)

# Add PCA scores to site-level table
bloodmeal_habitat_sites <- bloodmeal_habitat_sites %>%
  mutate(
    habitat_PC1 = bloodmeal_pca$x[,1],
    habitat_PC2 = bloodmeal_pca$x[,2]
  )

# Join PCA scores back to full WNV table
bloodmeal <- bloodmeal %>%
  dplyr::select(-any_of(c("habitat_PC1", "habitat_PC2"))) %>%
  left_join(
    bloodmeal_habitat_sites %>%
      dplyr::select(site_code, habitat_PC1, habitat_PC2),
    by = "site_code"
  )

urb_colors <- c(
  "Urban" = "#B22222",
  "Peri"  = "#DAA520",
  "Rural" = "#228B22"
)

# Site/trap-level PCA scores from bloodmeal object
pca_scores_df <- bloodmeal %>%
  distinct(
    site_code,
    trap_type2,
    urbanization,
    habitat_PC1,
    habitat_PC2
  ) %>%
  filter(
    !is.na(habitat_PC1),
    !is.na(habitat_PC2),
    !is.na(trap_type2)
  ) %>%
  mutate(
    urbanization = factor(
      urbanization,
      levels = c("Urban", "Peri", "Rural")
    ),
    PC1 = habitat_PC1,
    PC2 = habitat_PC2
  )

# PCA variable loadings
loadings_df <- as.data.frame(bloodmeal_pca$rotation) %>%
  rownames_to_column("variable") %>%
  mutate(
    variable = recode(
      variable,
      "impervious_500"  = "Impervious surface",
      "canopy_500"      = "Tree canopy",
      "dist_wetland_m"  = "Wetland distance",
      "summer_ndvi_500" = "NDVI"
    )
  )

arrow_scale <- 3.5

p1_biplot <- ggplot() +
  stat_ellipse(
    data = pca_scores_df,
    aes(x = PC1, y = PC2, color = urbanization),
    level = 0.75,
    linewidth = 0.8,
    linetype = "dashed"
  ) +
  geom_point(
    data = pca_scores_df,
    aes(
      x = PC1,
      y = PC2,
      color = urbanization,
      fill = urbanization,
      shape = trap_type2
    ),
    size = 3,
    alpha = 0.5,
    stroke = 0.8
  ) +
  geom_segment(
    data = loadings_df,
    aes(
      x = 0,
      y = 0,
      xend = PC1 * arrow_scale,
      yend = PC2 * arrow_scale
    ),
    arrow = arrow(length = unit(0.25, "cm"), type = "closed"),
    color = "gray30",
    linewidth = 0.7
  ) +
  geom_text(
    data = loadings_df,
    aes(
      x = PC1 * arrow_scale * 1.15,
      y = PC2 * arrow_scale * 1.15,
      label = variable
    ),
    size = 3.5,
    color = "gray20",
    fontface = "italic"
  ) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "gray70") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "gray70") +
  scale_color_manual(values = urb_colors, name = "Urbanization") +
  scale_fill_manual(values = urb_colors, name = "Urbanization") +
  scale_shape_manual(
    values = c(
      "CO2" = 20,
      "GRVD+" = 24
    ),
    name = "Trap type"
  ) +
  labs(
    title = "Habitat PCA (Bloodmeal Sites by Trap Type)",
       x = "PC1 - Urbanization (57.6% variance)",
    y = "PC2 - Vegetation (29.7% variance)"
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "gray40"),
    legend.position = "right"
  )

print(p1_biplot)
 
# Plot by mosquito species



# Site/species-level PCA scores from bloodmeal object
pca_scores_df <- bloodmeal %>%
  distinct(
    site_code,
    mosq_species,
    urbanization,
    habitat_PC1,
    habitat_PC2
  ) %>%
  filter(
    !is.na(habitat_PC1),
    !is.na(habitat_PC2),
    !is.na(mosq_species)
  ) %>%
  mutate(
    urbanization = factor(
      urbanization,
      levels = c("Urban", "Peri", "Rural")
    ),
    mosq_species = recode(
      mosq_species,
      "Cx_pipiens_sl" = "Cx. pipiens s.l.",
      "Cx_tarsalis" = "Cx. tarsalis"
    ),
    PC1 = habitat_PC1,
    PC2 = habitat_PC2
  )

p1_biplot_spp <- ggplot() +
  stat_ellipse(
    data = pca_scores_df,
    aes(x = PC1, y = PC2, color = urbanization),
    level = 0.75,
    linewidth = 0.8,
    linetype = "dashed"
  ) +
  geom_point(
    data = pca_scores_df,
    aes(
      x = PC1,
      y = PC2,
      color = urbanization,
      fill = urbanization,
      shape = mosq_species
    ),
    size = 3,
    alpha = 0.5,
    stroke = 0.8
  ) +
  geom_segment(
    data = loadings_df,
    aes(
      x = 0,
      y = 0,
      xend = PC1 * arrow_scale,
      yend = PC2 * arrow_scale
    ),
    arrow = arrow(length = unit(0.25, "cm"), type = "closed"),
    color = "gray30",
    linewidth = 0.7
  ) +
  geom_text(
    data = loadings_df,
    aes(
      x = PC1 * arrow_scale * 1.15,
      y = PC2 * arrow_scale * 1.15,
      label = variable
    ),
    size = 3.5,
    color = "gray20",
    fontface = "italic"
  ) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "gray70") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "gray70") +
  scale_color_manual(values = urb_colors, name = "Urbanization") +
  scale_fill_manual(values = urb_colors, name = "Urbanization") +
  scale_shape_manual(
    values = c(
      "Cx. pipiens s.l." = 10,
      "Cx. tarsalis" = 24
    ),
    name = "Species"
  ) +
  labs(
    title = "Habitat PCA (Bloodmeal Sites by Mosquito)",
    x = "PC1 - Urbanization (57.6% variance)",
    y = "PC2 - Vegetation (29.7% variance)"
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "gray40"),
    legend.position = "right"
  )

print(p1_biplot_spp)


```  


### PCA - WNV dataset

```{r PCA-wnv}

# Check mosq_species breakdown
table(wnv$mosq_species,wnv$urbanization, wnv$trap_type)

# Pool trap types and check breakdown by trap
wnv$trap_type2 <- wnv$trap_type

# Mutate trap_type to pool walk-in and trash can
wnv <- wnv %>%
  mutate(trap_type2 = case_when(
    trap_type2 == "BGSENT" ~ "CO2+",
    TRUE                     ~ trap_type2
  ))

wnv <- wnv %>%
  mutate(trap_type2 = case_when(
    trap_type2 == "CO2" ~ "CO2+",
    TRUE                     ~ trap_type2
  ))

wnv <- wnv %>%
  mutate(trap_type2 = case_when(
    trap_type2 == "GRVD" ~ "GRVD+",
    TRUE                     ~ trap_type2
  ))

wnv <- wnv %>%
  mutate(trap_type2 = case_when(
    trap_type2 == "OTHER" ~ "GRVD+",
    TRUE                     ~ trap_type2
  ))

wnv <- wnv %>%
  mutate(trap_type2 = case_when(
    trap_type2 == "REST" ~ "GRVD+",
    TRUE                     ~ trap_type2
  ))

table(wnv$mosq_species,wnv$urbanization,wnv$trap_type2)

# Use only sites represented in the WNV dataset to define the PCA
wnv_habitat_sites <- wnv %>%
  filter(site_code != "0") %>%
  distinct(
    site_code,
    impervious_500,
    canopy_500,
    dist_wetland_m,
    summer_ndvi_500
  ) %>%
  filter(complete.cases(
    impervious_500,
    canopy_500,
    dist_wetland_m,
    summer_ndvi_500
  ))

# Run PCA
wnv_pca_vars <- wnv_habitat_sites %>%
  dplyr::select(
    impervious_500,
    canopy_500,
    dist_wetland_m,
    summer_ndvi_500
  )

wnv_pca <- prcomp(
  wnv_pca_vars,
  center = TRUE,
  scale. = TRUE
)

summary(wnv_pca)
round(wnv_pca$rotation, 3)

# Add PCA scores to site-level table
wnv_habitat_sites <- wnv_habitat_sites %>%
  mutate(
    habitat_PC1 = wnv_pca$x[,1],
    habitat_PC2 = wnv_pca$x[,2]
  )

# Join PCA scores back to full WNV table
wnv <- wnv %>%
  dplyr::select(-any_of(c("habitat_PC1", "habitat_PC2"))) %>%
  left_join(
    wnv_habitat_sites %>%
      dplyr::select(site_code, habitat_PC1, habitat_PC2),
    by = "site_code"
  )

urb_colors <- c(
  "Urban" = "#B22222",
  "Peri"  = "#DAA520",
  "Rural" = "#228B22"
)

# Site/trap-level PCA scores from WNV object
pca_scores_df <- wnv %>%
  distinct(
    site_code,
    trap_type2,
    urbanization,
    habitat_PC1,
    habitat_PC2
  ) %>%
  filter(
    !is.na(habitat_PC1),
    !is.na(habitat_PC2),
    !is.na(trap_type2)
  ) %>%
  mutate(
    urbanization = factor(
      urbanization,
      levels = c("Urban", "Peri", "Rural")
    ),
    PC1 = habitat_PC1,
    PC2 = habitat_PC2
  )

# PCA variable loadings
loadings_df <- as.data.frame(wnv_pca$rotation) %>%
  rownames_to_column("variable") %>%
  mutate(
    variable = recode(
      variable,
      "impervious_500"  = "Impervious surface",
      "canopy_500"      = "Tree canopy",
      "dist_wetland_m"  = "Wetland distance",
      "summer_ndvi_500" = "NDVI"
    )
  )

arrow_scale <- 3.5

p2_biplot <- ggplot() +
  stat_ellipse(
    data = pca_scores_df,
    aes(x = PC1, y = PC2, color = urbanization),
    level = 0.75,
    linewidth = 0.8,
    linetype = "dashed"
  ) +
  geom_point(
    data = pca_scores_df,
    aes(
      x = PC1,
      y = PC2,
      color = urbanization,
      fill = urbanization,
      shape = trap_type2
    ),
    size = 3,
    alpha = 0.5,
    stroke = 0.8
  ) +
  geom_segment(
    data = loadings_df,
    aes(
      x = 0,
      y = 0,
      xend = PC1 * arrow_scale,
      yend = PC2 * arrow_scale
    ),
    arrow = arrow(length = unit(0.25, "cm"), type = "closed"),
    color = "gray30",
    linewidth = 0.7
  ) +
  geom_text(
    data = loadings_df,
    aes(
      x = PC1 * arrow_scale * 1.15,
      y = PC2 * arrow_scale * 1.15,
      label = variable
    ),
    size = 3.5,
    color = "gray20",
    fontface = "italic"
  ) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "gray70") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "gray70") +
  scale_color_manual(values = urb_colors, name = "Urbanization") +
  scale_fill_manual(values = urb_colors, name = "Urbanization") +
  scale_shape_manual(
    values = c(
      "CO2+" = 20,
      "GRVD+" = 24
    ),
    name = "Trap type"
  ) +
  labs(
    title = "Habitat PCA (WNV Sites)",
       x = "PC1 - Urbanization (57.6% variance)",
    y = "PC2 - Vegetation (29.7% variance)"
  ) +
  scale_x_reverse() +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "gray40"),
    legend.position = "right"
  )

print(p2_biplot)

```  
  
## PCA - pip only WNV
```{r pip-only-PCA}

# Make sure all random effect variables are factors
wnv <- wnv %>%
mutate(
site_code = factor(site_code),
trap_type2 = factor(trap_type2),
year_f = factor(year)
)

# Make pipiens-only dataset first
pipiens_data <- wnv %>%
  filter(mosq_species == "Cx_pipiens_sl")

# Use only pipiens sites to define PCA
pip_habitat_sites <- pipiens_data %>%
  filter(site_code != "0") %>%
  distinct(
    site_code,
    impervious_500,
    canopy_500,
    dist_wetland_m,
    summer_ndvi_500
  ) %>%
  filter(complete.cases(
    impervious_500,
    canopy_500,
    dist_wetland_m,
    summer_ndvi_500
  ))

# Run PCA
pip_pca_vars <- pip_habitat_sites %>%
  dplyr::select(
    impervious_500,
    canopy_500,
    dist_wetland_m,
    summer_ndvi_500
  )

pip_pca <- prcomp(
  pip_pca_vars,
  center = TRUE,
  scale. = TRUE
)

summary(pip_pca)
round(pip_pca$rotation, 3)

# Add PCA scores to pipiens site-level table
pip_habitat_sites <- pip_habitat_sites %>%
  mutate(
    habitat_PC1 = pip_pca$x[, 1],
    habitat_PC2 = pip_pca$x[, 2]
  )

# Join PCA scores back to pipiens-only data
pipiens_data <- pipiens_data %>%
  dplyr::select(-any_of(c("habitat_PC1", "habitat_PC2"))) %>%
  left_join(
    pip_habitat_sites %>%
      dplyr::select(site_code, habitat_PC1, habitat_PC2),
    by = "site_code"
  )

# Site/trap-level PCA scores from pipiens object
pip_pca_scores_df <- pipiens_data %>%
  distinct(
    site_code,
    trap_type2,
    urbanization,
    habitat_PC1,
    habitat_PC2
  ) %>%
  filter(
    !is.na(habitat_PC1),
    !is.na(habitat_PC2),
    !is.na(trap_type2)
  ) %>%
  mutate(
    urbanization = factor(
      urbanization,
      levels = c("Urban", "Peri", "Rural")
    ),
    PC1 = habitat_PC1,
    PC2 = habitat_PC2
  )

# PCA variable loadings
pip_loadings_df <- as.data.frame(pip_pca$rotation) %>%
  rownames_to_column("variable") %>%
  mutate(
    variable = recode(
      variable,
      "impervious_500"  = "Impervious surface",
      "canopy_500"      = "Tree canopy",
      "dist_wetland_m"  = "Wetland distance",
      "summer_ndvi_500" = "NDVI"
    )
  )

arrow_scale <- 3.5

pip_biplot <- ggplot() +
  stat_ellipse(
    data = pip_pca_scores_df,
    aes(x = PC1, y = PC2, color = urbanization),
    level = 0.75,
    linewidth = 0.8,
    linetype = "dashed"
  ) +
  geom_point(
    data = pip_pca_scores_df,
    aes(
      x = PC1,
      y = PC2,
      color = urbanization,
      fill = urbanization,
      shape = trap_type2
    ),
    size = 3,
    alpha = 0.5,
    stroke = 0.8
  ) +
  geom_segment(
    data = pip_loadings_df,
    aes(
      x = 0,
      y = 0,
      xend = PC1 * arrow_scale,
      yend = PC2 * arrow_scale
    ),
    arrow = arrow(length = unit(0.25, "cm"), type = "closed"),
    color = "gray30",
    linewidth = 0.7
  ) +
  geom_text(
    data = pip_loadings_df,
    aes(
      x = PC1 * arrow_scale * 1.15,
      y = PC2 * arrow_scale * 1.15,
      label = variable
    ),
    size = 3.5,
    color = "gray20",
    fontface = "italic"
  ) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "gray70") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "gray70") +
  scale_color_manual(values = urb_colors, name = "Urbanization") +
  scale_fill_manual(values = urb_colors, name = "Urbanization") +
  scale_shape_manual(
    values = c(
      "CO2+" = 20,
      "GRVD+" = 24
    ),
    name = "Trap type"
  ) +
  labs(
    title = "Habitat PCA: Cx. pipiens s.l. WNV Sites",
    x = "PC1 - Urbanization",
    y = "PC2 - Vegetation"
  ) +
  scale_x_reverse() +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "gray40"),
    legend.position = "right"
  )

print(pip_biplot)

# Make sure random effect variables are factors
pipiens_data <- pipiens_data %>%
  mutate(
    site_code = factor(site_code),
    trap_type2 = factor(trap_type2),
    year_f = factor(year)
  )
```

## PCA - pip only WNV paired-traps

```{r paired-pip}
# Identify truly paired pipiens sites
paired_pip_sites <- pipiens_data %>%
  filter(!is.na(trap_type2)) %>%
  distinct(site_code, trap_type2) %>%
  group_by(site_code) %>%
  summarize(
    trap_types_present = paste(sort(unique(trap_type2)), collapse = ", "),
    n_trap_types = n_distinct(trap_type2),
    has_CO2 = any(trap_type2 == "CO2+"),
    has_GRVD = any(trap_type2 == "GRVD+"),
    .groups = "drop"
  ) %>%
  filter(has_CO2, has_GRVD)

paired_pip_sites

pip_pca_scores_paired_df <- pip_pca_scores_df %>%
  semi_join(
    paired_pip_sites,
    by = "site_code"
  )

pip_biplot_paired <- ggplot() +
  stat_ellipse(
    data = pip_pca_scores_paired_df,
    aes(x = PC1, y = PC2, color = urbanization),
    level = 0.75,
    linewidth = 0.8,
    linetype = "dashed"
  ) +
  geom_point(
    data = pip_pca_scores_paired_df,
    aes(
      x = PC1,
      y = PC2,
      color = urbanization,
      fill = urbanization,
      shape = trap_type2
    ),
    size = 3,
    alpha = 0.6,
    stroke = 0.8
  ) +
  geom_segment(
    data = pip_loadings_df,
    aes(
      x = 0,
      y = 0,
      xend = PC1 * arrow_scale,
      yend = PC2 * arrow_scale
    ),
    arrow = arrow(length = unit(0.25, "cm"), type = "closed"),
    color = "gray30",
    linewidth = 0.7
  ) +
  geom_text(
    data = pip_loadings_df,
    aes(
      x = PC1 * arrow_scale * 1.15,
      y = PC2 * arrow_scale * 1.15,
      label = variable
    ),
    size = 3.5,
    color = "gray20",
    fontface = "italic"
  ) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "gray70") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "gray70") +
  scale_color_manual(values = urb_colors, name = "Urbanization") +
  scale_fill_manual(values = urb_colors, name = "Urbanization") +
  scale_shape_manual(
    values = c(
      "CO2+" = 20,
      "GRVD+" = 24
    ),
    name = "Trap type"
  ) +
  labs(
    title = "Habitat PCA: Paired Cx. pipiens s.l. WNV Sites",
    x = "PC1 - Urbanization",
    y = "PC2 - Vegetation"
  ) +
  scale_x_reverse() +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "right"
  )

print(pip_biplot_paired)
```
# RQ1: Bloodmeal  
```{r model1}
# MODEL 1: Bloodmeal host ~ mosquito species + urbanization
# Make sure it's a factor
bloodmeal <- bloodmeal %>%
mutate(host_3_categories = factor(host_3_categories))
# Check levels — brm will use the fir
levels(bloodmeal$host_3_categories)

# make "Other bird" reference category
bloodmeal <- bloodmeal %>%
mutate(host_3_categories = relevel(host_3_categories,
ref = "Other bird"))
levels(bloodmeal$host_3_categories)

# fit model
model_1 <- brm(
host_3_categories ~ mosq_species + habitat_PC1 + (1|site_code) + (1|year),
family = categorical(),
data = bloodmeal,
cores = 4,
chains = 4,
iter = 4000,
warmup = 2000,
control = list(adapt_delta = 0.99),
file = "../results/bloodmeal_host3_habitatPC1",
seed = 908275
)

summary(model_1)
pp_check(model_1, ndraws = 100) 
mcmc_trace(model_1, pars = vars(contains("habitat_PC1")))
mcmc_trace(model_1, pars = vars(contains("mosq_species")))
```
  
# RQ2: WNV probability

## Plot raw by urbanization:
```{r}
wnv_site <- wnv %>%
  group_by(site_code, mosq_species) %>%
  summarise(
    pct_infected = mean(infected, na.rm = TRUE) * 100,
    n_pools = n(),
    habitat_PC1 = first(habitat_PC1),
    .groups = "drop"
  ) %>%
  filter(!is.na(habitat_PC1)) %>%
  mutate(
    mosq_species = recode(
      mosq_species,
      "Cx_pipiens_sl" = "Cx. pipiens s.l.",
      "Cx_tarsalis" = "Cx. tarsalis"
    )
  )

ggplot(wnv_site, aes(x = habitat_PC1, y = pct_infected)) +
  geom_point(aes(size = n_pools), alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "firebrick", linewidth = 1) +
  facet_wrap(~ mosq_species, scales = "free_y") +
  scale_x_reverse() +
  scale_size_continuous(name = "Pools tested", range = c(2, 8)) +
  labs(
    title = "Urbanization Gradient vs WNV Infection Rate",
    x = "PC1 — Urbanization gradient",
    y = "% infected pools per site"
  ) +
  theme_classic(base_size = 13)

cols <- c(
  "Cx. pipiens s.l." = "#1bc8ea",
  "Cx. tarsalis" = "#FF2DA0"
)

wnv_site <- wnv %>%
  group_by(site_code, mosq_species) %>%
  summarise(
    pct_infected = mean(infected, na.rm = TRUE) * 100,
    n_pools = n(),
    habitat_PC1 = first(habitat_PC1),
    .groups = "drop"
  ) %>%
  filter(!is.na(habitat_PC1)) %>%
  mutate(
    mosq_species = recode(
      mosq_species,
      "Cx_pipiens_sl" = "Cx. pipiens s.l.",
      "Cx_tarsalis" = "Cx. tarsalis"
    )
  )

ggplot(
  wnv_site,
  aes(
    x = habitat_PC1,
    y = pct_infected,
    color = mosq_species
  )
) +
  geom_point(
    aes(size = n_pools),
    alpha = 0.7
  ) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    linewidth = 1
  ) +
  scale_color_manual(values = cols, name = "Species") +
  scale_x_reverse() +
  scale_size_continuous(
    name = "Pools tested",
    range = c(2, 8)
  ) +
  labs(
    title = NULL,
    x = "Urbanization PC1 (Rural --> Urban)",
    y = "% infected pools per site"
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5)
  )
```

  
## Pipiens

### Pip ~ habitat_PC1
```{r pip}
# Pip MODEL 2v1: WNV infection ~ habitat_PC1 + GAM smooths

# Make sure all random effect variables are factors
wnv <- wnv %>%
mutate(
site_code = factor(site_code),
trap_type2 = factor(trap_type2),
year_f = factor(year)
)

pipiens_data <- wnv %>% filter(mosq_species == "Cx_pipiens_sl")

gam_pipiens <- gam(
infected ~ habitat_PC1 +
  s(disease_week, bs = "tp", k = 8) +
s(disease_week, by = year_f, bs = "fs", k = 15) +
s(year_f, bs = "re") +
s(site_code, bs = "re") +
s(trap_type2, bs = "re") + #trap_type2 is random effect
offset(log(num_count)),
family = binomial(link = "cloglog"),
data = pipiens_data,
method = "REML"
)
summary(gam_pipiens)
```  

### Pip ~ s(habitat_PC1)

```{r pip-s} 

# Pip MODEL 2v1: WNV infection ~ s(habitat_PC1) + GAM smooths

#allow smooth to vary across urbanization gradient (non-linear)
gam_pipiens2 <- gam(
  infected ~
    s(habitat_PC1, k = 4) +
    s(disease_week, bs = "tp", k = 8) +
    s(disease_week, by = year_f, bs = "fs", k = 15) +
    s(year_f, bs = "re") +
    s(site_code, bs = "re") +
    s(trap_type2, bs = "re") +   #trap_type2 is random effect
    offset(log(num_count)),
  family = binomial(link = "cloglog"),
  data = pipiens_data,
  method = "REML"
)
summary(gam_pipiens2)

AIC(gam_pipiens, gam_pipiens2)

plot(
  gam_pipiens2,
  select = 1,
  shade = TRUE,
  residuals = TRUE,
  rug = TRUE
)

# Pipiens diagnostics
par(mfrow = c(2, 2))
gam.check(gam_pipiens)
par(mfrow = c(1, 1))
plot(gam_pipiens, pages = 1, residuals = TRUE, shade = TRUE, seWithMean = TRUE)

# Concurvity
concurvity(gam_pipiens, full = FALSE)

# Pipiens diagnostics
par(mfrow = c(2, 2))
gam.check(gam_pipiens2)
par(mfrow = c(1, 1))
plot(gam_pipiens2, pages = 1, residuals = TRUE, shade = TRUE, seWithMean = TRUE)

# Concurvity
concurvity(gam_pipiens2, full = FALSE)

```
Interpretation:

Formula:
infected ~ habitat_PC1 + s(disease_week, bs = "tp", k = 8) + 
    s(disease_week, by = year_f, bs = "fs", k = 15) + s(year_f, 
    bs = "re") + s(site_code, bs = "re") + s(trap_type2, bs = "re") + 
    offset(log(num_count))    
            Estimate Std. Error z value Pr(>|z|)      
habitat_PC1 -0.26895    0.08602  -3.127  0.00177 **   
s(disease_week) 4.36446  5.106 54.269  < 2e-16 ***  
 - Smooth term for disease week is significant.  
 - habitat_PC1 is statistically significant. 
 
Formula:
infected ~ s(habitat_PC1, k = 4) + s(disease_week, bs = "tp", 
    k = 8) + s(disease_week, by = year_f, bs = "fs", k = 15) + 
    s(year_f, bs = "re") + s(site_code, bs = "re") + s(trap_type2, 
    bs = "re") + offset(log(num_count))  
                      edf Ref.df  Chi.sq  p-value    
s(disease_week) 4.3730549  5.113  54.954  < 2e-16 ***  
s(habitat_PC1)  1.8748501  2.259  11.168  0.00643 **   
 - Smooth term for disease week is significant.  
 - The smooth term for habitat_PC1 is statistically significant.  
      - EDF ≈ 1 would be essentially linear.  
      - EDF ≈ 1.8 suggests modest nonlinearity.  
 - The AIC improvement with including a smooth term for habitat_PC1 is tiny (less than 1).  

I think this means there is evidence that habitat_PC1 matters for pipiens, and there may be some curvature, but the curvature is fairly weak. In other words, probability of infection in pipiens varies significantly across the urbanization gradient, even when we force a linear effect... but especially when we allow for some curvature in the relationship. 

### Pip ~ habitat_PC1 + trap_type2
```{r pip-trap}
# Pip MODEL 2v2: WNV infection ~ habitat_PC1 + trap_type2 + GAM smooths

gam_pipiens_trapfixed <- gam(
infected ~ habitat_PC1 + trap_type2 +
  s(disease_week, bs = "tp", k = 8) +
  s(disease_week, by = year_f, bs = "fs", k = 15) +           # bs = "fs" is likely ignored here, instead:
    # s(disease_week, year_f, bs = "fs", k = 15) +           # Option 1
    # s(disease_week, by = year_f, k = 15) +                 # Option 2
  s(year_f, bs = "re") +
  s(site_code, bs = "re") +
  offset(log(num_count)),
  family = binomial(link = "cloglog"),
  data = pipiens_data,
  method = "REML"
)

summary(gam_pipiens_trapfixed)

```

### Pip ~ s(habitat_PC1) + trap_type2

Allow smooth to vary across urbanization gradient (non-linear), and include trap_type2 as a fixed effect.

```{r pip-s-trap}
#
gam_pipiens2_trapfixed <- gam(
  infected ~ trap_type2 +
    s(habitat_PC1, k = 4) +
    s(disease_week, bs = "tp", k = 8) +
    s(disease_week, by = year_f, bs = "fs", k = 15) +
    s(year_f, bs = "re") +
    s(site_code, bs = "re") +
    offset(log(num_count)),
  family = binomial(link = "cloglog"),
  data = pipiens_data,
  method = "REML"
)
summary(gam_pipiens2_trapfixed)

AIC(gam_pipiens, gam_pipiens2, gam_pipiens_trapfixed, gam_pipiens2_trapfixed)
```
### Pip ~ s(habitat_PC1) + s(trap_type2)

Split data set into 2 trap types and allow smooth to vary across trap types... check that the curves look the same, check that the effect sizes match what we found before. 

```{r pip-by-trap}
# CO2+ only, allow smooth to vary across urbanization gradient (non-linear)

pipiens_data_CO2 <- pipiens_data %>%
  filter(trap_type2 == "CO2+")

gam_pipiens2_trapfixed_CO2 <- gam(
  infected ~
    s(habitat_PC1, k = 4) +
    s(disease_week, bs = "tp", k = 8) +
    s(disease_week, by = year_f, bs = "fs", k = 15) +
    s(year_f, bs = "re") +
    s(site_code, bs = "re") +
    offset(log(num_count)),
  family = binomial(link = "cloglog"),
  data = pipiens_data_CO2,
  method = "REML"
)
summary(gam_pipiens2_trapfixed_CO2)

# GRVD+ only, allow smooth to vary across urbanization gradient (non-linear)
gam_pipiens2_trapfixed_GRVD <- gam(
  infected ~ s(habitat_PC1, k = 4) +
    s(disease_week, bs = "tp", k = 8) +
    s(disease_week, by = year_f, bs = "fs", k = 15) +
    s(year_f, bs = "re") +
    s(site_code, bs = "re") +
    offset(log(num_count)),
  family = binomial(link = "cloglog"),
  data = pipiens_data[pipiens_data$trap_type2 == "GRVD+",],
  method = "REML"
)
summary(gam_pipiens2_trapfixed_GRVD)


# Predicted WNV positivity through time from separate CO2+ and GRVD+ models

pc1_values <- tibble::tribble(
  ~habitat_level, ~habitat_PC1,
  "Rural", quantile(wnv$habitat_PC1, 0.9, na.rm = TRUE),
  "Peri",  median(wnv$habitat_PC1, na.rm = TRUE),
  "Urban", quantile(wnv$habitat_PC1, 0.1, na.rm = TRUE)
)

week_seq <- seq(
  min(wnv$disease_week, na.rm = TRUE),
  max(wnv$disease_week, na.rm = TRUE),
  by = 1
)

pred_co2 <- expand.grid(
  disease_week = week_seq,
  year_f = "2021",
  site_code = levels(pipiens_data$site_code)[1],
  num_count = median(pipiens_data$num_count, na.rm = TRUE)
) %>%
  left_join(pc1_values, by = character())

pred_grvd <- pred_co2

co2_pred <- predict(
  gam_pipiens2_trapfixed_CO2,
  newdata = pred_co2,
  type = "link",
  se.fit = TRUE,
  exclude = c("s(site_code)", "s(year_f)")
)

grvd_pred <- predict(
  gam_pipiens2_trapfixed_GRVD,
  newdata = pred_grvd,
  type = "link",
  se.fit = TRUE,
  exclude = c("s(site_code)", "s(year_f)")
)

pred_co2 <- pred_co2 %>%
  mutate(
    fit = 1 - exp(-exp(co2_pred$fit)),
    lower = 1 - exp(-exp(co2_pred$fit - 1.96 * co2_pred$se.fit)),
    upper = 1 - exp(-exp(co2_pred$fit + 1.96 * co2_pred$se.fit)),
    trap_type2 = "CO2+"
  )

pred_grvd <- pred_grvd %>%
  mutate(
    fit = 1 - exp(-exp(grvd_pred$fit)),
    lower = 1 - exp(-exp(grvd_pred$fit - 1.96 * grvd_pred$se.fit)),
    upper = 1 - exp(-exp(grvd_pred$fit + 1.96 * grvd_pred$se.fit)),
    trap_type2 = "GRVD+"
  )

pred_pip_week <- bind_rows(pred_co2, pred_grvd) %>%
  mutate(
    habitat_level = factor(habitat_level, levels = c("Rural", "Peri", "Urban")),
    trap_type2 = factor(trap_type2, levels = c("CO2+", "GRVD+"))
  )

urb_colors <- c(
  "Urban" = "#B22222",
  "Peri"  = "#DAA520",
  "Rural" = "#228B22"
)

p_co2 <- pred_pip_week %>%
  filter(trap_type2 == "CO2+") %>%
  ggplot(
    aes(
      x = disease_week,
      y = fit * 100,
      color = habitat_level,
      fill = habitat_level
    )
  ) +
  geom_ribbon(
    aes(
      ymin = lower * 100,
      ymax = upper * 100
    ),
    alpha = 0.15,
    color = NA
  ) +
  geom_line(linewidth = 1.2) +
  facet_wrap(~ habitat_level, ncol = 1, scales = "free_y") +
  scale_color_manual(values = urb_colors, name = "Habitat") +
  scale_fill_manual(values = urb_colors, guide = "none") +
  labs(
    title = "Cx. pipiens CO2+ traps",
    x = "Disease week",
    y = "Predicted WNV positivity (%)"
  ) +
  theme_classic(base_size = 10) +
  theme(
    strip.background = element_blank(),
    legend.position = "none",
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

p_grvd <- pred_pip_week %>%
  filter(trap_type2 == "GRVD+") %>%
  ggplot(
    aes(
      x = disease_week,
      y = fit * 100,
      color = habitat_level,
      fill = habitat_level
    )
  ) +
  geom_ribbon(
    aes(
      ymin = lower * 100,
      ymax = upper * 100
    ),
    alpha = 0.15,
    color = NA
  ) +
  geom_line(linewidth = 1.2) +
  facet_wrap(~ habitat_level, ncol = 1, scales = "free_y") +
  scale_color_manual(values = urb_colors, name = "Habitat") +
  scale_fill_manual(values = urb_colors, guide = "none") +
  labs(
    title = "Cx. pipiens GRVD+ traps",
    x = "Disease week",
    y = "Predicted WNV positivity (%)"
  ) +
  theme_classic(base_size = 10) +
  theme(
    strip.background = element_blank(),
    legend.position = "none",
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

p_co2
p_grvd
```

Interpretation:  

Trap_type2 is significant as well. AIC is very similar with/without smooth term, and with trap_type as a fixed_effect or a random effect. When we split by trap type, CO2 remains signficant for habitat type, GRVD does not. So, I think its safe to proceed with CO2 traps only. This is the same trap type we used for Katie's analysis as well.  

Also, interestingly, in the final models for CO2 only, even though s(year_f)  p < 2e-16, year-specific seasonal smooths were not significant:
s(disease_week):year_f2018  p = 0.50
s(disease_week):year_f2019  p = 0.45
s(disease_week):year_f2020  p = 0.39
s(disease_week):year_f2021  p = 0.63
s(disease_week):year_f2022  p = 0.47
s(disease_week):year_f2023  p = 0.53

Maybe patterns are broadly similar among years and differ mainly in overall magnitude... so I'll try running it with year as a random effect.
  
### Pip try (fail) simplified s(year_f, bs = "re")

Trying to include year as random effect instead of year-specific smooths.

```{r pip-yr-random}

gam_pipiens_co2_simple <- gam(
  infected ~
    s(habitat_PC1, k = 4) +
    s(disease_week, bs = "tp", k = 8) +
    s(year_f, bs = "re") +
    s(site_code, bs = "re") +
    offset(log(num_count)),
  family = binomial(link = "cloglog"),
  data = pipiens_data %>% filter(trap_type2 == "CO2+"),
  method = "REML"
)

summary(gam_pipiens_co2_simple)

AIC(gam_pipiens2_trapfixed_CO2, gam_pipiens_co2_simple)
```

Models including year-specific seasonal variation were better supported than models with a single seasonal smooth (ΔAIC = 7.6), so year-specific smooths are better.

### Pip interpretation  

For Cx. pipiens s.l., both urbanization and trap type influenced predicted WNV positivity. However, trap type substantially affected infection estimates, indicating that comparisons across the urbanization gradient should be made within a single trapping method. When analyses were restricted to CO2 collections, WNV positivity was highest at urban sites and declined toward more rural habitats. In contrast, the urbanization effect was weak and not supported in GRVD collections. Because CO2 traps provided broad coverage across the urbanization gradient and yielded a consistent urbanization signal, subsequent interpretation focused on CO2 collections.  

### Pip Moran's I for Cx. pipiens residuals
```{r moran-pip}
# for CO2 model
pipiens_data_CO2 <- pipiens_data[pipiens_data$trap_type2 == "CO2+",]

pipiens_resids <- pipiens_data_CO2 %>%
  mutate(resid = residuals(gam_pipiens2_trapfixed_CO2, type = "deviance")) %>%
  group_by(site_code) %>%
  summarise(
    mean_resid = mean(resid, na.rm = TRUE),
    latitude = first(latitude),
    longitude = first(longitude),
    .groups = "drop"
  ) %>%
  filter(!is.na(mean_resid), !is.na(latitude), !is.na(longitude))

coords_pipiens <- cbind(pipiens_resids$longitude, pipiens_resids$latitude)

lw5_pipiens <- coords_pipiens %>%
  knearneigh(k = 5) %>%
  knn2nb() %>%
  nb2listw(style = "W")

moran.test(pipiens_resids$mean_resid, lw5_pipiens)

```
Moran's I shows no spatial autocorrelation in residuals for pipiens.

## Tarsalis   
### Tar ~ habitat_PC1
```{r tar}
tarsalis_data <- wnv %>% filter(mosq_species == "Cx_tarsalis")

gam_tarsalis <- gam(
infected ~ habitat_PC1 +
  s(disease_week, bs = "tp", k = 8) +
s(disease_week, by = year_f, bs = "fs", k = 15) +
s(year_f, bs = "re") +
s(site_code, bs = "re") +
s(trap_type2, bs = "re") +
offset(log(num_count)),
family = binomial(link = "cloglog"),
data = tarsalis_data,
method = "REML"
)
summary(gam_tarsalis)
AIC(gam_tarsalis)
```

### Tar ~ s(habitat_PC1)

```{r tar-s} 
#allow smooth to vary across urbanization gradient (non-linear)
gam_tarsalis2 <- gam(
  infected ~
    s(habitat_PC1, k = 4) +
    s(disease_week, bs = "tp", k = 8) +
    s(disease_week, by = year_f, bs = "fs", k = 15) +
    s(year_f, bs = "re") +
    s(site_code, bs = "re") +
    s(trap_type2, bs = "re") +   #trap_type2 is random effect
    offset(log(num_count)),
  family = binomial(link = "cloglog"),
  data = tarsalis_data,
  method = "REML"
)
summary(gam_tarsalis2)

AIC(gam_tarsalis, gam_tarsalis2)

plot(
  gam_tarsalis2,
  select = 1,
  shade = TRUE,
  residuals = TRUE,
  rug = TRUE
)


# Tarsalis diagnostics
par(mfrow = c(2, 2))
gam.check(gam_tarsalis)
par(mfrow = c(1, 1))
plot(gam_tarsalis, pages = 1, residuals = TRUE, shade = TRUE, seWithMean = TRUE)
concurvity(gam_tarsalis, full = FALSE)

par(mfrow = c(2, 2))
gam.check(gam_tarsalis2)
par(mfrow = c(1, 1))
plot(gam_tarsalis2, pages = 1, residuals = TRUE, shade = TRUE, seWithMean = TRUE)
concurvity(gam_tarsalis2, full = FALSE)



```

Interpretation:
habitat_PC1 and the smooth term for habitat_PC1 are NOT statistically significant.
The AIC improvement with including a smooth term for habitat_PC1 is tiny (less than 1).

I think this means there is no evidence that habitat_PC1 matters for tarsalis. In other words, probability of infection in tarsalis does not vary across the urbanization gradient.
  
### Tar ~ habitat_PC1 + trap_type2
```{r tar-trap}
# Tar MODEL 2v2: WNV infection ~ habitat_PC1 + trap_type2 + GAM smooths

gam_tarsalis_trapfixed <- gam(
infected ~ habitat_PC1 + trap_type2 +
  s(disease_week, bs = "tp", k = 8) +
s(disease_week, by = year_f, bs = "fs", k = 15) +
s(year_f, bs = "re") +
s(site_code, bs = "re") +
offset(log(num_count)),
family = binomial(link = "cloglog"),
data = tarsalis_data,
method = "REML"
)
summary(gam_tarsalis_trapfixed)
```

### Tar ~ s(habitat_PC1) + trap_type2

Allow smooth to vary across urbanization gradient (non-linear), and include trap_type2 as a fixed effect.

```{r tar-s-trap}
#allow smooth to vary across urbanization gradient (non-linear)
gam_tarsalis2_trapfixed <- gam(
  infected ~ trap_type2 +
    s(habitat_PC1, k = 4) +
    s(disease_week, bs = "tp", k = 8) +
    s(disease_week, by = year_f, bs = "fs", k = 15) +
    s(year_f, bs = "re") +
    s(site_code, bs = "re") +
    offset(log(num_count)),
  family = binomial(link = "cloglog"),
  data = tarsalis_data,
  method = "REML"
)
summary(gam_tarsalis2_trapfixed)

AIC(gam_tarsalis, gam_tarsalis2, gam_tarsalis_trapfixed, gam_tarsalis2_trapfixed)

plot(
  gam_tarsalis2_trapfixed,
  select = 1,
  shade = TRUE,
  residuals = TRUE,
  rug = TRUE
)

# tarsalis diagnostics
par(mfrow = c(2, 2))
gam.check(gam_tarsalis_trapfixed)
par(mfrow = c(1, 1))
plot(gam_tarsalis_trapfixed, pages = 1, residuals = TRUE, shade = TRUE, seWithMean = TRUE)

# Concurvity
concurvity(gam_tarsalis, full = FALSE)

# tarsalis diagnostics
par(mfrow = c(2, 2))
gam.check(gam_tarsalis2_trapfixed)
par(mfrow = c(1, 1))
plot(gam_tarsalis2_trapfixed, pages = 1, residuals = TRUE, shade = TRUE, seWithMean = TRUE)

# Concurvity
concurvity(gam_tarsalis2_trapfixed, full = FALSE)

```

### Tar ~ s(habitat_PC1) + s(trap_type2)

Split data set into 2 trap types and allow smooth to vary across trap types... check that the curves look the same, check that the effect sizes match what we found before. 

```{r tar-by-trap}
# CO2+ only, allow smooth to vary across urbanization gradient (non-linear)
tarsalis_data_CO2 <- tarsalis_data %>%
  filter(trap_type2 == "CO2+")

gam_tarsalis2_trapfixed_CO2 <- gam(
  infected ~
    s(habitat_PC1, k = 4) +
    s(disease_week, bs = "tp", k = 8) +
    s(disease_week, by = year_f, bs = "fs", k = 15) +
    s(year_f, bs = "re") +
    s(site_code, bs = "re") +
    offset(log(num_count)),
  family = binomial(link = "cloglog"),
  data = tarsalis_data_CO2,
  method = "REML"
)

summary(gam_tarsalis2_trapfixed_CO2)

# GRVD+ only, allow smooth to vary across urbanization gradient (non-linear)
gam_tarsalis2_trapfixed_GRVD <- gam(
  infected ~ s(habitat_PC1, k = 4) +
    s(disease_week, bs = "tp", k = 8) +
    s(disease_week, by = year_f, bs = "fs", k = 15) +
    s(year_f, bs = "re") +
    s(site_code, bs = "re") +
    offset(log(num_count)),
  family = binomial(link = "cloglog"),
  data = tarsalis_data[tarsalis_data$trap_type2 == "GRVD+",],
  method = "REML"
)
summary(gam_tarsalis2_trapfixed_GRVD)


# Predicted WNV positivity through time from separate CO2+ and GRVD+ models

pc1_values <- tibble::tribble(
  ~habitat_level, ~habitat_PC1,
  "Rural", quantile(wnv$habitat_PC1, 0.9, na.rm = TRUE),
  "Peri",  median(wnv$habitat_PC1, na.rm = TRUE),
  "Urban", quantile(wnv$habitat_PC1, 0.1, na.rm = TRUE)
)

week_seq <- seq(
  min(wnv$disease_week, na.rm = TRUE),
  max(wnv$disease_week, na.rm = TRUE),
  by = 1
)

pred_co2 <- expand.grid(
  disease_week = week_seq,
  year_f = "2021",
  site_code = levels(tarsalis_data$site_code)[1],
  num_count = median(tarsalis_data$num_count, na.rm = TRUE)
) %>%
  left_join(pc1_values, by = character())

pred_grvd <- pred_co2

co2_pred <- predict(
  gam_tarsalis2_trapfixed_CO2,
  newdata = pred_co2,
  type = "link",
  se.fit = TRUE,
  exclude = c("s(site_code)", "s(year_f)")
)

grvd_pred <- predict(
  gam_tarsalis2_trapfixed_GRVD,
  newdata = pred_grvd,
  type = "link",
  se.fit = TRUE,
  exclude = c("s(site_code)", "s(year_f)")
)

pred_co2 <- pred_co2 %>%
  mutate(
    fit = 1 - exp(-exp(co2_pred$fit)),
    lower = 1 - exp(-exp(co2_pred$fit - 1.96 * co2_pred$se.fit)),
    upper = 1 - exp(-exp(co2_pred$fit + 1.96 * co2_pred$se.fit)),
    trap_type2 = "CO2+"
  )

pred_grvd <- pred_grvd %>%
  mutate(
    fit = 1 - exp(-exp(grvd_pred$fit)),
    lower = 1 - exp(-exp(grvd_pred$fit - 1.96 * grvd_pred$se.fit)),
    upper = 1 - exp(-exp(grvd_pred$fit + 1.96 * grvd_pred$se.fit)),
    trap_type2 = "GRVD+"
  )

pred_pip_week <- bind_rows(pred_co2, pred_grvd) %>%
  mutate(
    habitat_level = factor(habitat_level, levels = c("Rural", "Peri", "Urban")),
    trap_type2 = factor(trap_type2, levels = c("CO2+", "GRVD+"))
  )

urb_colors <- c(
  "Urban" = "#B22222",
  "Peri"  = "#DAA520",
  "Rural" = "#228B22"
)

p_co2 <- pred_pip_week %>%
  filter(trap_type2 == "CO2+") %>%
  ggplot(
    aes(
      x = disease_week,
      y = fit * 100,
      color = habitat_level,
      fill = habitat_level
    )
  ) +
  geom_ribbon(
    aes(
      ymin = lower * 100,
      ymax = upper * 100
    ),
    alpha = 0.15,
    color = NA
  ) +
  geom_line(linewidth = 1.2) +
  facet_wrap(~ habitat_level, ncol = 1, scales = "free_y") +
  scale_color_manual(values = urb_colors, name = "Habitat") +
  scale_fill_manual(values = urb_colors, guide = "none") +
  labs(
    title = "Cx. tarsalis CO2+ traps",
    x = "Disease week",
    y = "Predicted WNV positivity (%)"
  ) +
  theme_classic(base_size = 10) +
  theme(
    strip.background = element_blank(),
    legend.position = "none",
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

p_grvd <- pred_pip_week %>%
  filter(trap_type2 == "GRVD+") %>%
  ggplot(
    aes(
      x = disease_week,
      y = fit * 100,
      color = habitat_level,
      fill = habitat_level
    )
  ) +
  geom_ribbon(
    aes(
      ymin = lower * 100,
      ymax = upper * 100
    ),
    alpha = 0.15,
    color = NA
  ) +
  geom_line(linewidth = 1.2) +
  facet_wrap(~ habitat_level, ncol = 1, scales = "free_y") +
  scale_color_manual(values = urb_colors, name = "Habitat") +
  scale_fill_manual(values = urb_colors, guide = "none") +
  labs(
    title = "Cx. tarsalis GRVD+ traps",
    x = "Disease week",
    y = "Predicted WNV positivity (%)"
  ) +
  theme_classic(base_size = 10) +
  theme(
    strip.background = element_blank(),
    legend.position = "none",
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

p_co2
p_grvd
```

Interpretation:

Trap_type is not significant in any of the analysis. Neither is habitat_PC1. Disease week is significant in the simplest model, gam_tarsalis.  


### Tar interpretation  

For Cx. tarsalis, neither urbanization nor trap type explained variation in WNV positivity. In addition, GRVD collections contained relatively few Cx. tarsalis compared with CO2 collections, limiting inference from GRVD-based analyses. To maximize comparability between species while focusing on the most data-rich component of the dataset, subsequent analyses emphasized CO2 collections. This approach is also consistent with the primary ecological signal observed in Cx. pipiens, where urban CO2 collections exhibited the highest predicted WNV positivity.  

### Tar Moran's I for Cx. tarsalis residuals
```{r moran-tar}
# for CO2 model
tarsalis_data_CO2 <- tarsalis_data[tarsalis_data$trap_type2 == "CO2+",]

tarsalis_resids <- tarsalis_data_CO2 %>%
  mutate(resid = residuals(gam_tarsalis2_trapfixed_CO2, type = "deviance")) %>%
  group_by(site_code) %>%
  summarise(
    mean_resid = mean(resid, na.rm = TRUE),
    latitude = first(latitude),
    longitude = first(longitude),
    .groups = "drop"
  ) %>%
  filter(!is.na(mean_resid), !is.na(latitude), !is.na(longitude))

coords_tarsalis <- cbind(tarsalis_resids$longitude, tarsalis_resids$latitude)

lw5_tarsalis <- coords_tarsalis %>%
  knearneigh(k = 5) %>%
  knn2nb() %>%
  nb2listw(style = "W")

moran.test(tarsalis_resids$mean_resid, lw5_tarsalis)
```

**Interpretation:**
No evidence of spatial autocorrelation in tarsalis. 

## Plot predicted  



SUMMARY TABLE:
Effect	| Cx. pipiens |	Cx. tarsalis
Urbanization (habitat_PC1)	| Significant	| Not significant
Trap type (CO2 vs GRVD) |	Significant	| Not significant
Seasonal effect (disease week)	| Significant	| Significant
Year effect	| Significant	| Significant
Year-specific seasonal smooths |	Not significant |	Not significant

STRONGEST RESULTS:
Urbanization matters for Cx. pipiens
Urbanization does not matter for Cx. tarsalis
Both species show seasonal dynamics
Pipiens peaks earlier; tarsalis peaks later
Urban sites have the highest predicted WNV positivity

INTERPRETATION:  
 - Cx. pipiens WNV positivity varies with urbanization, trap type, season, and year.  
 - Cx. tarsalis WNV positivity varies with season and year, but shows no detectable urbanization or trap-type effect.  
  - The strongest ecological signal is therefore in Cx. pipiens, where predicted positivity is highest in urban habitats and differs between trap types.  
 - Because trap type influences pipiens but not tarsalis, comparisons between species should be standardized to a single trap type (CO2).  
 - Because urbanization influences pipiens but not tarsalis, the urbanization figure is most informative when focused on pipiens, with tarsalis reported as a non-significant result in the text.  
 - The year-specific smooths contribute little, suggesting that seasonal timing is broadly consistent among years even though overall infection intensity differs among years.
 

### Pipiens by urbanization for CO2

Figure 1: Cx. pipiens CO2-only predicted WNV positivity across Urban / Peri / Rural.

Visualize predicted urbanization effects using CO2+ collections for pipiens only, where those effects are significant.  

```{r pred-week}
### Figure Cx. pipiens CO2-only predicted WNV positivity by disease week

cols <- c(
  "Cx. pipiens s.l." = "#1bc8ea"
)

pc1_values <- tibble::tribble(
  ~habitat_level, ~habitat_PC1,
  "Rural", quantile(wnv$habitat_PC1, 0.9, na.rm = TRUE),
  "Peri",  median(wnv$habitat_PC1, na.rm = TRUE),
  "Urban", quantile(wnv$habitat_PC1, 0.1, na.rm = TRUE)
)

week_seq <- seq(
  min(wnv$disease_week, na.rm = TRUE),
  max(wnv$disease_week, na.rm = TRUE),
  by = 1
)

pred_pip_co2_week <- expand.grid(
  disease_week = week_seq,
  habitat_PC1 = pc1_values$habitat_PC1,
  year_f = factor("2021", levels = levels(pipiens_data_CO2$year_f)),
  site_code = factor(
    levels(pipiens_data_CO2$site_code)[1],
    levels = levels(pipiens_data_CO2$site_code)
  ),
  num_count = median(pipiens_data_CO2$num_count, na.rm = TRUE)
) %>%
  left_join(pc1_values, by = "habitat_PC1")

pip_pred <- predict(
  gam_pipiens2_trapfixed_CO2,
  newdata = pred_pip_co2_week,
  type = "link",
  se.fit = TRUE,
  exclude = c("s(site_code)", "s(year_f)")
)

pred_wnv_pip_week <- pred_pip_co2_week %>%
  mutate(
    fit = 1 - exp(-exp(pip_pred$fit)),
    lower = 1 - exp(-exp(pip_pred$fit - 1.96 * pip_pred$se.fit)),
    upper = 1 - exp(-exp(pip_pred$fit + 1.96 * pip_pred$se.fit)),
    species = "Cx. pipiens s.l.",
    habitat_level = factor(habitat_level, levels = c("Rural", "Peri", "Urban"))
  )

ggplot(
  pred_wnv_pip_week,
  aes(
    x = disease_week,
    y = fit * 100,
    color = species,
    fill = species
  )
) +
  geom_ribbon(
    aes(
      ymin = lower * 100,
      ymax = upper * 100,
      group = species
    ),
    alpha = 0.25,
    color = NA
  ) +
  geom_line(linewidth = 1.4) +
  geom_vline(
    xintercept = 33,
    linetype = "dashed",
    linewidth = 0.8
  ) +
  geom_text(
    data = data.frame(
      habitat_level = factor(c("Rural", "Peri", "Urban"),
                             levels = c("Rural", "Peri", "Urban")),
      disease_week = 33.3,
      y = Inf
    ),
    aes(
      x = disease_week,
      y = y,
      label = "1st WNV cases"
    ),
    inherit.aes = FALSE,
    hjust = 0,
    vjust = 1.3,
    size = 4
  ) +
  facet_wrap(
    ~ habitat_level,
    ncol = 1,
    scales = "free_y"
  ) +
  scale_color_manual(values = cols, name = "Species") +
  scale_fill_manual(values = cols, name = "Species") +
  labs(
    title = "Predicted WNV positivity in Cx. pipiens s.l. across the urbanization gradient",
    x = "Disease week",
    y = "Predicted WNV positivity (%)"
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5),
    strip.background = element_blank(),
    strip.text = element_text(size = 14),
    legend.position = "none"
  )
```
Not very informative because the shape for 2021. looks pretty identical, magnitude is different...

### Seasonal dynamics

Figure 2. Seasonal dynamics of WNV positivity in Cx. pipiens and Cx. tarsalis.

```{r season-species}
### Figure 2: Seasonal dynamics of WNV positivity by species
### CO2 traps, habitat_PC1 held at median, one panel per year

cols <- c(
  "Cx. pipiens s.l." = "#1bc8ea",
  "Cx. tarsalis" = "#FF2DA0"
)

week_seq <- seq(
  min(wnv$disease_week, na.rm = TRUE),
  max(wnv$disease_week, na.rm = TRUE),
  by = 1
)

pred_season_base <- expand.grid(
  disease_week = week_seq,
  habitat_PC1 = median(wnv$habitat_PC1, na.rm = TRUE),
  trap_type2 = "CO2+",
  year_f = levels(wnv$year_f),
  num_count = 1
)

pred_pip <- pred_season_base %>%
  mutate(
    year_f = factor(year_f, levels = levels(pipiens_data$year_f)),
    trap_type2 = factor(trap_type2, levels = levels(pipiens_data$trap_type2)),
    site_code = factor(
      levels(pipiens_data$site_code)[1],
      levels = levels(pipiens_data$site_code)
    )
  )

pred_tar <- pred_season_base %>%
  mutate(
    year_f = factor(year_f, levels = levels(tarsalis_data$year_f)),
    trap_type2 = factor(trap_type2, levels = levels(tarsalis_data$trap_type2)),
    site_code = factor(
      levels(tarsalis_data$site_code)[1],
      levels = levels(tarsalis_data$site_code)
    )
  )

pip_pred <- predict(
  gam_pipiens_trapfixed,
  newdata = pred_pip,
  type = "link",
  se.fit = TRUE,
  exclude = "s(site_code)"
)

tar_pred <- predict(
  gam_tarsalis_trapfixed,
  newdata = pred_tar,
  type = "link",
  se.fit = TRUE,
  exclude = "s(site_code)"
)

pred_season <- bind_rows(
  pred_pip %>%
    mutate(
      fit = 1 - exp(-exp(pip_pred$fit)),
      lower = 1 - exp(-exp(pip_pred$fit - 1.96 * pip_pred$se.fit)),
      upper = 1 - exp(-exp(pip_pred$fit + 1.96 * pip_pred$se.fit)),
      species = "Cx. pipiens s.l."
    ),
  pred_tar %>%
    mutate(
      fit = 1 - exp(-exp(tar_pred$fit)),
      lower = 1 - exp(-exp(tar_pred$fit - 1.96 * tar_pred$se.fit)),
      upper = 1 - exp(-exp(tar_pred$fit + 1.96 * tar_pred$se.fit)),
      species = "Cx. tarsalis"
    )
)

ggplot(
  pred_season,
  aes(
    x = disease_week,
    y = fit * 100,
    color = species,
    fill = species
  )
) +
  geom_ribbon(
    aes(ymin = lower * 100, ymax = upper * 100, group = species),
    alpha = 0.22,
    color = NA
  ) +
  geom_line(aes(group = species), linewidth = 1.3) +
  geom_vline(xintercept = 33, linetype = "dashed", linewidth = 0.7) +
  facet_wrap(~ year_f, ncol = 3, scales = "free_y") +
  coord_cartesian(ylim = c(0, 1.5)) +
  scale_color_manual(values = cols, name = "Species") +
  scale_fill_manual(values = cols, guide = "none") +
  labs(
    title = "Seasonal dynamics of predicted WNV positivity",
    x = "Disease week",
    y = "Predicted WNV positivity (%)"
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5),
    strip.background = element_blank(),
    legend.position = "right"
  )
```  
### Pip by trap
Seasonal dynamics of WNV positivity in Cx. pipiens by trap type
habitat_PC1 held at median, one panel per year

```{r pip-season-traps}


pip_col <- "#1bc8ea"

week_seq <- seq(
  min(wnv$disease_week, na.rm = TRUE),
  max(wnv$disease_week, na.rm = TRUE),
  by = 1
)

pred_pip_trap <- expand.grid(
  disease_week = week_seq,
  habitat_PC1 = median(wnv$habitat_PC1, na.rm = TRUE),
  trap_type2 = levels(pipiens_data$trap_type2),
  year_f = levels(wnv$year_f),
  num_count = 1
) %>%
  mutate(
    year_f = factor(year_f, levels = levels(pipiens_data$year_f)),
    trap_type2 = factor(trap_type2, levels = levels(pipiens_data$trap_type2)),
    site_code = factor(
      levels(pipiens_data$site_code)[1],
      levels = levels(pipiens_data$site_code)
    )
  )

pip_pred <- predict(
  gam_pipiens_trapfixed,
  newdata = pred_pip_trap,
  type = "link",
  se.fit = TRUE,
  exclude = "s(site_code)"
)

pred_pip_trap <- pred_pip_trap %>%
  mutate(
    fit = 1 - exp(-exp(pip_pred$fit)),
    lower = 1 - exp(-exp(pip_pred$fit - 1.96 * pip_pred$se.fit)),
    upper = 1 - exp(-exp(pip_pred$fit + 1.96 * pip_pred$se.fit))
  )

ggplot(
  pred_pip_trap,
  aes(
    x = disease_week,
    y = fit * 100,
    linetype = trap_type2,
    group = trap_type2
  )
) +
  geom_ribbon(
    aes(
      ymin = lower * 100,
      ymax = upper * 100,
      group = trap_type2
    ),
    fill = pip_col,
    alpha = 0.16,
    color = NA
  ) +
  geom_line(
    color = pip_col,
    linewidth = 1.3
  ) +
  geom_vline(xintercept = 33, linetype = "dashed", linewidth = 0.7) +
  facet_wrap(~ year_f, ncol = 3, scales = "free_y") +
  coord_cartesian(ylim = c(0, 2.5)) +
  scale_linetype_manual(
    values = c(
      "CO2+" = "solid",
      "GRVD+" = "dashed"
    ),
    name = "Trap type"
  ) +
  labs(
    title = "Seasonal dynamics of predicted WNV positivity in Cx. pipiens s.l.",
    x = "Disease week",
    y = "Predicted WNV positivity (%)"
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5),
    strip.background = element_blank(),
    legend.position = "right"
  )
```


### Week 32:
```{r}
pred_compare <- expand.grid(
  habitat_level = c("Rural", "Peri", "Urban"),
  year_f = levels(wnv$year_f)
) %>%
  mutate(
    habitat_PC1 = case_when(
      habitat_level == "Rural" ~ quantile(wnv$habitat_PC1, 0.9, na.rm = TRUE),
      habitat_level == "Peri"  ~ median(wnv$habitat_PC1, na.rm = TRUE),
      habitat_level == "Urban" ~ quantile(wnv$habitat_PC1, 0.1, na.rm = TRUE)
    ),
    disease_week = 32,
    trap_type2 = "CO2+",
    num_count = 1,
    habitat_level = factor(habitat_level, levels = c("Rural", "Peri", "Urban"))
  )

pred_pip <- pred_compare %>%
  mutate(
    year_f = factor(year_f, levels = levels(pipiens_data$year_f)),
    trap_type2 = factor(trap_type2, levels = levels(pipiens_data$trap_type2)),
    site_code = factor(levels(pipiens_data$site_code)[1],
                       levels = levels(pipiens_data$site_code))
  )

pred_tar <- pred_compare %>%
  mutate(
    year_f = factor(year_f, levels = levels(tarsalis_data$year_f)),
    trap_type2 = factor(trap_type2, levels = levels(tarsalis_data$trap_type2)),
    site_code = factor(levels(tarsalis_data$site_code)[1],
                       levels = levels(tarsalis_data$site_code))
  )

pip <- predict(
  gam_pipiens_trapfixed,
  newdata = pred_pip,
  type = "link",
  se.fit = TRUE,
  exclude = "s(site_code)"
)

tar <- predict(
  gam_tarsalis_trapfixed,
  newdata = pred_tar,
  type = "link",
  se.fit = TRUE,
  exclude = "s(site_code)"
)

pred_species <- bind_rows(
  pred_pip %>%
    mutate(
      fit = 1 - exp(-exp(pip$fit)),
      lower = 1 - exp(-exp(pip$fit - 1.96 * pip$se.fit)),
      upper = 1 - exp(-exp(pip$fit + 1.96 * pip$se.fit)),
      species = "Cx. pipiens s.l."
    ),
  pred_tar %>%
    mutate(
      fit = 1 - exp(-exp(tar$fit)),
      lower = 1 - exp(-exp(tar$fit - 1.96 * tar$se.fit)),
      upper = 1 - exp(-exp(tar$fit + 1.96 * tar$se.fit)),
      species = "Cx. tarsalis"
    )
)

#six panel plot by year
ggplot(
  pred_species,
  aes(
    x = habitat_level,
    y = fit * 100,
    color = species,
    group = species
  )
) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(
      ymin = lower * 100,
      ymax = upper * 100
    ),
    width = 0.1,
    linewidth = 0.7
  ) +
  facet_wrap(~ year_f, ncol = 3, scales = "free_y") +
  scale_color_manual(values = cols, name = "Species") +
  labs(
    title = "Predicted WNV by urbanization and year (Week32 & CO2+)",
    x = "Urbanization",
    y = "Predicted WNV positivity (%)"
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5),
    strip.background = element_blank(),
    legend.position = "right"
  )
pred_species_avg <- pred_species %>%
  group_by(species, habitat_level) %>%
  summarise(
    fit = mean(fit, na.rm = TRUE),
    lower = mean(lower, na.rm = TRUE),
    upper = mean(upper, na.rm = TRUE),
    .groups = "drop"
  )
pred_species_avg <- pred_species %>%
  group_by(species, habitat_level) %>%
  summarise(
    fit = mean(fit, na.rm = TRUE),
    lower = mean(lower, na.rm = TRUE),
    upper = mean(upper, na.rm = TRUE),
    .groups = "drop"
  )
ggplot(
  pred_species_avg,
  aes(
    x = habitat_level,
    y = fit * 100,
    color = species,
    group = species
  )
) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 4) +
  geom_errorbar(
    aes(
      ymin = lower * 100,
      ymax = upper * 100
    ),
    width = 0.1,
    linewidth = 0.8
  ) +
  scale_color_manual(values = cols, name = "Species") +
  labs(
    title = "Average predicted WNV across years (Week32 & CO2+)",
    x = "Urbanization",
    y = "Predicted WNV positivity (%)"
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "right"
  )

```
**NOTE:** Averaging lower and upper is for visualization only, represents mean model-predicted intervals across years, not formal confidence intervals for the averaged prediction.

Also, the urbanization effect is not significant for tarsalis, so I'm including it here only for comparison with pipiens.

### Week 35:
```{r}
pred_compare <- expand.grid(
  habitat_level = c("Rural", "Peri", "Urban"),
  year_f = levels(wnv$year_f)
) %>%
  mutate(
    habitat_PC1 = case_when(
      habitat_level == "Rural" ~ quantile(wnv$habitat_PC1, 0.9, na.rm = TRUE),
      habitat_level == "Peri"  ~ median(wnv$habitat_PC1, na.rm = TRUE),
      habitat_level == "Urban" ~ quantile(wnv$habitat_PC1, 0.1, na.rm = TRUE)
    ),
    disease_week = 35,
    trap_type2 = "CO2+",
    num_count = 1,
    habitat_level = factor(habitat_level, levels = c("Rural", "Peri", "Urban"))
  )

pred_pip <- pred_compare %>%
  mutate(
    year_f = factor(year_f, levels = levels(pipiens_data$year_f)),
    trap_type2 = factor(trap_type2, levels = levels(pipiens_data$trap_type2)),
    site_code = factor(levels(pipiens_data$site_code)[1],
                       levels = levels(pipiens_data$site_code))
  )

pred_tar <- pred_compare %>%
  mutate(
    year_f = factor(year_f, levels = levels(tarsalis_data$year_f)),
    trap_type2 = factor(trap_type2, levels = levels(tarsalis_data$trap_type2)),
    site_code = factor(levels(tarsalis_data$site_code)[1],
                       levels = levels(tarsalis_data$site_code))
  )

pip <- predict(
  gam_pipiens_trapfixed,
  newdata = pred_pip,
  type = "link",
  se.fit = TRUE,
  exclude = "s(site_code)"
)

tar <- predict(
  gam_tarsalis_trapfixed,
  newdata = pred_tar,
  type = "link",
  se.fit = TRUE,
  exclude = "s(site_code)"
)

pred_species <- bind_rows(
  pred_pip %>%
    mutate(
      fit = 1 - exp(-exp(pip$fit)),
      lower = 1 - exp(-exp(pip$fit - 1.96 * pip$se.fit)),
      upper = 1 - exp(-exp(pip$fit + 1.96 * pip$se.fit)),
      species = "Cx. pipiens s.l."
    ),
  pred_tar %>%
    mutate(
      fit = 1 - exp(-exp(tar$fit)),
      lower = 1 - exp(-exp(tar$fit - 1.96 * tar$se.fit)),
      upper = 1 - exp(-exp(tar$fit + 1.96 * tar$se.fit)),
      species = "Cx. tarsalis"
    )
)

#six panel plot by year
ggplot(
  pred_species,
  aes(
    x = habitat_level,
    y = fit * 100,
    color = species,
    group = species
  )
) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(
      ymin = lower * 100,
      ymax = upper * 100
    ),
    width = 0.1,
    linewidth = 0.7
  ) +
  facet_wrap(~ year_f, ncol = 3, scales = "free_y") +
  scale_color_manual(values = cols, name = "Species") +
  labs(
    title = "Predicted WNV by urbanization and year (Week35 & CO2+)",
    x = "Urbanization",
    y = "Predicted WNV positivity (%)"
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5),
    strip.background = element_blank(),
    legend.position = "right"
  )
pred_species_avg <- pred_species %>%
  group_by(species, habitat_level) %>%
  summarise(
    fit = mean(fit, na.rm = TRUE),
    lower = mean(lower, na.rm = TRUE),
    upper = mean(upper, na.rm = TRUE),
    .groups = "drop"
  )
pred_species_avg <- pred_species %>%
  group_by(species, habitat_level) %>%
  summarise(
    fit = mean(fit, na.rm = TRUE),
    lower = mean(lower, na.rm = TRUE),
    upper = mean(upper, na.rm = TRUE),
    .groups = "drop"
  )
ggplot(
  pred_species_avg,
  aes(
    x = habitat_level,
    y = fit * 100,
    color = species,
    group = species
  )
) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 4) +
  geom_errorbar(
    aes(
      ymin = lower * 100,
      ymax = upper * 100
    ),
    width = 0.1,
    linewidth = 0.8
  ) +
  scale_color_manual(values = cols, name = "Species") +
  labs(
    title = "Average predicted WNV across years (Week35 & CO2+)",
    x = "Urbanization",
    y = "Predicted WNV positivity (%)"
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "right"
  )

```
  
### Week 38:
```{r}
pred_compare <- expand.grid(
  habitat_level = c("Rural", "Peri", "Urban"),
  year_f = levels(wnv$year_f)
) %>%
  mutate(
    habitat_PC1 = case_when(
      habitat_level == "Rural" ~ quantile(wnv$habitat_PC1, 0.9, na.rm = TRUE),
      habitat_level == "Peri"  ~ median(wnv$habitat_PC1, na.rm = TRUE),
      habitat_level == "Urban" ~ quantile(wnv$habitat_PC1, 0.1, na.rm = TRUE)
    ),
    disease_week = 38,
    trap_type2 = "CO2+",
    num_count = 1,
    habitat_level = factor(habitat_level, levels = c("Rural", "Peri", "Urban"))
  )

pred_pip <- pred_compare %>%
  mutate(
    year_f = factor(year_f, levels = levels(pipiens_data$year_f)),
    trap_type2 = factor(trap_type2, levels = levels(pipiens_data$trap_type2)),
    site_code = factor(levels(pipiens_data$site_code)[1],
                       levels = levels(pipiens_data$site_code))
  )

pred_tar <- pred_compare %>%
  mutate(
    year_f = factor(year_f, levels = levels(tarsalis_data$year_f)),
    trap_type2 = factor(trap_type2, levels = levels(tarsalis_data$trap_type2)),
    site_code = factor(levels(tarsalis_data$site_code)[1],
                       levels = levels(tarsalis_data$site_code))
  )

pip <- predict(
  gam_pipiens_trapfixed,
  newdata = pred_pip,
  type = "link",
  se.fit = TRUE,
  exclude = "s(site_code)"
)

tar <- predict(
  gam_tarsalis_trapfixed,
  newdata = pred_tar,
  type = "link",
  se.fit = TRUE,
  exclude = "s(site_code)"
)

pred_species <- bind_rows(
  pred_pip %>%
    mutate(
      fit = 1 - exp(-exp(pip$fit)),
      lower = 1 - exp(-exp(pip$fit - 1.96 * pip$se.fit)),
      upper = 1 - exp(-exp(pip$fit + 1.96 * pip$se.fit)),
      species = "Cx. pipiens s.l."
    ),
  pred_tar %>%
    mutate(
      fit = 1 - exp(-exp(tar$fit)),
      lower = 1 - exp(-exp(tar$fit - 1.96 * tar$se.fit)),
      upper = 1 - exp(-exp(tar$fit + 1.96 * tar$se.fit)),
      species = "Cx. tarsalis"
    )
)

#six panel plot by year
ggplot(
  pred_species,
  aes(
    x = habitat_level,
    y = fit * 100,
    color = species,
    group = species
  )
) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(
      ymin = lower * 100,
      ymax = upper * 100
    ),
    width = 0.1,
    linewidth = 0.7
  ) +
  facet_wrap(~ year_f, ncol = 3, scales = "free_y") +
  scale_color_manual(values = cols, name = "Species") +
  labs(
    title = "Predicted WNV by urbanization and year (Week38 & CO2+)",
    x = "Urbanization",
    y = "Predicted WNV positivity (%)"
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5),
    strip.background = element_blank(),
    legend.position = "right"
  )
pred_species_avg <- pred_species %>%
  group_by(species, habitat_level) %>%
  summarise(
    fit = mean(fit, na.rm = TRUE),
    lower = mean(lower, na.rm = TRUE),
    upper = mean(upper, na.rm = TRUE),
    .groups = "drop"
  )
pred_species_avg <- pred_species %>%
  group_by(species, habitat_level) %>%
  summarise(
    fit = mean(fit, na.rm = TRUE),
    lower = mean(lower, na.rm = TRUE),
    upper = mean(upper, na.rm = TRUE),
    .groups = "drop"
  )
ggplot(
  pred_species_avg,
  aes(
    x = habitat_level,
    y = fit * 100,
    color = species,
    group = species
  )
) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 4) +
  geom_errorbar(
    aes(
      ymin = lower * 100,
      ymax = upper * 100
    ),
    width = 0.1,
    linewidth = 0.8
  ) +
  scale_color_manual(values = cols, name = "Species") +
  labs(
    title = "Average predicted WNV across years (Week38 & CO2+)",
    x = "Urbanization",
    y = "Predicted WNV positivity (%)"
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "right"
  )

```

# RQ3: Bloodmeal and WNV probability  

Maybe this is just a question of interpretation, and showing how these aspects line up:

1. Cx. pipiens blood feeding on passerine is correlated with urbanization

2. Cx. pipiens WNV positiveity is correlated with urbanization

3. Cx. pipeins WNV positivity is higher than for Cx. tarsalis in peri and urban areas at week 33, time of first WNV cases every year. Also Cx. tarsalis has no significant effect of urbanization, or trap effect. Also consider abundance in traps in interpretation.






