---
title: "Bloodmeal + WNV positive GAM: SLC"
author: "Norah Saarman"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

# Setup
```{r 1-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  # Saves PNGs to the root /figures folder
  fig.path = "../figures/" 
)
```

```{r packages, warning=FALSE}
library(dplyr)
library(brms)
library(mgcv)
library(bayesplot)
library(ggplot2)
library(sf)
library(spdep)
library(tidyr)
library(maptiles)
library(ggspatial)
library(prettymapr)
library(tibble)

```

# Data 

Started with a folder in Emily's home directory called "saarman-group". I cleaned them up and added habitat data:  
 - "bloodmeal_master_FINAL.csv" --> "bloodmeal_Jun2026_NPS.csv"
 - "wnv_master_FINAL.csv" --> "wnv_Jun2026_NPS.csv"
 
 
```{r data}
bloodmeal <- read.csv("/uufs/chpc.utah.edu/common/home/saarman-group1/urbanindex/bloodmeal_Jun2026_NPS.csv")

wnv <- read.csv("/uufs/chpc.utah.edu/common/home/saarman-group1/urbanindex/wnv_Jun2026_NPS.csv")

```
# PCA of habitat/urbanization gradient  


### PCA - bloodmeal GPS points
```{r PCA-bloodmeal}
# Check mosq_species breakdown
table(bloodmeal$mosq_species,bloodmeal$urbanization)

# Pool trap types and check breakdown by trap
bloodmeal$trap_type2 <- bloodmeal$trap_type

# Mutate trap_type to pool walk-in and trash can
bloodmeal <- bloodmeal %>%
  mutate(trap_type2 = case_when(
    trap_type2 == "Walk-in" ~ "GRVD+",
    TRUE                     ~ trap_type2
  ))

bloodmeal <- bloodmeal %>%
  mutate(trap_type2 = case_when(
    trap_type2 == "Trash Can" ~ "GRVD+",
    TRUE                     ~ trap_type2
  ))

bloodmeal <- bloodmeal %>%
  mutate(trap_type2 = case_when(
    trap_type2 == "GRVD" ~ "GRVD+",
    TRUE                     ~ trap_type2
  ))

bloodmeal <- bloodmeal %>%
  mutate(trap_type2 = case_when(
    trap_type2 == "BOX" ~ "GRVD+",
    TRUE                     ~ trap_type2
  ))

table(bloodmeal$mosq_species,bloodmeal$urbanization,bloodmeal$trap_type2)

# Use only sites represented in the bloodmeal dataset to define the PCA
bloodmeal_habitat_sites <- bloodmeal %>%
  filter(site_code != "0") %>%
  distinct(
    site_code,
    impervious_500,
    canopy_500,
    dist_wetland_m,
    summer_ndvi_500
  ) %>%
  filter(complete.cases(
    impervious_500,
    canopy_500,
    dist_wetland_m,
    summer_ndvi_500
  ))

# Run PCA
bloodmeal_pca_vars <- bloodmeal_habitat_sites %>%
  dplyr::select(
    impervious_500,
    canopy_500,
    dist_wetland_m,
    summer_ndvi_500
  )

bloodmeal_pca <- prcomp(
  bloodmeal_pca_vars,
  center = TRUE,
  scale. = TRUE
)

summary(bloodmeal_pca)
round(bloodmeal_pca$rotation, 3)

# Add PCA scores to site-level table
bloodmeal_habitat_sites <- bloodmeal_habitat_sites %>%
  mutate(
    habitat_PC1 = bloodmeal_pca$x[,1],
    habitat_PC2 = bloodmeal_pca$x[,2]
  )

# Join PCA scores back to full WNV table
bloodmeal <- bloodmeal %>%
  dplyr::select(-any_of(c("habitat_PC1", "habitat_PC2"))) %>%
  left_join(
    bloodmeal_habitat_sites %>%
      dplyr::select(site_code, habitat_PC1, habitat_PC2),
    by = "site_code"
  )

urb_colors <- c(
  "Urban" = "#B22222",
  "Peri"  = "#DAA520",
  "Rural" = "#228B22"
)

# Site/trap-level PCA scores from bloodmeal object
pca_scores_df <- bloodmeal %>%
  distinct(
    site_code,
    trap_type2,
    urbanization,
    habitat_PC1,
    habitat_PC2
  ) %>%
  filter(
    !is.na(habitat_PC1),
    !is.na(habitat_PC2),
    !is.na(trap_type2)
  ) %>%
  mutate(
    urbanization = factor(
      urbanization,
      levels = c("Urban", "Peri", "Rural")
    ),
    PC1 = habitat_PC1,
    PC2 = habitat_PC2
  )

# PCA variable loadings
loadings_df <- as.data.frame(bloodmeal_pca$rotation) %>%
  rownames_to_column("variable") %>%
  mutate(
    variable = recode(
      variable,
      "impervious_500"  = "Impervious surface",
      "canopy_500"      = "Tree canopy",
      "dist_wetland_m"  = "Wetland distance",
      "summer_ndvi_500" = "NDVI"
    )
  )

arrow_scale <- 3.5

p1_biplot <- ggplot() +
  stat_ellipse(
    data = pca_scores_df,
    aes(x = PC1, y = PC2, color = urbanization),
    level = 0.75,
    linewidth = 0.8,
    linetype = "dashed"
  ) +
  geom_point(
    data = pca_scores_df,
    aes(
      x = PC1,
      y = PC2,
      color = urbanization,
      fill = urbanization,
      shape = trap_type2
    ),
    size = 3,
    alpha = 0.5,
    stroke = 0.8
  ) +
  geom_segment(
    data = loadings_df,
    aes(
      x = 0,
      y = 0,
      xend = PC1 * arrow_scale,
      yend = PC2 * arrow_scale
    ),
    arrow = arrow(length = unit(0.25, "cm"), type = "closed"),
    color = "gray30",
    linewidth = 0.7
  ) +
  geom_text(
    data = loadings_df,
    aes(
      x = PC1 * arrow_scale * 1.15,
      y = PC2 * arrow_scale * 1.15,
      label = variable
    ),
    size = 3.5,
    color = "gray20",
    fontface = "italic"
  ) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "gray70") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "gray70") +
  scale_color_manual(values = urb_colors, name = "Urbanization") +
  scale_fill_manual(values = urb_colors, name = "Urbanization") +
  scale_shape_manual(
    values = c(
      "CO2" = 20,
      "GRVD+" = 24
    ),
    name = "Trap type"
  ) +
  labs(
    title = "Habitat PCA (Bloodmeal Sites)",
       x = "PC1 - Urbanization (57.6% variance)",
    y = "PC2 - Vegetation (29.7% variance)"
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "gray40"),
    legend.position = "right"
  )

print(p1_biplot)


```  


### PCA - WNV dataset

```{r PCA-wnv}

# Check mosq_species breakdown
table(wnv$mosq_species,wnv$urbanization, wnv$trap_type)

# Pool trap types and check breakdown by trap
wnv$trap_type2 <- wnv$trap_type

# Mutate trap_type to pool walk-in and trash can
wnv <- wnv %>%
  mutate(trap_type2 = case_when(
    trap_type2 == "BGSENT" ~ "CO2+",
    TRUE                     ~ trap_type2
  ))

wnv <- wnv %>%
  mutate(trap_type2 = case_when(
    trap_type2 == "CO2" ~ "CO2+",
    TRUE                     ~ trap_type2
  ))

wnv <- wnv %>%
  mutate(trap_type2 = case_when(
    trap_type2 == "GRVD" ~ "GRVD+",
    TRUE                     ~ trap_type2
  ))

wnv <- wnv %>%
  mutate(trap_type2 = case_when(
    trap_type2 == "OTHER" ~ "GRVD+",
    TRUE                     ~ trap_type2
  ))

wnv <- wnv %>%
  mutate(trap_type2 = case_when(
    trap_type2 == "REST" ~ "GRVD+",
    TRUE                     ~ trap_type2
  ))

table(wnv$mosq_species,wnv$urbanization,wnv$trap_type2)

# Use only sites represented in the WNV dataset to define the PCA
wnv_habitat_sites <- wnv %>%
  filter(site_code != "0") %>%
  distinct(
    site_code,
    impervious_500,
    canopy_500,
    dist_wetland_m,
    summer_ndvi_500
  ) %>%
  filter(complete.cases(
    impervious_500,
    canopy_500,
    dist_wetland_m,
    summer_ndvi_500
  ))

# Run PCA
wnv_pca_vars <- wnv_habitat_sites %>%
  dplyr::select(
    impervious_500,
    canopy_500,
    dist_wetland_m,
    summer_ndvi_500
  )

wnv_pca <- prcomp(
  wnv_pca_vars,
  center = TRUE,
  scale. = TRUE
)

summary(wnv_pca)
round(wnv_pca$rotation, 3)

# Add PCA scores to site-level table
wnv_habitat_sites <- wnv_habitat_sites %>%
  mutate(
    habitat_PC1 = wnv_pca$x[,1],
    habitat_PC2 = wnv_pca$x[,2]
  )

# Join PCA scores back to full WNV table
wnv <- wnv %>%
  dplyr::select(-any_of(c("habitat_PC1", "habitat_PC2"))) %>%
  left_join(
    wnv_habitat_sites %>%
      dplyr::select(site_code, habitat_PC1, habitat_PC2),
    by = "site_code"
  )

urb_colors <- c(
  "Urban" = "#B22222",
  "Peri"  = "#DAA520",
  "Rural" = "#228B22"
)

# Site/trap-level PCA scores from WNV object
pca_scores_df <- wnv %>%
  distinct(
    site_code,
    trap_type2,
    urbanization,
    habitat_PC1,
    habitat_PC2
  ) %>%
  filter(
    !is.na(habitat_PC1),
    !is.na(habitat_PC2),
    !is.na(trap_type2)
  ) %>%
  mutate(
    urbanization = factor(
      urbanization,
      levels = c("Urban", "Peri", "Rural")
    ),
    PC1 = habitat_PC1,
    PC2 = habitat_PC2
  )

# PCA variable loadings
loadings_df <- as.data.frame(wnv_pca$rotation) %>%
  rownames_to_column("variable") %>%
  mutate(
    variable = recode(
      variable,
      "impervious_500"  = "Impervious surface",
      "canopy_500"      = "Tree canopy",
      "dist_wetland_m"  = "Wetland distance",
      "summer_ndvi_500" = "NDVI"
    )
  )

arrow_scale <- 3.5

p2_biplot <- ggplot() +
  stat_ellipse(
    data = pca_scores_df,
    aes(x = PC1, y = PC2, color = urbanization),
    level = 0.75,
    linewidth = 0.8,
    linetype = "dashed"
  ) +
  geom_point(
    data = pca_scores_df,
    aes(
      x = PC1,
      y = PC2,
      color = urbanization,
      fill = urbanization,
      shape = trap_type2
    ),
    size = 3,
    alpha = 0.5,
    stroke = 0.8
  ) +
  geom_segment(
    data = loadings_df,
    aes(
      x = 0,
      y = 0,
      xend = PC1 * arrow_scale,
      yend = PC2 * arrow_scale
    ),
    arrow = arrow(length = unit(0.25, "cm"), type = "closed"),
    color = "gray30",
    linewidth = 0.7
  ) +
  geom_text(
    data = loadings_df,
    aes(
      x = PC1 * arrow_scale * 1.15,
      y = PC2 * arrow_scale * 1.15,
      label = variable
    ),
    size = 3.5,
    color = "gray20",
    fontface = "italic"
  ) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "gray70") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "gray70") +
  scale_color_manual(values = urb_colors, name = "Urbanization") +
  scale_fill_manual(values = urb_colors, name = "Urbanization") +
  scale_shape_manual(
    values = c(
      "CO2+" = 20,
      "GRVD+" = 24
    ),
    name = "Trap type"
  ) +
  labs(
    title = "Habitat PCA (WNV Sites)",
       x = "PC1 - Urbanization (57.6% variance)",
    y = "PC2 - Vegetation (29.7% variance)"
  ) +
  scale_x_reverse() +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "gray40"),
    legend.position = "right"
  )

print(p2_biplot)

```  
  
# RQ1: Bloodmeal  

```{r model1}
# MODEL 1: Bloodmeal host ~ mosquito species + urbanization
# Make sure it's a factor
bloodmeal <- bloodmeal %>%
mutate(host_3_categories = factor(host_3_categories))
# Check levels — brm will use the fir
levels(bloodmeal$host_3_categories)

# make "Other bird" reference category
bloodmeal <- bloodmeal %>%
mutate(host_3_categories = relevel(host_3_categories,
ref = "Other bird"))
levels(bloodmeal$host_3_categories)

# fit model
model_1 <- brm(
host_3_categories ~ mosq_species + habitat_PC1 + (1|site_code) + (1|year),
family = categorical(),
data = bloodmeal,
cores = 4,
chains = 4,
iter = 4000,
warmup = 2000,
control = list(adapt_delta = 0.99),
file = "../results/bloodmeal_host3_habitatPC1",
seed = 908275
)

summary(model_1)
pp_check(model_1, ndraws = 100) 
mcmc_trace(model_1, pars = vars(contains("habitat_PC1")))
mcmc_trace(model_1, pars = vars(contains("mosq_species")))
```
# RQ2: WNV probability

## Pipiens

### Pip model 2v1
```{r model2v1-pip}
# Pip MODEL 2v1: WNV infection ~ urbanization + GAM smooths

# Make sure all random effect variables are factors
wnv <- wnv %>%
mutate(
site_code = factor(site_code),
trap_type2 = factor(trap_type2),
year_f = factor(year)
)

pipiens_data <- wnv %>% filter(mosq_species == "Cx_pipiens_sl")

gam_pipiens <- gam(
infected ~ habitat_PC1 +
  s(disease_week, bs = "tp", k = 8) +
s(disease_week, by = year_f, bs = "fs", k = 15) +
s(year_f, bs = "re") +
s(site_code, bs = "re") +
s(trap_type2, bs = "re") + #trap_type2 is random effect
offset(log(num_count)),
family = binomial(link = "cloglog"),
data = pipiens_data,
method = "REML"
)
summary(gam_pipiens)

#allow smooth to vary across urbanization gradient (non-linear)
gam_pipiens2 <- gam(
  infected ~
    s(habitat_PC1, k = 4) +
    s(disease_week, bs = "tp", k = 8) +
    s(disease_week, by = year_f, bs = "fs", k = 15) +
    s(year_f, bs = "re") +
    s(site_code, bs = "re") +
    s(trap_type2, bs = "re") +   #trap_type2 is random effect
    offset(log(num_count)),
  family = binomial(link = "cloglog"),
  data = pipiens_data,
  method = "REML"
)
summary(gam_pipiens2)

AIC(gam_pipiens, gam_pipiens2)

plot(
  gam_pipiens2,
  select = 1,
  shade = TRUE,
  residuals = TRUE,
  rug = TRUE
)

# Pipiens diagnostics
par(mfrow = c(2, 2))
gam.check(gam_pipiens)
par(mfrow = c(1, 1))
plot(gam_pipiens, pages = 1, residuals = TRUE, shade = TRUE, seWithMean = TRUE)

# Concurvity
concurvity(gam_pipiens, full = FALSE)

# Pipiens diagnostics
par(mfrow = c(2, 2))
gam.check(gam_pipiens2)
par(mfrow = c(1, 1))
plot(gam_pipiens2, pages = 1, residuals = TRUE, shade = TRUE, seWithMean = TRUE)

# Concurvity
concurvity(gam_pipiens2, full = FALSE)

```
Interpretation:

Formula:
infected ~ habitat_PC1 + s(disease_week, bs = "tp", k = 8) + 
    s(disease_week, by = year_f, bs = "fs", k = 15) + s(year_f, 
    bs = "re") + s(site_code, bs = "re") + s(trap_type2, bs = "re") + 
    offset(log(num_count))    
            Estimate Std. Error z value Pr(>|z|)      
habitat_PC1 -0.26895    0.08602  -3.127  0.00177 **   
s(disease_week) 4.36446  5.106 54.269  < 2e-16 ***  
 - Smooth term for disease week is significant.  
 - habitat_PC1 is statistically significant. 
 
Formula:
infected ~ s(habitat_PC1, k = 4) + s(disease_week, bs = "tp", 
    k = 8) + s(disease_week, by = year_f, bs = "fs", k = 15) + 
    s(year_f, bs = "re") + s(site_code, bs = "re") + s(trap_type2, 
    bs = "re") + offset(log(num_count))  
                      edf Ref.df  Chi.sq  p-value    
s(disease_week) 4.3730549  5.113  54.954  < 2e-16 ***  
s(habitat_PC1)  1.8748501  2.259  11.168  0.00643 **   
 - Smooth term for disease week is significant.  
 - The smooth term for habitat_PC1 is statistically significant.  
      - EDF ≈ 1 would be essentially linear.  
      - EDF ≈ 1.8 suggests modest nonlinearity.  
 - The AIC improvement with including a smooth term for habitat_PC1 is tiny (less than 1).  

I think this means there is evidence that habitat_PC1 matters for pipiens, and there may be some curvature, but the curvature is fairly weak. In other words, probability of infection in pipiens varies significantly across the urbanization gradient, even when we force a linear effect. So, for now I'll let habitat_PC1 have a linear relationship. 


### Pip model 2v2
```{r model2v2-pip}
# Pip MODEL 2v2: WNV infection ~ urbanization + trap_type2 + GAM smooths

gam_pipiens_trapfixed <- gam(
infected ~ habitat_PC1 + trap_type2 +
  s(disease_week, bs = "tp", k = 8) +
s(disease_week, by = year_f, bs = "fs", k = 15) +
s(year_f, bs = "re") +
s(site_code, bs = "re") +
offset(log(num_count)),
family = binomial(link = "cloglog"),
data = pipiens_data,
method = "REML"
)
summary(gam_pipiens_trapfixed)

#allow smooth to vary across urbanization gradient (non-linear)
gam_pipiens2_trapfixed <- gam(
  infected ~ trap_type2 +
    s(habitat_PC1, k = 4) +
    s(disease_week, bs = "tp", k = 8) +
    s(disease_week, by = year_f, bs = "fs", k = 15) +
    s(year_f, bs = "re") +
    s(site_code, bs = "re") +
    offset(log(num_count)),
  family = binomial(link = "cloglog"),
  data = pipiens_data,
  method = "REML"
)
summary(gam_pipiens2_trapfixed)

AIC(gam_pipiens, gam_pipiens2, gam_pipiens_trapfixed, gam_pipiens2_trapfixed)

plot(
  gam_pipiens2_trapfixed,
  select = 1,
  shade = TRUE,
  residuals = TRUE,
  rug = TRUE
)

# Pipiens diagnostics
par(mfrow = c(2, 2))
gam.check(gam_pipiens_trapfixed)
par(mfrow = c(1, 1))
plot(gam_pipiens_trapfixed, pages = 1, residuals = TRUE, shade = TRUE, seWithMean = TRUE)

# Concurvity
concurvity(gam_pipiens, full = FALSE)

# Pipiens diagnostics
par(mfrow = c(2, 2))
gam.check(gam_pipiens2_trapfixed)
par(mfrow = c(1, 1))
plot(gam_pipiens2_trapfixed, pages = 1, residuals = TRUE, shade = TRUE, seWithMean = TRUE)

# Concurvity
concurvity(gam_pipiens2_trapfixed, full = FALSE)

```
Interpretation:  

Formula:
infected ~ habitat_PC1 + trap_type2 + s(disease_week, bs = "tp", 
    k = 8) + s(disease_week, by = year_f, bs = "fs", k = 15) + 
    s(year_f, bs = "re") + s(site_code, bs = "re") + offset(log(num_count))  
                Estimate Std. Error z value Pr(>|z|)      
habitat_PC1     -0.25376    0.08757  -2.898  0.00376 **   
trap_type2GRVD+  1.11828    0.23886   4.682 2.85e-06 ***  
s(disease_week)  4.365484  5.107  54.819  <2e-16 ***  


Formula:
infected ~ trap_type2 + s(habitat_PC1, k = 4) + s(disease_week, 
    bs = "tp", k = 8) + s(disease_week, by = year_f, bs = "fs", 
    k = 15) + s(year_f, bs = "re") + s(site_code, bs = "re") + 
    offset(log(num_count))  
                Estimate Std. Error z value Pr(>|z|)      
trap_type2GRVD+   1.0689     0.2488   4.297 1.73e-05 ***  
                                 edf Ref.df Chi.sq p-value    
s(habitat_PC1)             1.8668036  2.252  9.572  0.0135 *  
s(disease_week)            4.3739313  5.114 55.484  <2e-16 ***

                             df      AIC  
gam_pipiens            27.64861 1231.843  
gam_pipiens2           28.99880 1231.156  
gam_pipiens_trapfixed  27.65630 1231.729  
gam_pipiens2_trapfixed 29.02111 1231.057  

Trap_type2 is significant as well. AIC is very similar with/without smooth term, and with trap_type as a fixed_effect or a random effect.  Do we need to allow an interaction term, or let the smooth vary by trap type?  
  



### Pip Moran's I for Cx. pipiens residuals
```{r moran-pip}
pipiens_resids <- pipiens_data %>%
  mutate(resid = residuals(gam_pipiens, type = "deviance")) %>%
  group_by(site_code) %>%
  summarise(
    mean_resid = mean(resid, na.rm = TRUE),
    latitude = first(latitude),
    longitude = first(longitude),
    .groups = "drop"
  ) %>%
  filter(!is.na(mean_resid), !is.na(latitude), !is.na(longitude))

coords_pipiens <- cbind(pipiens_resids$longitude, pipiens_resids$latitude)

lw5_pipiens <- coords_pipiens %>%
  knearneigh(k = 5) %>%
  knn2nb() %>%
  nb2listw(style = "W")

moran.test(pipiens_resids$mean_resid, lw5_pipiens)

```
Moran's I shows no spatial autocorrelation in residuals for pipiens.

## Tarsalis   
### Tar model 2v1
```{r gam-tar}
tarsalis_data <- wnv %>% filter(mosq_species == "Cx_tarsalis")

gam_tarsalis <- gam(
infected ~ habitat_PC1 +
  s(disease_week, bs = "tp", k = 8) +
s(disease_week, by = year_f, bs = "fs", k = 15) +
s(year_f, bs = "re") +
s(site_code, bs = "re") +
s(trap_type2, bs = "re") +
offset(log(num_count)),
family = binomial(link = "cloglog"),
data = tarsalis_data,
method = "REML"
)
summary(gam_tarsalis)
AIC(gam_tarsalis)

#allow smooth to vary across urbanization gradient (non-linear)
gam_tarsalis2 <- gam(
  infected ~
    s(habitat_PC1, k = 4) +
    s(disease_week, bs = "tp", k = 8) +
    s(disease_week, by = year_f, bs = "fs", k = 15) +
    s(year_f, bs = "re") +
    s(site_code, bs = "re") +
    s(trap_type2, bs = "re") +   #trap_type2 is random effect
    offset(log(num_count)),
  family = binomial(link = "cloglog"),
  data = tarsalis_data,
  method = "REML"
)
summary(gam_tarsalis2)

AIC(gam_tarsalis, gam_tarsalis2)

plot(
  gam_tarsalis2,
  select = 1,
  shade = TRUE,
  residuals = TRUE,
  rug = TRUE
)


# Tarsalis diagnostics
par(mfrow = c(2, 2))
gam.check(gam_tarsalis)
par(mfrow = c(1, 1))
plot(gam_tarsalis, pages = 1, residuals = TRUE, shade = TRUE, seWithMean = TRUE)
concurvity(gam_tarsalis, full = FALSE)

par(mfrow = c(2, 2))
gam.check(gam_tarsalis2)
par(mfrow = c(1, 1))
plot(gam_tarsalis2, pages = 1, residuals = TRUE, shade = TRUE, seWithMean = TRUE)
concurvity(gam_tarsalis2, full = FALSE)



```

Interpretation:
habitat_PC1 and the smooth term for habitat_PC1 are NOT statistically significant.
The AIC improvement with including a smooth term for habitat_PC1 is tiny (less than 1).

I think this means there is no evidence that habitat_PC1 matters for tarsalis. In other words, probability of infection in tarsalis does not vary across the urbanization gradient.
  
### Tar model 2v2
```{r model2v2-tar}
# Tar MODEL 2v2: WNV infection ~ urbanization + trap_type2 + GAM smooths

gam_tarsalis_trapfixed <- gam(
infected ~ habitat_PC1 + trap_type2 +
  s(disease_week, bs = "tp", k = 8) +
s(disease_week, by = year_f, bs = "fs", k = 15) +
s(year_f, bs = "re") +
s(site_code, bs = "re") +
offset(log(num_count)),
family = binomial(link = "cloglog"),
data = tarsalis_data,
method = "REML"
)
summary(gam_tarsalis_trapfixed)

#allow smooth to vary across urbanization gradient (non-linear)
gam_tarsalis2_trapfixed <- gam(
  infected ~ trap_type2 +
    s(habitat_PC1, k = 4) +
    s(disease_week, bs = "tp", k = 8) +
    s(disease_week, by = year_f, bs = "fs", k = 15) +
    s(year_f, bs = "re") +
    s(site_code, bs = "re") +
    offset(log(num_count)),
  family = binomial(link = "cloglog"),
  data = tarsalis_data,
  method = "REML"
)
summary(gam_tarsalis2_trapfixed)

AIC(gam_tarsalis, gam_tarsalis2, gam_tarsalis_trapfixed, gam_tarsalis2_trapfixed)

plot(
  gam_tarsalis2_trapfixed,
  select = 1,
  shade = TRUE,
  residuals = TRUE,
  rug = TRUE
)

# tarsalis diagnostics
par(mfrow = c(2, 2))
gam.check(gam_tarsalis_trapfixed)
par(mfrow = c(1, 1))
plot(gam_tarsalis_trapfixed, pages = 1, residuals = TRUE, shade = TRUE, seWithMean = TRUE)

# Concurvity
concurvity(gam_tarsalis, full = FALSE)

# tarsalis diagnostics
par(mfrow = c(2, 2))
gam.check(gam_tarsalis2_trapfixed)
par(mfrow = c(1, 1))
plot(gam_tarsalis2_trapfixed, pages = 1, residuals = TRUE, shade = TRUE, seWithMean = TRUE)

# Concurvity
concurvity(gam_tarsalis2_trapfixed, full = FALSE)

```
Interpretation:

Formula:
infected ~ habitat_PC1 + s(disease_week, bs = "tp", k = 8) + 
    s(disease_week, by = year_f, bs = "fs", k = 15) + s(year_f, 
    bs = "re") + s(site_code, bs = "re") + s(trap_type2, bs = "re") + 
    offset(log(num_count))  
            Estimate Std. Error z value Pr(>|z|)      
(Intercept)  -9.0764     0.3880 -23.392   <2e-16 ***  
habitat_PC1  -0.1224     0.1405  -0.871    0.384      
                                 edf    Ref.df Chi.sq  p-value      
s(disease_week)            2.484e+00 3.070e+00 24.425 2.66e-05 ***  

Formula:  
infected ~ habitat_PC1 + trap_type2 + s(disease_week, bs = "tp",   
    k = 8) + s(disease_week, by = year_f, bs = "fs", k = 15) +   
    s(year_f, bs = "re") + s(site_code, bs = "re") + offset(log(num_count))  
                  Estimate Std. Error z value Pr(>|z|)    
(Intercept)     -9.087e+00  3.844e-01 -23.638   <2e-16 ***
habitat_PC1     -1.434e-01  1.422e-01  -1.009    0.313    
trap_type2GRVD+ -1.144e+02  7.698e+06   0.000    1.000    
  
                              df      AIC  
gam_tarsalis            18.53465 781.5625  
gam_tarsalis2           19.03913 782.0304  
gam_tarsalis_trapfixed  18.84940 779.8560  

Trap_type is not significant in any of the analysis. Neither is habitat_PC1. Disease week is significant in the simplest model, gam_tarsalis.

### Tar Moran's I for Cx. tarsalis residuals
```{r moran-tar}
tarsalis_resids <- tarsalis_data %>%
  mutate(resid = residuals(gam_tarsalis, type = "deviance")) %>%
  group_by(site_code) %>%
  summarise(
    mean_resid = mean(resid, na.rm = TRUE),
    latitude = first(latitude),
    longitude = first(longitude),
    .groups = "drop"
  ) %>%
  filter(!is.na(mean_resid), !is.na(latitude), !is.na(longitude))

coords_tarsalis <- cbind(tarsalis_resids$longitude, tarsalis_resids$latitude)

lw5_tarsalis <- coords_tarsalis %>%
  knearneigh(k = 5) %>%
  knn2nb() %>%
  nb2listw(style = "W")

moran.test(tarsalis_resids$mean_resid, lw5_tarsalis)
```

**Interpretation:**
No evidence of spatial autocorrelation in tarsalis.

## Plot raw by urbanization:
```{r}
wnv_site <- wnv %>%
  group_by(site_code, mosq_species) %>%
  summarise(
    pct_infected = mean(infected, na.rm = TRUE) * 100,
    n_pools = n(),
    habitat_PC1 = first(habitat_PC1),
    .groups = "drop"
  ) %>%
  filter(!is.na(habitat_PC1)) %>%
  mutate(
    mosq_species = recode(
      mosq_species,
      "Cx_pipiens_sl" = "Cx. pipiens s.l.",
      "Cx_tarsalis" = "Cx. tarsalis"
    )
  )

ggplot(wnv_site, aes(x = habitat_PC1, y = pct_infected)) +
  geom_point(aes(size = n_pools), alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "firebrick", linewidth = 1) +
  facet_wrap(~ mosq_species, scales = "free_y") +
  scale_x_reverse() +
  scale_size_continuous(name = "Pools tested", range = c(2, 8)) +
  labs(
    title = "Urbanization Gradient vs WNV Infection Rate",
    x = "PC1 — Urbanization gradient",
    y = "% infected pools per site"
  ) +
  theme_classic(base_size = 13)

cols <- c(
  "Cx. pipiens s.l." = "#1bc8ea",
  "Cx. tarsalis" = "#FF2DA0"
)

wnv_site <- wnv %>%
  group_by(site_code, mosq_species) %>%
  summarise(
    pct_infected = mean(infected, na.rm = TRUE) * 100,
    n_pools = n(),
    habitat_PC1 = first(habitat_PC1),
    .groups = "drop"
  ) %>%
  filter(!is.na(habitat_PC1)) %>%
  mutate(
    mosq_species = recode(
      mosq_species,
      "Cx_pipiens_sl" = "Cx. pipiens s.l.",
      "Cx_tarsalis" = "Cx. tarsalis"
    )
  )

ggplot(
  wnv_site,
  aes(
    x = habitat_PC1,
    y = pct_infected,
    color = mosq_species
  )
) +
  geom_point(
    aes(size = n_pools),
    alpha = 0.7
  ) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    linewidth = 1
  ) +
  scale_color_manual(values = cols, name = "Species") +
  scale_x_reverse() +
  scale_size_continuous(
    name = "Pools tested",
    range = c(2, 8)
  ) +
  labs(
    title = NULL,
    x = "PC1 — Urbanization gradient",
    y = "% infected pools per site"
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5)
  )
```

  
## Plot predicted  

For Cx. pipiens:   
 - habitat_PC1 is significant   
 - trap_type2 is significant   
 - allowing habitat_PC1 to be smooth provides a slight improvement   
 - the smooth is only mildly nonlinear (edf ≈ 1.9)    

For Cx. tarsalis:   
 - habitat_PC1 is not significant  
 - trap_type2 is not significant  
 - there is no evidence for a nonlinear habitat effect  - the fitted relationship is essentially flat

So, I think it's best to fit and plot both species using the same model structure that I landed on for pipiens.  

infected ~ habitat_PC1 + trap_type2 +  
  s(disease_week, bs = "tp", k = 8) +  
  s(disease_week, by = year_f, bs = "fs", k = 15) +  
  s(year_f, bs = "re") +  
  s(site_code, bs = "re") +  
  offset(log(num_count))  
  
gam_pipiens_trapfixed  
gam_tarsalis_trapfixed   
  
###Urbanization gradient
```{r plot-pred}
pc1_seq <- seq(
  min(wnv$habitat_PC1, na.rm = TRUE),
  max(wnv$habitat_PC1, na.rm = TRUE),
  length.out = 200
)

pred_pip <- expand.grid(
  habitat_PC1 = pc1_seq,
  trap_type2 = levels(pipiens_data$trap_type2)
)

pred_tar <- expand.grid(
  habitat_PC1 = pc1_seq,
  trap_type2 = levels(tarsalis_data$trap_type2)
)

pred_pip$disease_week <- median(pipiens_data$disease_week)
pred_pip$year_f <- levels(pipiens_data$year_f)[1]
pred_pip$site_code <- levels(pipiens_data$site_code)[1]
pred_pip$num_count <- 1

pred_tar$disease_week <- median(tarsalis_data$disease_week)
pred_tar$year_f <- levels(tarsalis_data$year_f)[1]
pred_tar$site_code <- levels(tarsalis_data$site_code)[1]
pred_tar$num_count <- 1


pip_pred <- predict(
  gam_pipiens_trapfixed,
  newdata = pred_pip,
  type = "link",
  se.fit = TRUE,
  exclude = c("s(site_code)", "s(year_f)")
)

pred_pip <- pred_pip %>%
  mutate(
    fit = 1 - exp(-exp(pip_pred$fit)),
    lower = 1 - exp(-exp(pip_pred$fit - 1.96 * pip_pred$se.fit)),
    upper = 1 - exp(-exp(pip_pred$fit + 1.96 * pip_pred$se.fit)),
    species = "Cx. pipiens s.l."
  )

tar_pred <- predict(
  gam_tarsalis_trapfixed,
  newdata = pred_tar,
  type = "link",
  se.fit = TRUE,
  exclude = c("s(site_code)", "s(year_f)")
)

pred_tar <- pred_tar %>%
  mutate(
    fit = 1 - exp(-exp(tar_pred$fit)),
    lower = 1 - exp(-exp(tar_pred$fit - 1.96 * tar_pred$se.fit)),
    upper = 1 - exp(-exp(tar_pred$fit + 1.96 * tar_pred$se.fit)),
    species = "Cx. tarsalis"
  )

pred_wnv <- bind_rows(pred_pip, pred_tar)
cols <- c(
  "Cx. pipiens s.l." = "#1bc8ea",
  "Cx. tarsalis" = "#FF2DA0"
)

ggplot( pred_wnv, aes( x = habitat_PC1, y = fit * 100, color = species, fill = species, linetype = trap_type2 ) ) + 
  geom_ribbon( aes( ymin = lower * 100, ymax = upper * 100, group = interaction(species, trap_type2) ), alpha = 0.15, color = NA ) + 
  geom_line(linewidth = 1.2) + 
  scale_x_reverse() + 
  scale_color_manual(values = cols) + 
  scale_fill_manual(values = cols) + 
  labs( 
    x = "PC1 — Urbanization gradient", 
    y = "Predicted WNV positivity (%)", 
    color = "Species", 
    fill = "Species", 
    linetype = "Trap type" 
  ) + 
  theme_classic(base_size = 13) +
  # 1. Zoom into the y-axis without clipping data points outside the window
  coord_cartesian(ylim = c(0, 1.2)) +
  # 2. Place the legend on right
  theme(
    legend.position = "right",
    legend.background = element_rect(fill = "transparent")
  )
```  
### Disease week

```{r pred-plot}
# Predicted WNV positivity through time

cols <- c(
  "Cx. pipiens s.l." = "#1bc8ea",
  "Cx. tarsalis" = "#FF2DA0"
)

pc1_values <- wnv %>%
  filter(!is.na(habitat_PC1)) %>%
  summarise(
    Urban = quantile(habitat_PC1, 0.1, na.rm = TRUE),
    Peri = median(habitat_PC1, na.rm = TRUE),
    Rural = quantile(habitat_PC1, 0.9, na.rm = TRUE)
  ) %>%
  tidyr::pivot_longer(
    everything(),
    names_to = "habitat_level",
    values_to = "habitat_PC1"
  )

week_seq <- seq(
  min(wnv$disease_week, na.rm = TRUE),
  max(wnv$disease_week, na.rm = TRUE),
  by = 1
)

pred_pip <- expand.grid(
  disease_week = week_seq,
  trap_type2 = levels(pipiens_data$trap_type2),
  year_f = "2021",
  site_code = levels(pipiens_data$site_code)[1],
  num_count = median(pipiens_data$num_count, na.rm = TRUE)
) %>%
  left_join(pc1_values, by = character())

pred_tar <- expand.grid(
  disease_week = week_seq,
  trap_type2 = levels(tarsalis_data$trap_type2),
  year_f = "2021",
  site_code = levels(tarsalis_data$site_code)[1],
  num_count = median(tarsalis_data$num_count, na.rm = TRUE)
) %>%
  left_join(pc1_values, by = character())

pip_pred <- predict(
  gam_pipiens_trapfixed,
  newdata = pred_pip,
  type = "link",
  se.fit = TRUE,
  exclude = c("s(site_code)", "s(year_f)")
)

tar_pred <- predict(
  gam_tarsalis_trapfixed,
  newdata = pred_tar,
  type = "link",
  se.fit = TRUE,
  exclude = c("s(site_code)", "s(year_f)")
)

pred_pip <- pred_pip %>%
  mutate(
    fit = 1 - exp(-exp(pip_pred$fit)),
    lower = 1 - exp(-exp(pip_pred$fit - 1.96 * pip_pred$se.fit)),
    upper = 1 - exp(-exp(pip_pred$fit + 1.96 * pip_pred$se.fit)),
    species = "Cx. pipiens s.l."
  )

pred_tar <- pred_tar %>%
  mutate(
    fit = 1 - exp(-exp(tar_pred$fit)),
    lower = 1 - exp(-exp(tar_pred$fit - 1.96 * tar_pred$se.fit)),
    upper = 1 - exp(-exp(tar_pred$fit + 1.96 * tar_pred$se.fit)),
    species = "Cx. tarsalis"
  )

pred_wnv_week <- bind_rows(pred_pip, pred_tar) %>%
  mutate(
    habitat_level = factor(habitat_level, levels = c("Rural", "Peri","Urban" ))
  )
ggplot(
  pred_wnv_week,
  aes(
    x = disease_week,
    y = fit * 100,
    color = species,
    fill = species,
    linetype = trap_type2
  )
) +
  geom_ribbon(
    aes(
      ymin = lower * 100,
      ymax = upper * 100,
      group = interaction(species, trap_type2, habitat_level)
    ),
    alpha = 0.12,
    color = NA
  ) +
  geom_line(
    aes(group = interaction(species, trap_type2, habitat_level)),
    linewidth = 1.1
  ) +
  facet_grid(species ~ habitat_level, scales = "free_y") +
  scale_color_manual(values = cols, name = "Species") +
  scale_fill_manual(values = cols, guide = "none") +
  scale_y_continuous(limits = c(0, 50)) +
  labs(
    x = "Disease week",
    y = "Predicted WNV positivity (%)",
    linetype = "Trap type"
  ) +
  theme_classic(base_size = 13) +
  theme(
    strip.background = element_blank(),
    strip.text.y = element_blank(),
    strip.text.x = element_text(),
    legend.position = "right"
  )
```
# RQ3: Bloodmeal and WNV probability




