---
title: "GAM_Cx.pip/tar_abundance_SLC2025"
author: "Katie Graybeal"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

# Setup
```{r 1-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  # Saves PNGs to the root /figures folder
  fig.path = "../figures/" 
)
```
**Research Topic:** testing whether habitat and seasonal partitioning between Culex pipiens s.l. and Culex tarsalis shapes West Nile Virus (WNV) dynamics across urban–rural gradients. 

**Core hypothesis:** early/mid-season amplification dominated by pipiens in urban areas, later spillover involving tarsalis moving into urban/peri-urban areas.

**Approach:** Preliminary results visualized via mapping, with species identity and abundance as primary response variables. Model mosquito abundance and proportions using GLMM with GAM smoothing:

count ~ season*urbanization + trap_type + (1|site/date), family = poisson(link = "log"):   

 - Response variable = mosquito abundance  
 - Predictors = season*urbanization  
 - The trap type could be important, so we will add that as a fixed effect (covariate)... is this correct? We do think that the response variable of count of mosquitoes depends on trap type, since tarsalis seems to be more attracted to CO2 than pipiens, and we want to quantify that effect. Note that poisson model does not give a fixed offset (due to the log link)... The structure of this model means that it will estimate an effect that scales with the total number of mosquitos caught, which is exactly what we want.  
 - The data are grouped into sites and are also linked through time, so we'll add those as random effects. I think the sites should be coded as factors, **but I'm not sure what format to use for the date. I think it should be disease week so that week 18 is treated closer to 19 than 20, etc., but I'm not totally confident in this.** 
 - The family = poisson (link = "log")... why again?  
 
**For simple model:** 
count ~ disease_week*urbanization + (1|site/date), family = poisson(link = "log")


Load libraries
```{r}
library(tidyverse) # for data wrangling
library(glmmTMB)   # for model fitting
library(DHARMa)    # for residual plots
library(mgcViz)    # for residual plots
library(emmeans)   # for estimating marginal effects
library(multcomp)  # for statistical comparisons on fitted models
library(dplyr)     # for mutating dataframe to change labels in dataset
library(mgcv)      # fits GAM
library(broom)
library(ggplot2)
```
  
# Prepare Data

## Combined data 
```{r}
## tarsalis datasets from SLCMAD:
tarsalis <- read.csv("../data/tarsalis_2025.csv")
## pipiens datasets from SLCMAD:
pipiens <- read.csv("../data/pipiens_2025.csv")
## combine

combined <- bind_rows(tarsalis, pipiens)

## Set factor levels
combined <- combined %>%
  mutate(
    species = factor(
      species,
      levels = c("Culex pipiens", "Culex tarsalis")
    ),
    urban_cat = trimws(tolower(urban_cat)),
    urbanization = factor(
      urban_cat,
      levels = c("rural", "peri", "urban")
    ),
    season = factor(
      season,
      levels = c("early", "mid", "late")
    )
  )

#check
table(combined$species)
table(combined$season, combined$species)

combined <- combined %>%
  mutate(
    species = factor(species),
    urbanization = factor(urbanization),
    trap_type = factor(trap_type),
    site_name = factor(site_name),
    disease_week = as.numeric(disease_week)
  )
```

# GAM with each species separately, count by urbanization + trap_type (random effect = site name)

## Fit model
```{r fit-gam-separated-spp}
# pull out pipiens
pipiens <- combined[combined$species == "Culex pipiens",]
head(pipiens$site_name)

# Fit GAM model with site_name random effect
pip_gam <- gam(
  count ~ urbanization + trap_type + 
    s(disease_week, 
      by = urbanization, 
      bs = "fs",  # bs = "fs" for independent smooths
      k=10, 
      m=3) +  # m = 3 restricts wigglyness
    s(site_name, bs = "re"),
  family = nb(),
  data = pipiens,
  method = "REML"
)

# pull out tarsalis
tarsalis <- combined[combined$species == "Culex tarsalis",]
head(tarsalis$site_name)

# Fit GAM model with site_name only
tar_gam <- gam(
  count ~ urbanization + trap_type + 
    s(disease_week, 
      by = urbanization, 
      bs = "fs",  # bs = "fs" for independent smooths
      k=20, 
      m=3) +
    s(site_name, bs = "re"),
  family = nb(),
  data = tarsalis,
  method = "REML"
)


```

#Splitting up tarsalis by trap type
  
```{r tarsalis_CO2/GRVD}

table(tarsalis$urbanization,tarsalis$trap_type)

tarsalis_CO2 <- combined %>%
  filter(species == "Culex tarsalis",
         trap_type == "CO2")

tarsalis_GRVD <- combined %>%
  filter(species == "Culex tarsalis",
         trap_type == "GRVD")

head(tarsalis_CO2)
head(tarsalis_GRVD)

#tarsalis_C02 gam
gam_tarsalis_CO2 <- gam(
  count ~ urbanization +
    s(disease_week, k = 10, m = 2) +
    s(site_name, bs = "re"),
  family = nb(),
  data = tarsalis_CO2,
  method = "REML"
)

table(tarsalis_CO2$urbanization)
table(tarsalis_CO2$site_name)
str(tarsalis_CO2)


```

#splitting up pipiens by trap type
```{r}

table(pipiens$urbanization,pipiens$trap_type)

pipiens_CO2 <- combined %>%
  filter(species == "Culex pipiens",
         trap_type == "CO2")

#pipiens_CO2 gam
gam_pipiens_CO2 <- gam(
  count ~ urbanization +
    s(disease_week, k = 10, m = 2) +
    s(site_name, bs = "re"),
  family = nb(),
  data = pipiens_CO2,
  method = "REML"
)


#pipiens_GRVD <- combined %>%
  #filter(species == "Culex pipiens",
        # trap_type == "GRVD")

head(pipiens_CO2)
#head(pipiens_GRVD)

table(pipiens_CO2$urbanization)
table(pipiens_CO2$site_name)
str(pipiens_CO2)

```

#Compare models on the same dataset. If your scientific question is "Does trap type matter?", fit both models to the full dataset:
```{r}
m1 <- gam(
  count ~ species * urbanization +
    s(disease_week, by = species, bs = "fs", k = 10, m = 2) +
    s(site_name, bs = "re"),
  family = nb(),
  data = combined,
  method = "REML"
)

m2 <- gam(
  count ~ species * urbanization + trap_type +
    s(disease_week, by = species, bs = "fs", k = 10, m = 2) +
    s(site_name, bs = "re"),
  family = nb(),
  data = combined,
  method = "REML"
)

AIC(m1, m2)

#Likelihood ratio test (nested models) If both models are fit to the same data and one is nested within the other.
anova(m1, m2, test = "Chisq")

```

#Compare Cx.tarsalis abundance by trap type Urban
```{r}
#Analyze Urban sites only:If your question is simply:

"Within Urban sites, does trap type affect abundance?fit the model on Urban observations only"

urban_dat <- subset(tarsalis, urbanization == "urban")
nrow(urban_dat)
summary(urban_dat)

urban_gam <- mgcv::gam(
  count ~ trap_type +
    s(disease_week, k = 5) +
    s(site_name, bs = "re"),
  family = mgcv::nb(),
  data = urban_dat,
  method = "REML"
)

summary(urban_gam)

#Then test the trap effect:

#drop1(urban_gam, test = "Chisq")
```
#Compare Cx.tarsalis abundance by trap type Urban
```{r}
#Analyze Urban sites only:If your question is simply:
"Within Urban sites, does trap type affect abundance?fit the model on Urban observations only"
urban_dat <- subset(tarsalis, urbanization == "urban")
nrow(urban_dat)
summary(urban_dat)
urban_gam <- mgcv::gam(
  count ~ trap_type +
    s(disease_week, k = 5) +
    s(site_name, bs = "re"),
  family = mgcv::nb(),
  data = urban_dat,
  method = "REML"
)
summary(urban_gam)
#Then test the trap effect:
#drop1(urban_gam, test = "Chisq")
```
